From multiplicity of infection to force of infection for sparsely sampled Plasmodium falciparum populations at high transmission

Version 1

Abstract

Abstract

High multiplicity of infection or MOI, the number of genetically distinct parasite strains co-infecting a single human host, characterizes infectious diseases including falciparum malaria at high transmission. It accompanies high asymptomatic Plasmodium falciparum prevalence despite high exposure, creating a large transmission reservoir challenging intervention. High MOI and asymptomatic prevalence are enabled by immune evasion of the parasite achieved via vast antigenic diversity. Force of infection or FOI, the number of new infections acquired by an individual host over a given time interval, is the dynamic sister quantity of MOI, and a key epidemiological parameter for monitoring the impact of antimalarial interventions and assessing vaccine or drug efficacy in clinical trials. FOI remains difficult, expensive, and labor-intensive to accurately measure, especially in high-transmission regions, whether directly via cohort studies or indirectly via the fitting of epidemiological models to repeated cross-sectional surveys. We propose here the application of queuing theory to obtain FOI on the basis of MOI, in the form of either a two-moment approximation method or Little’s law. We illustrate these methods with MOI estimates obtained under sparse sampling schemes with the recently proposed “varcoding” method, based on sequences of the var multigene family encoding for the major variant surface antigen of the blood stage of malaria infection. The methods are evaluated with simulation output from a stochastic agent-based model, and are applied to an interrupted time-series study from Bongo District in northern Ghana before and immediately after a three-round transient indoor residual spraying (IRS) intervention. We incorporate into the sampling of the simulation output, limitations representative of those encountered in the collection of field data, including under-sampling of var genes, missing data, and usage of antimalarial drug treatment. We address these limitations in MOI estimates with a Bayesian framework and an imputation bootstrap approach. We demonstrate that both proposed methods give good and consistent FOI estimates across various simulated scenarios. Their application to the field surveys shows a pronounced reduction in annual FOI during intervention, of more than 70%. The proposed approach should be applicable to the many geographical locations where cohort or cross-sectional studies with regular and frequent sampling are lacking but single-time-point surveys under sparse sampling schemes are available, and for MOI estimates obtained in different ways. They should also be relevant to other pathogens of humans, wildlife and livestock whose immune evasion strategies are based on large antigenic variation resulting in high multiplicity of infection.

Introduction

Despite substantial control and elimination efforts, falciparum malaria in high-transmission regions continues to be a major public health concern, as a cause of mortality among young children and considerable economic burden in many countries especially in sub-Saharan Africa (World Health Organization, 2023). It thus remains important to robustly evaluate the effects of intervention efforts in these regions, including on transmission intensity. Collectively, epidemiological studies have revealed a multiplicity of metrics that are associated with parasite transmission (Tusting et al., 2014). Chief among them is the force of infection (FOI), defined as the number of new Plasmodium falciparum infections acquired by an individual host over a given time interval. FOI has been found to reflect risk of infection and to be tightly associated with risk of clinical episodes (Mueller et al., 2012). While other metrics may describe the relationship between transmission intensity and the burden of malaria illness on global or continental scales (Carneiro et al., 2010; Beier et al., 1994), FOI can relate local variation in malaria burden to transmission (Mueller et al., 2012). Despite being recognized as a key epidemiological parameter for malaria surveillance, FOI remains difficult, expensive, and labor-intensive to accurately measure, either directly or indirectly via the fitting of epidemiological models. As molecular tools for parasite genomics become more readily available, they are contributing to new approaches. In particular, molecular advances are providing a basis for estimating a sister “static” quantity, the multiplicity of infection (MOI), defined as the number of genetically distinct parasite strains co-infecting a single human host (Zhong et al., 2018; Chang et al., 2017; Ruybal-Pesántez et al., 2022; Tiedje et al., 2022; Labbé et al., 2023). We can therefore ask whether we can go further and on the basis of MOI obtain the dynamical, rate, quantity of force of infection?

Prior efforts to directly measure FOI included clearing infections and observing time to re-infection. Infant conversion rate, the time to a patent infection in uninfected children less than 1-year of age, was widely used in the 1950s to 1970s (Macdonald, 1950; Pull and Grab, 1974). An alternative way was to perform detection of parasites by light microscopy for subjects following antimalarial treatment (Msuya and Curtis, 1991; Alonso et al., 2004). In both cases, FOI was measured directly in either a naturally uninfected cohort (e.g. infants or uninfected immigrants) or by artificially creating a cohort of uninfected individuals through treatment with antimalarial drugs, which was then followed to estimate the attack rate or proportion acquiring infection over some time period. More recently, with the advent of molecular approaches to malaria epidemiology, direct and more precise measures of FOI can be generated by genotyping individual parasite infections (Mueller et al., 2012; Hofmann et al., 2017). Individual clones can be tagged by a set of molecular markers which are polymorphic, a precondition for longitudinally tracking and differentiating individual parasite clones. PCR-based techniques also offer the added advantage of higher sensitivity by early infection detection, when parasite density is under the detection limit of microscopy (Johnston et al., 2006). It remains challenging however to accurately differentiate a truly new infection from the temporary absence from the peripheral blood of an old infection and its subsequent reemergence (Daubersies et al., 1996). This is primarily because some of these popular polymorphic molecular markers have a relatively low resolution for differentiating different strains from one another. Despite increased access to better techniques, the determination of molecular FOI remains challenging, labor-intensive, and costly, as it still involves closely monitoring and genotyping a large cohort over some period of time.

Alternatives to direct measurements involve cross-sectional surveys with FOI estimated by fitting simple epidemiological models, including Markov models corresponding to the reverse catalytic formulations parameterized from changes in prevalence with age (Bekessy et al., 1976; Muench, 1959; Smith and Vounatsou, 2003), or immigration-death models informed by the values of MOI across age groups (Felger et al., 2012), or iterations between a generalized linear mixed model and a mathematical Susceptible-Infected-Susceptible model (Mugenyi et al., 2017). These model-fitting procedures require empirical data sampled regularly and frequently, such as age-stratified cohorts with a large number of individuals sampled six times a year, to account for the high heterogeneity and variation in FOI and infection duration (Laishram et al., 2012; Simpson et al., 2002; Langhorne et al., 2008; Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014), both of which are dependent on the interplay between hosts immunity profiles and antigen composition of any infection (Piper et al., 1999; Molineaux et al., 2002; Doolan et al., 2009; Barry et al., 2007). In addition, they may suffer from identifiability problems with more than one value of given parameters of interest resulting in similar likelihood or other model selection metrics (Mugenyi et al., 2017). Hence, these indirect measurements share limitations with direct ones.

As a result of the above-described challenges, FOI has not become a readily available epidemiological quantity across time and geographical locations. In contrast, various approaches have been proposed to estimate MOI from clinical samples, involving size-polymorphic antigenic markers, microsatellites, and panels of biallelic single nucleotide polymorphisms (SNPs) (Felger et al., 1994; Konaté et al., 1999; Anderson et al., 2000b; Daniels et al., 2008; Chang et al., 2017). Because Plasmodium parasites reproduce asexually during haploid stages within human hosts (Guttery et al., 2012), signatures of polymorphic genotypes are evidence of multiclonal infection. As methods for disentangling the molecular complexity of natural parasite infections across various transmission settings emerge and mature, MOI becomes much more commonly and easily surveyed across space and time. Although this quantity remains one of the most frequently used genetic metrics of parasite transmission (Arnot, 1998; Sondo et al., 2020), it is by definition a number and not a rate.

A natural correlation exists however between MOI and FOI mediated by the duration of infection, which provides an opportunity to convert the former into the latter. This conversion has been particularly challenging because disease transmission and dynamics display heterogeneity in time and across individuals (Laishram et al., 2012; Simpson et al., 2002; Langhorne et al., 2008), duration of infection information is highly heterogeneous across a multiplicity of factors and mostly lacking (Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014; Bretscher et al., 2011), and the majority of MOI estimates are obtained under sparse sampling schemes. Rather than cohort-studies or cross-sectional studies with regular and frequent sampling, most often single-time-point surveys are implemented, typically in wet (high-transmission) and dry (low-transmission) seasons (Tiedje et al., 2022; Abukari et al., 2019). Although MOI estimates from those single-time-point cross-sectional surveys have been used to address a number of questions such as whether multiple infections protect against clinical attacks, and whether each additional infecting clone is associated with a higher prospective risk or a higher odd of treatment failure (Earland et al., 2019; Lee et al., 2006), the sparse sampling scheme has limited their translation into a transmission rate.

In this work, we propose to specifically use these MOI estimates for the purpose of FOI inference. We present two mathematical modeling frameworks based on queuing theory for this purpose (Choi et al., 2005; Little and Graves, 2008), and illustrate their application with empirical molecular data collected in the Bongo District of northern Ghana (Tiedje et al., 2022, 2023). The proposed methods can be easily scaled up and applied across different geographical locations and time. They are potentially applicable to the many geographical locations where cohort or cross-sectional studies with regular and frequent sampling are lacking but single-time-point surveys are available. We evaluate their performance with numerical simulation of an extended stochastic agent-based model (ABM) (He et al., 2018) for both a closed system and an open one under two different spatial scenarios, specifically two populations coupled via migration and a local population with migration from a regional pool, with either constant or seasonal transmission. We further consider cases of heterogeneous transmission in which a high-risk group of hosts receives the majority of infectious bites. We also examine scenarios in which the times of local transmission events follow different statistical distributions. We incorporate limitations representative of those encountered in the collection of field data into the sampling of simulation output, including under-sampling of var genes, missing data, and usage of antimalarial drug treatment. We obtain MOI estimates with a Bayesian framework and a bootstrap imputation approach correcting for all the aforementioned errors. The Bayesian method accounts for the under-sampling of var genes for individual isolates. The bootstrap imputation approach accounts for the missing data error and the usage of antimalarial treatment by removing the samples for individuals who are under the impact of the two and then inferring their MOI estimates by sampling from those of the remaining population. We show that our Bayesian method and bootstrap imputation approach are robust to considerable fractions of individuals with their MOI information missing or under antimalarial treatment, comparable to those from our field surveys in northern Ghana. We demonstrate further that based on MOI estimates obtained in this way, both proposed methods give good and consistent FOI estimates across various simulated scenarios. After validating their performance with simulation output, we illustrate the application of the proposed methods to empirical data collected from an interrupted time-series study from Bongo District in northern Ghana that involved a three-round transient indoor residual spraying (IRS) intervention (Tiedje et al., 2022, 2023). We conclude with discussion of the relationship between FOI and another commonly used surrogate variable for transmission intensity, the entomological inoculation rate or EIR, defined as the number of infectious bites received by an individual over a given time period (Shaukat et al., 2010). We discuss how this relationship explains the known difficulty in relating transmission to malaria burden at more refined local scales, and that of achieving sustainable and considerable transmission reductions in high-transmission endemic regions. We further discuss how this relationship might be informative of malaria immunity, and within-host competitive interactions of co-infecting strains.

Results

The MOI estimates at the population level for surveys in Ghana resemble those under high-intensity and heterogeneous transmission from numerical simulation

While any polymorphic marker should be suitable in theory for estimating MOI, the common high number of multiclonal infections in high-transmission endemic regions limits the ability to accurately do so in practice (Tiedje et al., 2022; Ruybal-Pesántez et al., 2017; Touray et al., 2020; Labbé et al., 2023). A multigene family known as var which encodes for hyper-diverse antigenic variation in the parasite provides one proposed solution, described in the Materials and Methods and referred to as “varcoding” (Ruybal-Pesántez et al., 2022; Tiedje et al., 2022). We recently extended the method to a Bayesian formulation to account for the measurement error from under-sampling or imperfect detection of var genes in the collection of empirical data (Tiedje et al., 2023), with details summarized in the Materials and Methods-Bayesian approach to MOI estimation with “varcoding”. Additionally, empirical surveys suffer from a missing data error due to factors including the detection limit of the selected method and low DNA quality, and they contain a fraction of individuals under antimalarial drug treatment during a previous window of time. Antimalarial treatment either prevents new infections or clears existing ones by shortening their duration, which potentially violates the assumptions of our two proposed methods for FOI inference. This is because the two methods rely on a known and fixed distribution of duration of infection for 1-5 year-old children. Because these children’s immune profile is similar to that of naive hosts, for this specific sub-population class, the duration of infection can be approximated by that of malaria therapy patients with neurosyphilis and no history of prior exposure to malaria (Collins and Jeffery, 1999; Maire et al., 2006). Details can be found in Materials and Methods-Inferring FOI from MOI estimates. We address both types of errors with a bootstrap imputation approach whose details are included in Materials and Methods-The under-sampling of infections in estimates of MOI from empirical surveys, and Materials and Methods-Antimalarial drug treatment of infections in estimates of MOI from empirical surveys.

Because we rely on MOI estimates to derive FOI ones, we first conduct here an investigation of the MOI estimates obtained under the impact of each of these aforementioned errors, and of all combined, in comparison with true MOI values. We rely on the simulated output from our agent-based model (ABM) of malaria transmission for which true MOI values are known. The ABM is described in detail in previous studies of P. falciparum’s population structure based on var genes (He et al., 2018). We provide additional details on its formulation, including extensions implemented to the model for this work, and on the experimental design of simulation output in Appendix 1-Simulation data. See also Appendix 1-Figure 5A-C for illustrations of the experimental design.

Confidence intervals for estimated mean FOI values in simulated risk scenarios with homogeneous exposure, before and during IRS interventions with three different coverage levels.

The times of the local transmission events follow a Gamma distribution, and transmission is seasonal with the spatial setting corresponding to a closed system. We provide FOI estimates based respectively on the true MOI values, on the MOI estimates with the Bayesian method and a bootstrap imputation approach correcting for all aforementioned errors, and on the MOI estimates with one source of error at a time corrected by either the Bayesian method or the bootstrap imputation approach, namely under the missing data error, the under-sampling of var genes, and the antimalarial treatment issue. To obtain the true mean FOI per host per year, we divide the total number of infections acquired by the population by the total number of individual hosts in the population. Minimum, 5% quantile, median, 95% quantile, and maximum values are shown in the boxplot.

Confidence intervals for the estimated mean FOI values in simulated risk scenarios with heterogeneous exposure, before and during IRS interventions with three different coverage levels.

The times of the local transmission events follow a Gamma distribution, and the transmission is seasonal with the spatial setting corresponding to a semi-open system. We provide FOI estimates based respectively on the true MOI values, on the MOI estimates with the Bayesian method and a bootstrap imputation approach correcting for all aforementioned errors, and on the MOI estimates with one source of error at a time corrected by either the Bayesian method or the bootstrap imputation approach, namely under the missing data error, the under-sampling of var genes only, and the antimalarial treatment issue. To obtain the true mean FOI per host per year, we divide the total number of infections acquired by the population by the total number of individual hosts in the population. Minimum, 5% quantile, median, 95% quantile, and maximum values are shown in the boxplot.

Confidence intervals for the estimated mean FOI values in Ghana surveys across a transient three-round IRS intervention based on the combination of surveys from the wet/high-transmission and dry/low-transmission seasons

(A) The estimated FOI values when considering that antimalarial treatment does not affect infection status and MOI values in treated individuals (assumption 1). (B) The estimated FOI values when considering instead that antimalarial treatment affects infection status and MOI values in treated individuals (assumption 2). High, mid, and low detactability correspond to 0%, 5%, and 10% of PCR-negative individuals carrying infections respectively. Minimum, 5% quantile, median, 95% quantile, and maximum are shown in the boxplot.

Saturation in FOI with EIR and the non-linear relationship between these two quantities from previous field studies.

(A) and (B) includes our empirical estimates under the assumption that antimalarial treatment does not affect infection status and MOI values in treated individuals. (C) and (D) includes our estimates under the assumption that the treatment affects such quantities. The points correspond to paired EIR-FOI values from the literature, summarized in (Smith et al., 2010), and the crosses indicate the range of those values when several estimates of the EIR or the FOI were reported or estimated for the same location. Namely, a line or a cross was plotted with its center at the arithmetic mean and with legs that connect the center to each one of the estimates, and similarly for the scenario when ranges were reported for both variables. The red function curve is the best-fit to these paired EIR-FOI values from (Smith et al., 2010). The black hollow diamond and circle correspond to the Ghana data, for our respective estimates of FOI with the two methods and the EIR measure in the field by the entomological team (Tiedje et al., 2022).

(A) Each simulation follows three stages: a “pre-IRS” period during which the transmission in the local population reaches a stationary state, followed by an “IRS” intervention period of three years which reduces transmission rate (in what we call transient IRS), and a “post-IRS” period when transmission rates go back to their original levels. We let the system run for some years to reach a (semi-) stationary state before applying IRS interventions. After initial seeding, closed systems do not receive migrant genomes from the regional pool. Semi-open systems are explicit simulations of two individual local populations coupled via migration events. Regionally-open systems receive continuous migrant genomes from the regional pool throughout the simulation course. (B) Transmission intensity or effective contact rate (Appendix 1-Simulation data) varies as a function of time, from pre-, to during, to post-intervention. We simulate three different coverages of perturbations, ranging from low-coverage one which reduces the system’s transmission by around 20% to high-coverage one which reduces the system’s transmission by around 70-75%. (C) We incorporate heterogeneity in transmission across individual hosts by varying two things. First, we consider different statistical distributions for times of local transmission events: exponential and Gamma. Second, we consider homogeneous and heterogeneous exposure risks. For the latter, we set of the total population as being at high-risk and receiving more than 90% of all bites whereas the rest population receives only less than 10% of all bites. (D) The measurement error which describes potential sequencing errors of var genes: histogram of the number of non-upsA (i.e., upsB and upsC) DBLα var gene types per repertoire. The molecular sequences were previously sequenced from infections that are expected to be monoclonal (as they had less than or equal to 45 non-upsA DBLα types), sampled during six cross-sectional surveys made from 2012 to 2016 in Bongo District. (E) Study design showing the timing of the four age-stratified cross-sectional surveys conducted in Bongo District, Ghana at the end of the wet/high-transmission seasons (blue circles) and at the end of the dry/low-transmission seasons (gold circles). The study can be broken down into two phases: (1) Pre-IRS: Survey 1 (S1) October 2012 and Survey 2 (S2) May/June 2013; (2) Right post-IRS: Survey 3 (S3) October 2015 and Survey 4 (S4) May/June 2016. IRS were implemented against a background of widespread LLIN usage which were distributed across Bongo District between 2010 – 2012 and again in 2016 (Tiedje et al., 2023; Gogue et al., 2020).

Our analysis shows that MOI estimates obtained with the Bayesian method and the bootstrap imputation approach with all the aforementioned errors present, are close to the true MOI values. We measure the difference between estimated and true MOI by the difference in their mean value and the Kolmogorov-Smirnov statistic, which quantifies the distance in their cumulative distribution function. Results are included in the supplementary file 1-MOImethodsPerformance.xlsx, and illustrated in Appendix 1-Figure 6. The bootstrap imputation approach corrects well for the missing data error and the antimalarial drug treatment issue. The Bayesian method reduces the under-sampling of var genes error, resulting in a noticeable improvement relative to the original varcoding method which relies on a constant repertoire size to convert the number of detected non-upsA DBLα types to the estimated MOI value, and hence does not consider the variation in repertoire size introduced by the under-sampling of var genes. See supplementary file 2-BayesianImprovement.xlsx, Appendix 1-Figure 7, and Appendix 1-A comparison between the Bayesian method and the original varcoding one for further details.

Comparison of the estimated MOI distribution under only one type of error among those commonly encountered in the collection of empirical data (missing data, under-sampling of var genes, and usage of antimalarial drugs) versus under all the types of the errors together and the distribution of true MOI values (indicated by different shades of the red). The scenarios shown here are simulated as semi-open systems under seasonal transmission. Individuals are under heterogeneous exposure risk (Appendix 1-Figure 5C) and the arrival of infection follows a Gamma distribution. The Bayesian varcoding method and the bootstrap imputation approach are able to address all types of the errors. Results of the comparison across other simulated scenarios can be found in supplementary file 1-MOImethodsPerformance.xlsx

Comparison of the estimated MOI distribution based on the original varcoding method, the Bayesian approach, and the true values. The scenarios shown here correspond to a semi-open system under seasonal transmission. Arrival of infection follows a Gamma distribution. Both the Bayesian varcoding approach and the original method perform well, with the former showing a clear improvement in capturing the tail of high MOI values. Results of the comparison across the remaining different simulated scenarios can be found in supplementary file 2-BayesianImprovement.xlsx.

The MOI estimates at the population level for simulation output under high-intensity and heterogeneous transmission are similar in range to the ones for surveys in Ghana (Appendix 1-Figure 6 and 8-9). The respective distributions of MOI estimates across the different surveys in Ghana and from the simulated outputs are not Poisson-distributed (Appendix 1-Test of deviation from Poisson homogeneity in MOI estimates, supplementary file 3-deviationFromPoissonTest.xlsx) (Potthoff and Whittinghill, 1966; Lloyd-Smith et al., 2005). The deviation of empirical MOI estimates from a Poisson distribution indicates that either the arrival of inoculations deviates from a homogeneous Poisson process, or some latent heterogeneity causes inoculation duration distribution to be over-dispersed. Despite these conditions which challenge the conversion of MOI estimates to their corresponding FOI values, the proposed methods below are still applicable (see Materials and Methods-Inferring FOI from MOI estimates for details).

The distribution of MOI estimates based on the Bayesian varcoding method and the bootstrap imputation approach for longitudinal surveys in Bongo District, from pre-IRS to right post-IRS. See Appendix 1-Figure 5E for the study design of the IRS intervention. The wet/high season survey for pre-IRS phase was collected in 2012 and denoted as S1 in Figure 5E. The dry/low season survey for pre-IRS phase was collected in 2013 and denoted as S2. The wet/high season survey for right post-IRS phase was collected in 2015, corresponding to S3. The dry/low season survey for right post-IRS phase was collected in 2016 and denoted as S4. As explained in the methods, the empirical surveys suffer from missing data, i.e., only microscopy-positive individuals have their var genes typed and sequenced. We quantify the number of individuals with missing MOI by utilizing the empirically measured detectability/sensitivity of microscopy relative to PCR, and assuming a range for the PCR detectability. We consider a high detectability in which PCR picks up all infected individuals, to a mid value in which 5% PCR-negative individuals actually carry infections, to a low value in which 10% PCR-negative individuals carry infections. The empirical surveys also suffer from missing data due to other factors than the detection limit of the selected method, for example, low DNA quality. We impute MOI values for all those individuals with missing MOI by sampling from MOI estimates of microscopy-positive individuals with their var sequenced and typed successfully who are not treated. In addition, some individuals may seek and get antimalarial treatment in a previous window of time. We consider here the extreme scenario that antimalarial treatment has no impact on treated individuals’ infection status and MOI values, and use their infection status and MOI values directly for FOI inference.

The Two-Moment Approximation and Little’s Law approaches give good and consistent estimates for FOI across different simulated scenarios

We start with the result of the homogeneous exposure risk scenario for seasonal transmission and a closed system based on the two-moment approximation approach and Little’s law. The times of local transmission events follow a Gamma distribution (Appendix 1-Figure 5C). Across pre-IRS, low-, mid-, and high-coverage IRS, the 95% confidence intervals of annual FOI per host and the full sampling distribution from bootstrap analysis are narrow and concentrated, containing, or very close to, the true FOI values (Figure 1). We infer FOI estimates based respectively on the true MOI values, on the MOI estimates with the Bayesian method and a bootstrap imputation approach correcting for all aforementioned errors, and on the MOI estimates with one source of error at a time corrected by either the Bayesian method or the bootstrap imputation approach, namely under the missing data error, the under-sampling of var genes, and the antimalarial treatment issue. The FOI estimates obtained with the two methods based on the true MOI values differ only slightly from the true FOI values. The FOI estimates based on the MOI estimates with the Bayesian method and a bootstrap imputation approach correcting for all errors exhibit a somewhat larger, but still small, difference from the true FOI values.

We continue with the result of the heterogeneous exposure risk scenarios in which a high-risk group ( of the total population) receives around 94% of all bites whereas a low-risk group( of the total population) receives the remaining bites (Appendix 1-Figure 5C). Transmission is seasonal and the spatial setting corresponds to a semi-open system (Appendix 1-Figure 5A-B, and Appendix 1-Simulation data). The times of local transmission events are Gamma-distributed (Appendix 1-Figure 5C). Across pre-IRS, low-, mid-, and high-coverage IRS, the 95% confidence intervals and the full sampling distribution from bootstrap analysis are narrow and concentrated, containing, or very close to, the true FOI values (Figure 2).

The performance of the two methods across other simulated scenarios is demonstrated in Appendix 1-Figure 10-14. Details for deriving the confidence intervals are provided in Appendix 1-Confidence intervals for mean and variance.

The Two-Moment Approximation and Little’s Law approaches give consistent FOI estimates for empirical surveys from Bongo District in northern Ghana

The two-moment approximation approach and Little’s law also give consistent estimates of the FOI using the empirical MOI estimates from surveys in Ghana. We provide in Materials and Methods, further details and proposed solutions regarding the under-sampling of var genes, as well as antimalarial drug treatment and missing data due to factors such as the low detection limit of the selected method and low DNA quality, in obtaining empirical MOI estimates. Since PCR has a high, but yet not perfect power of infection detection, we further assume three levels of sensitivity, ranging from 0% of PCR-negative individuals carrying infection or a detection of 100% of the infected individuals (high detectability), to 5% of PCR-negative individuals carrying infections (mid detectability), to 10% of PCR-negative individuals carrying infections (low detectability). The impact of antimalarial treatment on infection status and duration, and therefore on MOI values, is difficult if not impossible to disentangle due to the complex biology of the parasite. We thus consider the following two extremes which should bound reality. Antimalarial treatment either does not impact treated individuals’ infection status and MOI values, or it does affect both. For the former assumption (hereafter assumption 1), we include the infection status and MOI estimates of these treated individuals when inferring FOI. For the latter assumption (hereafter assumption 2), we discard their infection status and MOI estimates and sample instead their MOI values from the microscopy-positive individuals who were not under the missing data error or antimalarial treatment.

We start with the two-moment approximation approach. The estimated values of the annual mean FOI before IRS (Appendix 1-Figure 5E, the pre-IRS phase) are based on Survey 1 (S1; Oct 2012) and Survey 2 (S2; May/Jun 2013), undertaken at the end of the wet/high-transmission and dry/low-transmission seasons, respectively. Under assumption 1, the FOI estimates are respectively 5, 5.214286, and 5.367647, assuming respectively that 0%, 5%, 10% PCR-negative individuals carry infections. Under assumption 2, they are equal to 5.983607, 6.186441, and 6.293103, assuming respectively that 0%, 5%, 10% PCR-negative individuals carry infections. The estimated values for the year when IRS was discontinued (Appendix 1-Figure 5E, the post-IRS phase right after IRS was interrupted) are based on Survey 3 (S3; Oct 2015) and Survey 4 (S4; May/Jun 2016), undertaken at the end of the wet/high-transmission and dry/low-transmission seasons, respectively. The corresponding FOI estimates under assumption 1 are 0.7635983, 0.9864865, 1.140625, assuming respectively that 0%, 5%, 10% PCR-negative individuals carry infections. The corresponding FOI estimates under assumption 2 are are 1.46, 1.615044, 1.738095, assuming respectively that 0%, 5%, 10% PCR-negative individuals carry infections.

We then consider Little’s law. The estimated annual mean FOI values before IRS under assumption 1 are 4.83786, 5.089157, 5.237725, for 0%, 5%, 10% PCR-negative individuals carrying infections respectively. For comparison, the estimated annual mean FOI values before IRS under assumption 2 are 5.92716, 6.099559, 6.224758. The corresponding estimated values immediately post-IRS under assumption 1 are 0.764853, 0.9870203, 1.14117. Those values under assumption 2 are 1.456373, 1.622026, 1.741665.

The two-moment approximation and Little’s law approaches give consistent FOI estimates for empirical surveys from Bongo District in northern Ghana. There is a considerable reduction in FOI, of more than 80% or 70% under the two assumptions respectively, which indicates that the three-round IRS intervention although transient was efficient and strong. Reality is likely to lie somewhere in between the two bounding assumptions on the impact of antimalarial treatment on infection status and MOI values, so that such treatment may have a non-negligible effect on some individuals but not others. Regardless, the actual estimates of FOI are similar in magnitude and the trend in the change of FOI across intervention, and so the evaluation of the impact of such intervention, remain qualitatively the same.

The 95% confidence intervals and the full sampling distribution from bootstrap analysis are narrow and concentrated (Figure 3).

The inferred FOI and the directly measured EIR for surveys in Ghana are consistent with the relationship between the two measures from previous studies

We plot the measured mean annual EIR against the estimated mean annual FOI, and the transmission efficiency (the ratio between FOI and EIR) against the measured mean annual EIR from previous field studies, summarized in (Smith et al., 2010) (Figure 4). The estimation of FOI in these studies was primarily based on fitting a simple epidemiological model to age-stratified prevalence from cross-sectional parasitological studies (Smith et al., 2010; Pull and Grab, 1974); that of EIR was based on diverse methods including exit bait collection, human bait collection, pyrethrum spray collection, night bite collection, and outdoor resting collection (Smith et al., 2010). The red line corresponds to the functional curve fitted to these data points as described in Materials and Methods-Conversion between FOI and EIR, as initially proposed in (Smith et al., 2010).

Based on these data pairs, we see that the values for mean annual FOI are well below an empirical limit of 20. Due to the highly non-linear pattern between FOI and EIR, there is no single constant factor that converts FOI to EIR (or the opposite) across different empirical settings. The transmission efficiency in high-intensity endemic regions is extremely low, with an average annual EIR reaching values of a few hundred to a thousand, whereas the average annual FOI takes values of only about 5-10. In future applications, when applying our proposed methods to obtain FOI estimates from MOI under sparse sampling schemes from specific geographical regions, we can rely on this functional curve to convert the FOI estimates to corresponding EIR values. There is however an inherently high variance associated with the conversion. Nevertheless, overall, our paired EIR (directly measured by the entomological team in Ghana (Tiedje et al., 2022)) and FOI values are reasonably consistent with the data points from previous studies, suggesting the robustness of our proposed methods.

The variance inferred reflects transmission intensity and heterogeneity across individuals

We have focused so far on estimation of the mean for FOI in both the simulation output and the empirical data. The two-moment approximation method further gives an estimate for the variance of the inter-arrival times of infections. We examine the variance inferred across all simulated scenarios. We find that regardless of the magnitude of the variance for FOI or the degree of heterogeneity in exposure risks across individual human hosts, the estimated mean FOI at the individual level by the two proposed methods, when aggregated across all individuals within the local population, reflects the total number of infections accumulated by that population.

Overall, seasonal runs have slightly larger variance than non-seasonal runs (Appendix 1-Figure 15). Runs with a heterogeneous transmission, in which a high-risk group of individuals receives more than 90% of the total bites, have slightly higher variance than those with a homogeneous transmission in which bites do not exhibit such variation. These findings are consistent with expectations since seasonality and heterogeneity in transmission should increase the dispersion of FOI across individuals, and consequently of MOI. The inferred variance correlates most significantly with the inferred mean. In particular, when the mean FOI is close to 0, the variance is large. This is intuitive as significant deviation from the close-to-zero mean FOI (or, large inter-arrival times, for example, at the order of several years) would be required to generate some fraction of individual hosts with MOI greater than zero.

Discussion

Building on estimates of the multiplicity of infection under sparse sampling schemes, we have demonstrated the feasibility of converting this quantity to estimates of the force of infection. Across a wide range of numerical simulation scenarios with the extended stochastic agent-based model, the two-moment approximation approach and Little’s law are shown to provide consistent and good FOI estimates, and to do so based on the MOI estimates obtained with the Bayesian method and bootstrap imputation approach in the presence of various types of errors commonly encountered in the collection of empirical data. The Bayesian method addresses in particular the undersampling of var genes and helps alleviate the underestimation of MOI estimates.

In high-transmission endemic regions which continue to pose major challenges to malaria elimination, clinical infections typically constitute a small fraction of the overall reservoir of infection behind the force of infection, of approximately 2% at the end of wet season and less than 1% at the end of dry one (Tiedje et al., 2022). Individuals can seek however antimalarial drug treatment in response to symptoms or the perception of transmission risk, which affects duration of infection and would interfere with our FOI estimates. Rather than attempting a correction of the duration of infection (which given the complexity of the biology involved would be difficult if not impossible at this stage), we implement instead the removal of the samples for treated individuals, and impute their MOI estimates with a bootstrap approach based on the remaining population. The proposed bootstrap imputation approach is robust to considerable fractions of individuals treated or missing MOI information, comparable to values from our field surveys in Ghana.

The application of both methods typically requires an evolving time series as it is observed over a long period of time. Because we do not have time series observation of MOI estimates for any specific individual host, we treat different individual hosts sampled at the same time point as if each one represented an observation of the process at a single time point and ignore any individual heterogeneity in their biology. This approximation turns out to be a sensible one and the bias due to interval edge effects when the system does not begin and end empty turns out to be small because the FOI estimates obtained with the two methods based on the true MOI values differ by only a very small amount from the true FOI values in the simulation output. Despite seasonality and heterogeneity in exposure risk, MOI estimates obtained under sparse sampling schemes are sufficient to recover the mean annual FOI. The inferred mean annual FOI when aggregated across all individuals faithfully reflects the total number of new P. falciparum infections acquired by the entire population. The inferred variance indicates transmission intensity and the degree of deviation from a homogeneous Poisson process, for the arrival of infections at the individual level. Thus, the flexibility and generality of these two proposed methods allow robust estimation of an important metric for malaria transmission across geographical locations and time, where single-time-point surveys are implemented, typically in wet (high-transmission) and dry (low-transmission) seasons, and information about FOI was previously unobtainable.

We illustrate the application of these two methods with molecular data from Bongo District in northern Ghana, a high-transmission endemic region for which estimation of multiplicity of infection and related FOI has been challenging. The three-round transient IRS intervention was strong and effective, resulting in a significant, more than 70% reduction in FOI.

Beyond describing basic malaria epidemiology and measuring outcomes of intervention, the FOI estimates obtained with these two proposed methods can inform process-based models for the population dynamics of complex infectious diseases. They can serve as values for priors when seeking to parameterize and validate more complex agent-based or equation-based models, in a way that reduces computational cost, improves efficiency, and reduce the degree of freedom or identifiability issues.

Although we have focused on infections caused by P. falciparum, our approach should be applicable to other Plasmodium parasites responsible for malaria transmission, including P. vivax, with distinctive life cycles. For P. falciparum, FOI is closely linked to the number of newly acquired infections from infectious mosquito bites, whereas for P. vivax, clones appearing in the blood-stream can either originate directly from recent infective mosquito bites or relapsing liver hypnozoites (Chu and White, 2016). Regardless of infection source, MOI estimates and related FOI ones should represent the number of effective infections which reach the blood stage.

As expected from estimating FOI based on MOI values, the estimation accuracy of the latter is crucial for that of the former. In high-transmission endemic regions, repertoires of var genes in co-infecting genomes can still be partially overlapping in their composition. This low overlap results in a reduced number of different var genes being sequenced and identified, which can lead to an underestimation of MOI. In addition, recent studies have identified a signature of parasite co-transmission from a single mosquito bite even in high-transmission endemic regions (Nkhoma et al., 2020; Wong et al., 2017). This conclusion was based primarily on clinical or symptomatic infections whereas asymptomatic prevalence is high and constitutes the majority of transmission reservoir for malaria (Andolina et al., 2021; Lindblade et al., 2013) in these regions. Hence, additional data needs to be collected on polygenomic relatedness across diverse epidemiological settings. These co-transmitted recombinant parasites are more related to each other than parasites from different mosquito bites, potentially resulting in a further reduced number of different var genes sequenced and identified (Wong et al., 2022; Nkhoma et al., 2020; Wong et al., 2017). Low or variable parasite densities resulting from small blood volumes, genomic DNA quality, clinical status, and/or within host dynamics, including synchronicity will all create sampling problems (Peyerl-Hoffmann et al., 2001; Bruce et al., 2000; Okell et al., 2012; Farnert et al., 1997; Färnert et al., 2008; Barry et al., 2021; Hergott et al., 2024). But these issues are common to all measures of MOI, as well as to all direct measures of FOI. Our previously proposed framework of MOI estimation could be extended in the future to account for these confounding effects in recovering the number of co-infecting strains.

A salient characteristic of malaria transmission is the saturation in FOI at high transmission, and the highly non-linear relationship between FOI and another central epidemiological measure related to transmission intensity, the entomological inoculation rate, EIR. Whereas EIR can reach values up to a few hundred or even a thousand per year, the annual FOI saturates rapidly and is typically kept below a limit of 20 (Smith et al., 2010). Although different choices are made on the way EIR and FOI are measured in the field, these cannot account for such drastically different magnitudes of the two quantities. For example, adults are used as bait to attract and trap mosquitoes, but the attack rates are typically evaluated in children, and there is an observed correlation between body size and biting rates. Transmission is highly inefficient in high-transmission endemic regions with high annual EIRs. Because FOI refers directly to detectable blood-stage infections whereas EIR concerns human-vector contact rates, the difference between these quantities is mediated primarily by immunity, or within-host dynamics, measurement bias, and heterogeneous transmission (Donovan et al., 2007; Macdonald, 1950; John et al., 2005; Doolan and Martinez-Alier, 2006). The probability of transmission from an infectious mosquito bite is commonly used in mathematical models to mediate the difference between the two transmission quantities, as a general parameter encapsulating a variety of processes. Though historically paired EIR and FOI values are only available from regions under frequent sampling schemes, our methods, when applied to geographical locations under sparse sampling schemes, could generate further information on this relationship and in so doing, provide further insights into within-host dynamics.

The saturation of FOI underlies the challenge of control and elimination efforts and the associated resilience of the transmission system, whereby relaxation of intervention leads to rapid rebounds in prevalence in high-transmission regions. In such regions of high EIR and FOI saturation, intervention efforts need to sufficiently reduce EIR to bring FOI below saturation. Only then further intervention efforts would have a significant impact by reducing the number of effective infections which advance to the blood stage. This apparent inefficiency of malaria transmission highlights why high-coverage interventions are often needed to achieve small observable impacts on individual exposure risk. Theory suggests a sharp nonlinear transition towards sustainable low transmission or elimination, including in a recent mathematical model highlighting the role of the high antigenic diversity of P. falciparum in high transmission (de Roos et al., 2023). The same molecular information used here to estimate MOI and FOI in Bongo District underlies such diversity. Our recent modeling work with the agent-based model on the resurgence dynamics of the transmission system below the saturation regime found that local systems can be highly resilient, and can compensate for the reduction in FOI following intervention with the heterogeneity in the duration of infection and more importantly the selection for infections with longer duration (Zhan et al., 2024). As a result, local systems can maintain during intervention the high prevalence and MOI levels comparable to original values.

The proposed methods are generally relevant to evaluate transmission intensity in many pathogens exhibiting multi-genomic infection as the result of large antigenic diversity including those encoding such variation by multigene families (Deitsch et al., 2009). More easily estimating changes in transmission intensity should contribute to the efficiency and evaluation of control programs across a wider range of infectious diseases.

Materials and Methods

High genetic diversity of var and the associated strain structure of limiting similarity

We present briefly here aspects of the biology of the malaria parasite P. falciparum that constitute the basis for the MOI estimation method used here to then apply the proposed FOI inference.

Individual human hosts acquire malaria infections through the infectious bites of mosquitoes, the rate of which can be either fairly constant or seasonal. In high-transmission endemic regions, individual human hosts remain susceptible to malaria re-infection (but not clinical disease) throughout their lifetime (Doolan et al., 2009). Indeed, the prevalence of asymptomatic infections is high within adults and older age groups in high-transmission regions, and these infections constitute a large transmission reservoir (Andolina et al., 2021; Lindblade et al., 2013). High asymptomatic infections result not only from high transmission rates, but also from hosts’ incomplete specific immunity conferred by high antigenic variation of the parasite (Deitsch et al., 2009).

A common strategy for immune evasion among parasites that exhibit such high asymptomatic prevalence under high transmission is to encode important variant surface antigens (VSAs) with multigene families (Deitsch et al., 2009). One important multigene family in the malaria parasite P. falciparum known as var encodes for the major VSA during the blood stage of infection, PfEMP1 (Plasmodium falciparum erythrocyte membrane protein 1) (Zhang and Deitsch, 2022; Baruch et al., 1995; Smith et al., 1995; Su et al., 1995). Each parasite carries 50-60 var genes across its 14 chromosomes and these encode different variants of this protein. Parasites express these genes largely sequentially. Under high-transmission settings, the local parasite population exhibits a large pool of var gene variants documented to reach thousands to tens of thousands (Day et al., 2017; Tiedje et al., 2022). This high level of diversity is enabled primarily by ectopic (mitotic) and meiotic recombination but also mutation (Claessens et al., 2014; Frank et al., 2008; Freitas-Junior et al., 2000; Bopp et al., 2013). Migration can also introduce new variants and contribute to the extensive diversity of a local population. Negative frequency-dependent selection (NFDS) mediated by the acquisition of specific immunity by hosts contributes to the maintenance of such a high var gene diversity, and to parasite population structure, with individual genomes (or strains or repertoires), and sets of genomes infecting different hosts (or isolates), exhibiting no, or extremely low, overlap of var gene composition. This pattern of limiting similarity has been shown to be both non-random and non-neutral (Day et al., 2017; He et al., 2018).

As hosts typically have not seen many of the VSAs within a single strain, or across different circulating strains, infection duration lengthens which increases the fitness of the parasite by increasing its chance of getting transmitted (Ruybal-Pesántez et al., 2022). Multi-genomic infection is common, with infections within individual hosts consisting of different parasite clones that have arrived successively via different mosquito bites (superinfection) or simultaneously via one single mosquito bite containing multiple clones (co-transmission) (Portugal et al., 2011; Nkhoma et al., 2020; Wong et al., 2017; Volkman et al., 2012; Anderson et al., 2000a). The duration of an infection is influenced by the overlap between the constituent var genes of that infection and the immunological memory of the infected host. The measured duration of an individual infection with P. falciparum in similar settings from empirical observations has been reported to be highly variable (Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014; Bretscher et al., 2011).

Empirical sequencing of var genes specifically concerns the DBLα tag, a small conserved ∼450bp region within var genes which encodes for the immunogenic Duffy-binding-like alpha domain of PfEMP1 (Tiedje et al., 2023; Ruybal-Pesántez et al., 2017, 2022; Day et al., 2017). Bioinformatic analyses of a large database of exon 1 sequences of var genes showed a predominantly 1-to-1 DBLα-var relationship, such that each DBLα tag typically represents a unique var gene (Tan et al., 2023). We use DBLα and var interchangeably hereafter.

Bayesian approach to MOI estimation with “varcoding”

The low to non-existent overlap of var repertoires enables estimation of MOI on the basis of the number of non-upsA DBLα types sequenced from an individual’s isolate. The original method uses a constant repertoire length or number of non-upsA DBLα types in a parasite genome to convert the number of types sequenced in an isolate to the estimated MOI. As such, the method does not take into account the measurement error (Appendix 1-Figure 5D) in this length introduced by the sampling of var genes in an infection. We recently extended the method to a Bayesian formulation that does consider this error and provides a posterior distribution of MOI values for each sampled individual (Tiedje et al., 2023). We have previously documented the major steps of the Bayesian approach, compared two ways of obtaining an MOI distribution at the population level (either pooling the maximum a posteriori MOI estimates or calculating a mixture distribution), and examined the impact of different priors on the final MOI distribution at the population level. We provide in our analyses hereafter the estimated MOI distribution at the population level obtained from a mixture distribution and using an uniform prior for individuals.

We include an evaluation of this Bayesian varcoding approach in Appendix 1-A comparison between the performance of the Bayesian approach and the original varcoding method from the simulation output.

The under-sampling of infections in estimates of MOI from empirical surveys

The empirical MOI estimates in many epidemiological studies, including ours in Bongo District from northern Ghana, rely on individuals who are microscopy-positive (Tiedje et al., 2023, 2022). Given the sensitivity of microscopy, a significant fraction of individuals who carry infections are not detected. A subset of the Ghana surveys includes in addition submicroscopic infections detected by PCR (Tiedje et al., 2023, 2022). PCR is considerably more sensitive than microscopy, detecting a higher fraction, if not 100%, of individuals with P. falciparum infections. Using surveys for which both microscopy and PCR detection were used, we estimated a conversion factor between the proportion of hosts that are microscopy-positive and those that are PCR-positive within the targeted age group, i.e., children of 1-5 years old, which is 0.711329.

For surveys with available detection by both microscopy and PCR, we can directly calculate the number of individuals with undetected infections by microscopy, i.e., individuals who are microscopy-negative but PCR-positive. For surveys with detection only via microscopy, we use the estimated conversion factor to estimate the number of individuals who are microscopy-negative but PCR-positive. In addition, because the underlying sensitivity of PCR is high but not exactly known, we assume it varies within a reasonable range of values, from a relatively low detectability, i.e, 10% of all PCR-negative individuals carrying undetected infections, to a high and perfect detectability, i.e, none of PCR-negative individuals carrying undetected infections. We calculate the number of individuals with undetected infections by PCR for each sensitivity value respectively.

After calculating the number of individuals with undetected infections, including those who are microscopy-negative but PCR-positive and those who carry infections but are PCR-negative, we then sample values from the existing MOI estimates of microscopy-positive individuals who are not under antimalarial treatment (see the following section) to represent those missing MOI estimates.

In addition to the detection limit of microscopy and PCR, a fraction of microscopy-positive individuals may not have their var information sequenced and typed successfully due to factors including low DNA quality. Similarly, we sample values from the existing MOI estimates who are not under antimalarial treatment to represent those missing MOI estimates.

Antimalarial drug treatment of infections in estimates of MOI from empirical surveys

In addition to the missing data error described in the previous section, individuals can also seek and get antimalarial treatment in response to symptoms or the perception of transmission risk. In our field surveys from northern Ghana, more than 50% of children (1-5 years) responded that they had received an antimalarial treatment in the previous two-weeks (i.e., participants that reported they were sick, sought treatment, and were provided with an antimalarial treatment) prior to their blood samples being collected for the wet/high-transmission survey before IRS (i.e., Survey 1, Appendix 1-Figure 5E) (Tiedje et al., 2022). In contrast, the fraction is much lower for the dry/low-transmission survey before IRS or for the surveys collected right after IRS (i.e., Appendix 1-Figure 5E) (Tiedje et al., 2022).

To bound reality, which here is difficult if not impossible to determine, given the complexity of the biology, we consider two opposite extremes. For the first one (assumption 1), we assume that antimalarial treatment has no impact on treated individuals’ infection status and MOI values. For the second one (assumption 2), we assume that antimalarial treatment affects infection status and MOI values in treated individuals. Thus, we remove the corresponding samples for treated individuals, and implement a bootstrap imputation approach based on the remaining population. We specifically sample values from the MOI estimates of microscopy-positive individuals who are neither under the aforementioned missing data error nor treated, to replace and represent MOI estimates for the treated individuals. This correction takes care of individuals who have used drugs and show either no infection (MOI=0) or infection (MOI>0). We show with numerical simulations that our bootstrap imputation approach is robust to considerable fractions of individuals treated, comparable to those from our survey in Ghana. Reality should lie somewhere in between these two assumptions, with antimalarial treatment affecting the infection status and MOI estimates in some individuals significantly but not in others.

The final distribution of MOI estimates at the population level, which includes values for microscopy-positive individuals, the imputed values for individuals with missing MOI information or the false negative individuals, the imputed values for individuals under malarial treatment (applicable for assumption 2), and the true zeros for uninfected individuals, is used for FOI inference.

Empirical surveys from Ghana

For this investigation we rely on empirical data collected from an interrupted time-series study in Bongo District northern Ghana, involving four age-stratified cross-sectional surveys of ∼2,000 participants per survey between 2012 and 2016. This study was undertaken to investigate the impacts of a transient three-round IRS intervention, in combination with long-lasting insecticidal nets (LLINs), on the asymptomatic P. falciparum reservoir (Tiedje et al., 2017, 2022, 2023). The surveys were undertaken either at the end of the wet/high-transmission season (i.e., October) or at the end of the dry/low-transmission season (i.e., May/June). Additionally the study can be separated into two distinct study phases: (1) Pre-IRS: two surveys prior to the IRS intervention (Survey 1 October 2012; Survey 2 May/June 2013); and (2) Right post-IRS: two surveys immediately following the three-rounds of IRS (Survey 3 October 2015; Survey 4 May/June 2016) (Appendix 1-Figure 5E). Details on the study area/population, study design, malaria control interventions (i.e., IRS and LLINs), inclusion/exclusion criteria, data collection/generation procedures, routine surveys on whom in the population having taken antimalarial treatment in a previous window of time etc. have been previously described (Tiedje et al., 2017, 2022, 2023).

Inferring FOI from MOI estimates

Mathematical consideration of malaria transmission in relation to queuing theory Mathematical models often consider that in a cohort of uninfected people acquiring and clearing infections independently, inoculations arrive according to a homogeneous Poisson process with rate equal to the mean FOI, with each inoculation exhibiting an exponentially distributed duration. It follows that at equilibrium, MOI is a Poisson variable with mean given by the mean FOI divided by the mean clearance rate (Dietz et al., 1974). In empirical settings, the arrival of inoculations often deviates from a homogeneous Poisson process due to various factors including seasonality and heterogeneity in exposure risk. Under heterogeneous biting of different hosts, the MOI distribution can be captured by a negative binomial distribution to account for a higher variance. There is no simple relationship in this case, however, between the parameters of the negative binomial distribution and the mean FOI and mean clearance rate. We perform a formal test for any deviation from Poisson homogeneity of the empirical and simulated MOI estimates, with details and results of the test provided in Appendix 1-Test of deviation from Poisson homogeneity in MOI estimates and supplementary file 3-deviationFromPoissonTest.xlsx.

Despite these complexities, the process of acquisition of infectious bites is structurally equivalent to stochastic queuing theory, which aims to explain how the length of a queue is related to the intensity of arrivals to the system, the systems’ priority schedule, and the service and waiting times. Queuing theory is indeed widely employed across multiple applications including in communications, computer architecture, and operation management. The queuing system is usually modeled by a system of differential equations (with the Kolmogorov equations). Their structure depends on specific assumptions of the model, for example, single server vs. multiple servers, stationary vs. non-stationary arrivals, different priory schedules, and so on. The three main information components of a queuing system are the length of the queue, the intensity of arrivals to the system, and the service times, all of which are coupled. In theory, knowing any two components would allow inference about the third. A simple quantitative description of the process of malaria transmission would be analogous to customers arriving to, and departing from, a service facility. In other words, each individual host is equivalent to a service facility center with a given number of servers, i.e, the carrying capacity for blood-stage infections in the parasite. Moreover, each infection is equivalent to a customer, with infections arriving at an individual host over time and each infection lasting a given amount of time before clearance. The empirical MOI estimates provide information about the length of queues. Hypothetically, when the duration of infection is known, one could infer backwards the information about the rate of arrival of infections to individuals, in other words, the FOI.

However, the duration of a single infection is difficult to determine in field samples from endemic areas, where mono-clonal infections are rare but multi-genomic infections are common. Commonly-used polymorphic markers may not have a high power to differentiate different strains that are co-infecting individual human hosts (Argyropoulos et al., 2023). Reasons include that the diversity of those polymorphic markers can be limited, with two parasites sharing the same marker but having distinct sets of antigen-encoding genes. It is thus difficult to accurately measure the emergence and absence from the peripheral blood of any individual strain. Because var genes frequently undergo ectopic recombination and change the location of their homology blocks on different chromosomes across the genome, it is difficult to assign var genes to an individual genome and to a specific chromosomal location. The difficulty of phasing precludes the integrity of individual infections and thus the ability to isolate these, as well as the tracking of their corresponding first appearance in the blood and their clearance in time. Also, the duration of infection measured in field studies is highly variable. Examples include infections within individuals of different age groups and therefore different immune histories, but also infections from different geographical locations and times and thus circulating under different transmission intensity and different genetic diversity (Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014; Bretscher et al., 2011).

We therefore propose the alternative strategy of focusing on the MOI estimates of the youngest age group in the population, that is, on those of the 1-5 year-old children. Because these children have received a much lower number of infections than the rest of the population, they have not yet seen much of the circulating diversity. It is thus safe to consider negligible the impact of immune memory accumulated from previous infections on the duration of a current infection. Therefore, infection duration in these children can be approximated by that of naive hosts. Estimates of the latter are available from a historical medical intervention in which patients with neurosyphilis were infected with malaria as a therapy. Between 1940 and 1963, 318 patients with syphilis were intentionally infected with a single strain of P. falciparum (Collins and Jeffery, 1999; Maire et al., 2006). Data on fever and the number of malaria parasites in the blood were recorded. Because those patients had never been infected with P. falciparum malaria prior to the treatment, the documented infection duration is representative of naive infections, and can serve as a proxy for infection duration in young children. With MOI estimates and the duration of infection at hand, we can then infer the FOI experienced by these children using methods from queuing theory as described next. An additional motivation for focusing on 1-5 year-old children is that they are the most vulnerable group of individuals in the population, for whom mortality, severe and clinical cases are most prominent.

A two-moment approximation for a queue of finite capacity

Analysis of multi-server models in general is notoriously difficult. Exact results are only available for some special models, including the previously mentioned M/M/c/k ones which require that both the inter-arrival times and service times are exponentially distributed. Additional examples of those special models include the M/G/c/c (including the infinite-server M/G/∞) and GI/M/c/c+r queues which require that either the inter-arrival times or the service times are exponentially distributed. As such, those special models are often too restrictive to apply to real-world queuing situations in which arrivals do not follow a homogeneous Poisson process (i.e, for which the interarrival and service times are not exponentially distributed). Application of a two-moment approximation method is examined here for the steady state analysis as proposed by Kim and Cha (Choi et al., 2005), which considers the inter-arrival and service times as independent sequences of independent and identically distributed (i.i.d.) general random variables (r.v.’s). The approximations are obtained by replacing unknown arrival- and departure-average quantities by their corresponding (well-known) time-average counterparts, which are exact for exponential inter-arrival and service times. More mathematical details and derivations can be found in Appendix 1-The mathematical details for a two-moment approximation for a queue of finite capacity.

We vary here the mean and variance parameters for inter-arrival times across wide ranges. For each combination of a specific mean and variance value, we calculate a probability density distribution for the queue length, i.e., a probability density distribution of MOI. We identify the combination of mean and variance values that maximizes the likelihood of observing the given distribution of MOI estimates in either the simulation output or the empirical data from Bongo District. We provide the details for deriving confidence intervals for the estimated mean and variance in Appendix 1-Confidence intervals for mean and variance.

In the empirical settings with seasonal transmission and heterogeneous exposure risks, the inter-arrival times of infection are likely to be complex and heterogeneous in time. By applying this two-moment approximation method, we treat those inter-arrival times as if they followed a general distribution of any kind with a specific mean and a specific variance to be inferred, and consider consecutive infections independent. Those assumptions are general and flexible enough to be applicable to a variety of empirical settings.

Finally, because the goal of this work is to utilize widely available MOI estimates sampled under sparse schemes across geographical space and time, alternative approaches such as fitting a non-homogeneous Poisson process or Gamma process to derive time-varying average arrival rates cannot be considered due to their requirements for densely sampled time-series data.

When applying this approach, one needs to specify values for the number of parallel servers (c) and the number of waiting places (r). The former corresponds to the maximum number or carrying capacity of blood-stage infections. The maximum MOI observed in the empirical data from Bongo is 20. Due to additional factors which result in a reduced number of var genes sequenced and typed but are not explicitly accounted for in current MOI estimation (see Discussion), the actual carrying capacity could be higher. For simplicity, we assume it to be 30 in the simulation. Without loss of generality, the chosen value for this quantity should not raise any issues for FOI inference based on simulation output, as long as one keeps it consistent when generating simulation output and when applying the two-moment approach to the simulation output. For FOI inference based on the empirical surveys from Bongo, we set this quantity to be either 25 or 30. Because the results of FOI inference are robust against those two values, we report the one with c = 30. Because FOI represents the number of effective infections which reach the blood stage and MOI is defined for blood-stage infections only, by default, the number of waiting places r should be set to be 0.

The mean arrival rate of infection from Little’s law

The second method is known as Little’s law (Little and Graves, 2008), which describes a relationship between the three main information components of queuing systems. This law states that the average number of items in a queuing system L equals the average arrival rate λ multiplied by the average waiting time of an item in the system, W. Reformulating Little’s law in terms of malaria transmission, the average arrival rate of infection λ equals the average number of blood-stage active infections present in an individual L divided by the average duration of infection W.

The relationship is remarkably simple and general. It requires that in the long run the average number of arrivals is less than or equal to the average number of exits, such that the queue does not overflow. The law applies however irrespective of the number of servers (here, the carrying capacity of blood-stage infections of hosts), the service time distribution (here, infection duration distribution), the distribution of inter-arrival times, the order of items’ service, and whether each server has its own queue or a single queue feeds all servers. In theory, individual human hosts can experience “overflowing” under high-transmission settings with high EIR values, but evidence has suggested that the majority of these infectious bites do not result in detectable blood-stage infections. Because FOI by definition represents the rate of arrival of effective infections acquired by human hosts that have advanced to the blood-stage, and not the rate of arrival of all infections (which is known as EIR), we consider that the requirement approximately applies in the context of FOI inference for malaria transmission. In other words, values of FOI do not include and hence should not be influenced by the loss of excessive arrivals due to various factors including immunity, measurement bias, and so on.

Little’s law holds for an evolving time series as it is observed over a long period of time. Although we do not have time series of MOI estimates for any individual host, we treat different sampled individual hosts as if each one represented an observation of the process at a single time point. This is sensible because different individual hosts are likely to receive infections in an asynchronous manner, and therefore, the aggregation of MOI information across different individual hosts can function as a proxy for a time series of MOI values for an individual host.

Conversion between FOI and the Entomological Inoculation Rate, or EIR

As an indirect proxy for transmission intensity, malariologists typically measure EIR by counting how many infectious bites are received within a time interval by a human host seated in a fixed place (Shaukat et al., 2010). The resulting value of EIR is considered a standard metric of malaria transmission. Though FOI and EIR both reflect transmission intensity, the former refers directly to detectable blood-stage infections whereas the latter concerns human-vector contact rates. Studies reporting both the annual P. falciparum EIR and FOI estimates from detailed age-stratified prevalence in cross-sectional parasitological studies, have obtained very different magnitudes for these two quantities (Smith et al., 2010). We refer here to the number of infections (in the blood stage) per infectious bite (FOI/EIR) as transmission efficiency. Several studies have shown that malaria transmission is inefficient in high-intensity settings, and there has been a long-standing debate on the reason why. Heterogeneous biting, immunity, or measurement bias have been proposed as causes for explaining this discrepancy between EIR and FOI (Donovan et al., 2007; John et al., 2005; Doolan and Martinez-Alier, 2006; Smith et al., 2010). Within host competition for resources between co-infecting strains and the resulting dominance of a subset of those strains, with the remaining others at low parasitemia, can also account for this discrepancy. Parasites may not be transmitted from an infectious mosquito to a human during the blood meal due simply to chance. We borrow a functional curve with estimated parameters under the assumption of heterogeneous transmission, which describes the highly non-linear relationship between these reported EIR-FOI pairs (Smith et al., 2010). The functional curve is in the format of the following equation with h denoting FOI and E denoting EIR. The corresponding parameters are as follows: b = 0.55, α = 4.6 and t = 43 days.

The combination of these reported EIR-FOI pairs (Smith et al., 2010) and the functional curve provides a basis for converting from one quantity to the other.

The EIR information for a subset of surveys in Bongo District from northern Ghana was obtained (Tiedje et al., 2022). Together with the FOI estimates from the two methods from queuing theory, we have the EIR-FOI pairs for empirical surveys in Bongo District. Therefore, we can address whether our EIR-pairs are consistent with those previous historical collections of EIR-FOI pairs and the functional curve with the best-fit parameter values to those collections.

Supporting information

eLife Assessment

The ability to estimate the force of infection for Plasmodium falciparum from other more directly measurable epidemiological quantities is a useful contribution to malaria epidemiology. The authors propose a method to accomplish this using genetic data from the var genes of the Pf genome and novel applications of existing methods from queueing theory. While the simulations are sophisticated, the real-world application of the method is incomplete in its analysis and would benefit from clearer articulation of the assumptions being made. Given the lack of clarity in the methods and presentation of results, it is difficult to fully assess the performance of their proposed estimation procedure.

None

Reviewer #1 (Public Review):

Summary:

In their paper, Zhan et al. have used Pf genetic data from simulated data and Ghanaian field samples to elucidate a relationship between multiplicity of infection (MOI) (the number of distinct parasite clones in a single host infection) and force of infection (FOI). Specifically, they use sequencing data from the var genes of Pf along with Bayesian modeling to estimate MOI individual infections and use these values along with methods from queueing theory that rely on various assumptions to estimate FOI. They compare these estimates to known FOIs in a simulated scenario and describe the relationship between these estimated FOI values and another commonly used metric of transmission EIR (entomological inoculation rate).

This approach does fill an important gap in malaria epidemiology, namely estimating the force of infection, which is currently complicated by several factors including superinfection, unknown duration of infection, and highly genetically diverse parasite populations. The authors use a new approach borrowing from other fields of statistics and modeling and make extensive efforts to evaluate their approach under a range of realistic sampling scenarios. However, the write-up would greatly benefit from added clarity both in the description of methods and in the presentation of the results. Without these clarifications, rigorously evaluating whether the author's proposed method of estimating FOI is sound remains difficult. Additionally, there are several limitations that call into question the stated generalizability of this method that should at minimum be further discussed by authors and in some cases require a more thorough evaluation.

Major comments:

(1) Description and evaluation of FOI estimation procedure.

a. The methods section describing the two-moment approximation and accompanying appendix is lacking several important details. Equations on lines 891 and 892 are only a small part of the equations in Choi et al. and do not adequately describe the procedure notably several quantities in those equations are never defined some of them are important to understand the method (e.g. A, S as the main random variables for inter-arrival times and service times, aR and bR which are the known time average quantities, and these also rely on the squared coefficient of variation of the random variable which is also never introduced in the paper). Without going back to the Choi paper to understand these quantities, and to understand the assumptions of this method it was not possible to follow how this works in the paper. At a minimum, all variables used in the equations should be clearly defined.

b. Additionally, the description in the main text of how the queueing procedure can be used to describe malaria infections would benefit from a diagram currently as written it's very difficult to follow.

c. Just observing the box plots of mean and 95% CI on a plot with the FOI estimate (Figures 1, 2, and 10-14) is not sufficient to adequately assess the performance of this estimator. First, it is not clear whether the authors are displaying the bootstrapped 95%CIs or whether they are just showing the distribution of the mean FOI taken over multiple simulations, and then it seems that they are also estimating mean FOI per host on an annual basis. Showing a distribution of those per-host estimates would also be helpful. Second, a more quantitative assessment of the ability of the estimator to recover the truth across simulations (e.g. proportion of simulations where the truth is captured in the 95% CI or something like this) is important in many cases it seems that the estimator is always underestimating the true FOI and may not even contain the true value in the FOI distribution (e.g. Figure 10, Figure 1 under the mid-IRS panel). But it's not possible to conclude one way or the other based on this visualization. This is a major issue since it calls into question whether there is in fact data to support that these methods give good and consistent FOI estimates.

d. Furthermore the authors state in the methods that the choice of mean and variance (and thus second moment) parameters for inter-arrival times are varied widely, however, it's not clear what those ranges are there needs to be a clear table or figure caption showing what combinations of values were tested and which results are produced from them, this is an essential component of the method and it's impossible to fully evaluate its performance without this information. This relates to the issue of selecting the mean and variance values that maximize the likelihood of observing a given distribution of MOI estimates, this is very unclear since no likelihoods have been written down in the methods section of the main text, which likelihood are the authors referring to, is this the probability distribution of the steady state queue length distribution? At other places the authors refer to these quantities as Maximum Likelihood estimators, how do they know they have found the MLE? There are no derivations in the manuscript to support this. The authors should specify the likelihood and include in an appendix an explanation of why their estimation procedure is in fact maximizing this likelihood, preferably with evidence of the shape of the likelihood, and how fine the grid of values they tested is for their mean and variance since this could influence the overall quality of the estimation procedure.

(2) Limitation of FOI estimation procedure.

a. The authors discuss the importance of the duration of infection to this problem. While I agree that empirically estimating this is not possible, there are other options besides assuming that all 1-5-year-olds have the same duration of infection distribution as naïve adults co-infected with syphilis. E.g. it would be useful to test a wide range of assumed infection duration and assess their impact on the estimation procedure. Furthermore, if the authors are going to stick to the described method for duration of infection, the potentially limited generalizability of this method needs to be further highlighted in both the introduction, and the discussion. In particular, for an estimated mean FOI of about 5 per host per year in the pre-IRS season as estimated in Ghana (Figure 3) it seems that this would not translate to 4-year-old being immune naïve, and certainly this would not necessarily generalize well to a school-aged child population or an adult population.

b. The evaluation of the capacity parameter c seems to be quite important and is set at 30, however, the authors only describe trying values of 25 and 30, and claim that this does not impact FOI inference, however it is not clear that this is the case. What happens if the carrying capacity is increased substantially? Alternatively, this would be more convincing if the authors provided a mathematical explanation of why the carrying capacity increase will not influence the FOI inference, but absent that, this should be mentioned and discussed as a limitation.

Reviewer #2 (Public Review):

Summary:

The authors combine a clever use of historical clinical data on infection duration in immunologically naive individuals and queuing theory to infer the force of infection (FOI) from measured multiplicity of infection (MOI) in a sparsely sampled setting. They conduct extensive simulations using agent-based modeling to recapitulate realistic population dynamics and successfully apply their method to recover FOI from measured MOI. They then go on to apply their method to real-world data from Ghana before and after an indoor residual spraying campaign.

Strengths:

(1) The use of historical clinical data is very clever in this context.

(2) The simulations are very sophisticated with respect to trying to capture realistic population dynamics.

(3) The mathematical approach is simple and elegant, and thus easy to understand.

Weaknesses:

(1) The assumptions of the approach are quite strong and should be made more clear. While the historical clinical data is a unique resource, it would be useful to see how misspecification of the duration of infection distribution would impact the estimates.

(2 )Seeing as how the assumption of the duration of infection distribution is drawn from historical data and not informed by the data on hand, it does not substantially expand beyond MOI. The authors could address this by suggesting avenues for more refined estimates of infection duration.

(3) It is unclear in the example how their bootstrap imputation approach is accounting for measurement error due to antimalarial treatment. They supply two approaches. First, there is no effect on measurement, so the measured MOI is unaffected, which is likely false and I think the authors are in agreement. The second approach instead discards the measurement for malaria-treated individuals and imputes their MOI by drawing from the remaining distribution. This is an extremely strong assumption that the distribution of MOI of the treated is the same as the untreated, which seems unlikely simply out of treatment-seeking behavior. By imputing in this way, the authors will also deflate the variability of their estimates.

- For similar reasons, their imputation of microscopy-negative individuals is also questionable, as it also assumes the same distributions of MOI for microscopy-positive and negative individuals.

Reviewer #3 (Public Review):

Summary:

It has been proposed that the FOI is a method of using parasite genetics to determine changes in transmission in areas with high asymptomatic infection. The manuscript attempts to use queuing theory to convert multiplicity of infection estimates (MOI) into estimates of the force of infection (FOI), which they define as the number of genetically distinct blood-stage strains. They look to validate the method by applying it to simulated results from a previously published agent-based model. They then apply these queuing theory methods to previously published and analysed genetic data from Ghana. They then compare their results to previous estimates of FOI.

Strengths:

It would be great to be able to infer FOI from cross-sectional surveys which are easier and cheaper than current FOI estimates which require longitudinal studies. This work proposes a method to convert MOI to FOI for cross-sectional studies. They attempt to validate this process using a previously published agent-based model which helps us understand the complexity of parasite population genetics.

Weaknesses:

(1) I fear that the work could be easily over-interpreted as no true validation was done, as no field estimates of FOI (I think considered true validation) were measured. The authors have developed a method of estimating FOI from MOI which makes a number of biological and structural assumptions. I would not call being able to recreate model results that were generated using a model that makes its own (probably similar) defined set of biological and structural assumptions a validation of what is going on in the field. The authors claim this at times (for example, Line 153 ) and I feel it would be appropriate to differentiate this in the discussion.

(2) Another aspect of the paper is adding greater realism to the previous agent-based model, by including assumptions on missing data and under-sampling. This takes prominence in the figures and results section, but I would imagine is generally not as interesting to the less specialised reader. The apparent lack of impact of drug treatment on MOI is interesting and counterintuitive, though it is not really mentioned in the results or discussion sufficiently to allay my confusion. I would have been interested in understanding the relationship between MOI and FOI as generated by your queuing theory method and the model. It isn't clear to me why these more standard results are not presented, as I would imagine they are outputs of the model (though happy to stand corrected - it isn't entirely clear to me what the model is doing in this manuscript alone).

(3) I would suggest that outside of malaria geneticists, the force of infection is considered to be the entomological inoculation rate, not the number of genetically distinct blood-stage strains. I appreciate that FOI has been used to explain the latter before by others, though the authors could avoid confusion by stating this clearly throughout the manuscript. For example, the abstract says FOI is "the number of new infections acquired by an individual host over a given time interval" which suggests the former, please consider clarifying.

(4) Line 319 says "Nevertheless, overall, our paired EIR (directly measured by the entomological team in Ghana (Tiedje et al., 2022)) and FOI values are reasonably consistent with the data points from previous studies, suggesting the robustness of our proposed methods". I would agree that the results are consistent, given that there is huge variation in Figure 4 despite the transformed scales, but I would not say this suggests a robustness of the method.

(5) The text is a little difficult to follow at times and sometimes requires multiple reads to understand. Greater precision is needed with the language in a few situations and some of the assumptions made in the modelling process are not referenced, making it unclear whether it is a true representation of the biology.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In their paper, Zhan et al. have used Pf genetic data from simulated data and Ghanaian field samples to elucidate a relationship between multiplicity of infection (MOI) (the number of distinct parasite clones in a single host infection) and force of infection (FOI). Specifically, they use sequencing data from the var genes of Pf along with Bayesian modeling to estimate MOI individual infections and use these values along with methods from queueing theory that rely on various assumptions to estimate FOI. They compare these estimates to known FOIs in a simulated scenario and describe the relationship between these estimated FOI values and another commonly used metric of transmission EIR (entomological inoculation rate).

This approach does fill an important gap in malaria epidemiology, namely estimating the force of infection, which is currently complicated by several factors including superinfection, unknown duration of infection, and highly genetically diverse parasite populations. The authors use a new approach borrowing from other fields of statistics and modeling and make extensive efforts to evaluate their approach under a range of realistic sampling scenarios. However, the write-up would greatly benefit from added clarity both in the description of methods and in the presentation of the results. Without these clarifications, rigorously evaluating whether the author's proposed method of estimating FOI is sound remains difficult. Additionally, there are several limitations that call into question the stated generalizability of this method that should at minimum be further discussed by authors and in some cases require a more thorough evaluation.

Major comments:

(1) Description and evaluation of FOI estimation procedure.

a. The methods section describing the two-moment approximation and accompanying appendix is lacking several important details. Equations on lines 891 and 892 are only a small part of the equations in Choi et al. and do not adequately describe the procedure notably several quantities in those equations are never defined some of them are important to understand the method (e.g. A, S as the main random variables for inter-arrival times and service times, aR and bR which are the known time average quantities, and these also rely on the squared coefficient of variation of the random variable which is also never introduced in the paper). Without going back to the Choi paper to understand these quantities, and to understand the assumptions of this method it was not possible to follow how this works in the paper. At a minimum, all variables used in the equations should be clearly defined.

We thank the reviewer for this useful comment. We have clarified the method and defined all relevant variables in the revised manuscript (Line 537-573). The reviewer correctly pointed out additional sections and equations in Choi et al., including the derivation of an exact expression for the steady-state queue-length distribution and the two-moment approximation. Since our work directly utilized the two-moment approximation, our previous manuscript included only material on that section. However, we agree that providing additional details on the derivation of the exact expression would benefit readers. Therefore, we have summarized this derivation in the revised manuscript (Line 561-564). Additionally, we clarified the method’s assumptions, particularly those involved in transitioning from the exact expression to the two-moment approximation (Line 565-570).

b. Additionally, the description in the main text of how the queueing procedure can be used to describe malaria infections would benefit from a diagram currently as written it's very difficult to follow.

We thank the reviewer for this suggestion. In the revised manuscript, we included a diagram illustrating the connection between the queueing procedure and malaria transmission (Appendix 1-Figure 8).

c. Just observing the box plots of mean and 95% CI on a plot with the FOI estimate (Figures 1, 2, and 10-14) is not sufficient to adequately assess the performance of this estimator. First, it is not clear whether the authors are displaying the bootstrapped 95%CIs or whether they are just showing the distribution of the mean FOI taken over multiple simulations, and then it seems that they are also estimating mean FOI per host on an annual basis. Showing a distribution of those per-host estimates would also be helpful. Second, a more quantitative assessment of the ability of the estimator to recover the truth across simulations (e.g. proportion of simulations where the truth is captured in the 95% CI or something like this) is important in many cases it seems that the estimator is always underestimating the true FOI and may not even contain the true value in the FOI distribution (e.g. Figure 10, Figure 1 under the mid-IRS panel). But it's not possible to conclude one way or the other based on this visualization. This is a major issue since it calls into question whether there is in fact data to support that these methods give good and consistent FOI estimates.

There seems to be some confusion on what we display in some key figures. Figures 1-2 and 10-14 (labeled as Figure 1-2 and Appendix 1-Figure 11-15 in the revised manuscript) display bootstrapped distributions including the 95% CIs, not the distribution of the mean FOI taken over multiple simulations. To estimate the mean FOI per host on an annual basis, the two proposed methods require either the steady-state queue length distribution (MOI distribution) or the moments of this distribution. Obtaining such a steady-state queue length distribution necessitates either densely tracked time-series observations per host or many realizations at the same sampling time per host. However, under the sparse sampling schemes, we only have two one-time-point observations per host: one at the end of wet/high-transmission and another at the end of dry/low-transmission. This is typically the case for empirical data, although numerical simulations could circumvent this limitation and generate such output. Nonetheless, we have a population-level queue length distribution from both simulation outputs and empirical data by aggregating MOI estimates across all sampled individuals. We use this population-level distribution to represent and approximate the steady-state queue length distribution at the individual level, not explicitly considering any individual heterogeneity due to transmission. The estimated FOI is per host in the sense of representing the FOI experienced by an individual host whose queue length distribution is approximated from the collection of all sampled individuals. The true FOI per host per year in the simulation is the total FOI of all hosts per year divided by the number of hosts. Therefore, our estimator, combined with the demographic information on population size, estimates the total number of Plasmodium falciparum infections acquired by all individual hosts in the population of interest per year. We clarified this point in the revised manuscript in the subsection of the Materials and Methods, entitled ‘Population-level MOI distribution for approximating time-series observation of MOI per host or many realizations at the same sampling time per host’ (Line 623-639).

We evaluated the impact of individual heterogeneity due to transmission on FOI inference using simulation outputs (Line 157-184, Figure 1-2 and Appendix 1-Figure 11-15). Even with significant heterogeneity among individuals (2/3 of the population receiving approximately 94% of all bites whereas the remaining 1/3 receives the rest of the bites), our methods performed comparably to scenarios with homogeneous transmission. Furthermore, our methods demonstrated similar performance for both non-seasonal and seasonal transmission scenarios.

Regarding the second point, we quantitatively assessed the ability of the estimator to recover the truth across simulations and included this information in a supplementary table in the revised manuscript (supplementary file 3-FOImethodsPerformance.xlsx). Specifically, we indicated whether the truth lies within the bootstrap distribution and provided a measure of relative deviation, which is defined as the true FOI value minus the median of the bootstrap distribution for the estimate, normalized by the true FOI value . This assessment is a valuable addition which enhances clarity, but please note that our previous graphical comparisons do illustrate the ability of the methods to estimate “sensible” values, close to the truth despite multiple sources of errors. “Close” here is relative to the scale of variation of FOI in the field and to the kind of precision that would be useful in an empirical context. From a practical perspective based on the potential range of variation of FOI, the graphical results already illustrate that the estimated distributions would be informative.

We also thank the reviewer for highlighting instances where our proposed methods for FOI inference perform sub-optimally (e.g. Figure 10, Figure 1 under the mid-IRS panel in the previous manuscript). This feedback prompted us to examine these instances more closely and identify the underlying causes related to the stochastic impact introduced during various sampling processes. These include sampling the host population and their infections at a specific sampling depth in the simulated output, matching the depth used for collecting empirical data. In addition, previously, we imputed MOI estimates for treated individuals by sampling only once from non-treated individuals. This time, we conducted 200 samplings and used the final weighted MOI distribution for FOI inference. By doing so, we reduced the impact of extreme single-sampling efforts on MOI distribution and FOI inference. In other words, some of these suboptimal instances correspond to the scenarios where the one-time sampled MOIs from non-treated individuals do not fully capture the MOI distribution of non-treated individuals. We added a section titled ‘Reducing stochastic impact in sampling processes’ to Appendix 1 on this matter (Line 841-849).

The reviewer correctly noted that our proposed methods tend to underestimate FOI (Figure 1-2, 10-14, ‘Estimated All Errors’ and ‘Estimated Undersampling of Var’ panels in the previous manuscript, corresponding to Figure 1-2 and Appendix 1-Figure 11-15 in the revised manuscript). This underestimation arises from the underestimation of MOI. The Bayesian formulation of the varcoding method does not account for the limited overlap between co-infecting strains, an additional factor that reduces the number of var genes detected per individual. We have elaborated on this matter in the Results and Discussion sections of the revised manuscript (Line 142-149, 252-256).

d. Furthermore the authors state in the methods that the choice of mean and variance (and thus second moment) parameters for inter-arrival times are varied widely, however, it's not clear what those ranges are there needs to be a clear table or figure caption showing what combinations of values were tested and which results are produced from them, this is an essential component of the method and it's impossible to fully evaluate its performance without this information. This relates to the issue of selecting the mean and variance values that maximize the likelihood of observing a given distribution of MOI estimates, this is very unclear since no likelihoods have been written down in the methods section of the main text, which likelihood are the authors referring to, is this the probability distribution of the steady state queue length distribution? At other places the authors refer to these quantities as Maximum Likelihood estimators, how do they know they have found the MLE? There are no derivations in the manuscript to support this. The authors should specify the likelihood and include in an appendix an explanation of why their estimation procedure is in fact maximizing this likelihood, preferably with evidence of the shape of the likelihood, and how fine the grid of values they tested is for their mean and variance since this could influence the overall quality of the estimation procedure.

We thank the reviewer for pointing out these aspects of the work that can be further clarified. In response, we maximized the likelihood of observing the population-level MOI distribution in the sampled population (see our responses to your previous comment c), given queue length distributions, derived from the two-moment approximation method for various mean and variance combinations of inter-arrival times. We added a new section to the Materials and Methods in the revised manuscript with an explicit likelihood formulation (Line 574-585).

Additionally, we specified the ranges for the mean and variance parameters for inter-arrival times and provided the grid of values tested in a supplementary table (supplementary file 4-meanVarianceParams.xlsx). Example figures illustrating the shape of the likelihood have also been included in Appendix 1-Figure 9. We tested the impact of different grid value choices on estimation quality by refining the grid to include more points, ensuring the FOI inference results are consistent. The results of the test are documented in the revised manuscript (Line 587-593, Appendix 1-Figure 10).

(2) Limitation of FOI estimation procedure.

a. The authors discuss the importance of the duration of infection to this problem. While I agree that empirically estimating this is not possible, there are other options besides assuming that all 1-5-year-olds have the same duration of infection distribution as naïve adults co-infected with syphilis. E.g. it would be useful to test a wide range of assumed infection duration and assess their impact on the estimation procedure. Furthermore, if the authors are going to stick to the described method for duration of infection, the potentially limited generalizability of this method needs to be further highlighted in both the introduction, and the discussion. In particular, for an estimated mean FOI of about 5 per host per year in the pre-IRS season as estimated in Ghana (Figure 3) it seems that this would not translate to 4-year-old being immune naïve, and certainly this would not necessarily generalize well to a school-aged child population or an adult population.

We thank the reviewer for this useful comment. The reviewer correctly noted the challenge in empirically measuring the duration of infection for 1-5-year-olds and comparing it to that of naïve adults co-infected with syphilis. We nevertheless continued to use the described method for the duration of infection, while more thoroughly acknowledging and discussing the limitations this aspect of the method introduces. We have highlighted this potential limitation in the Abstract, Introduction, and Discussion sections of the revised manuscript (Line 26-28, 99-103, 270-292). It is important to note that the infection duration from the historical clinical data we have relied on has been used, and is still used, in the malaria modeling community as a credible source for this parameter in untreated natural infections of malaria-naïve individuals in endemic settings of Africa (e.g. in the agent-based model OpenMalaria, see 1).

To reduce misspecification in infection duration and fully utilize our proposed methods, future data collection and sampling could prioritize subpopulations with minimal prior infections and an immune profile similar to naïve adults, such as infants and toddlers. As these individuals are also the most vulnerable, prioritizing them aligns with the priority of all intervention efforts in the short term, which is to monitor and protect the most vulnerable individuals from severe symptoms and death. We discuss this aspect in detail in the Discussion section of the revised manuscript (Line 287-292).

In the pre-IRS phase of Ghana surveys, an estimated mean FOI of about 5 per host per year indicates that a 4-year-old child would have experienced around 20 infections, which could suggest they are far from naïve. The extreme diversity of circulating var genes (2) implies, however, that even after 20 infections, a 4-year-old may have only developed immunity to a small fraction of the variant surface antigens (PfEMP1, Plasmodium falciparum erythrocyte membrane protein 1) encoded by this important gene family. Consequently, these children are not as immunologically experienced as it might initially seem. Moreover, studies have shown that long-lived infections in older children and adults can persist for months or even years, including through the dry season. This persistence is driven by high antigenic variation of var genes and associated incomplete immunity. Additionally, parasites can skew PfEMP1 expression to produce less adhesive erythrocytes, enhancing splenic clearance, reducing virulence, and maintaining sub-clinical parasitemia (3, 4, 5). The impact of immunity on infection duration with age for falciparum malaria remains a challenging open question.

Lastly, the FOI for naïve hosts is a key basic parameter for epidemiological models of complex infectious diseases like falciparum malaria, in both agent-based and equation-based formulations. This is because FOI for non-naïve hosts is typically a function of their immune status, body size, and the FOI of naïve hosts. Thus, knowing the FOI of naïve hosts helps parameterize and validate these models by reducing degrees of freedom.

b. The evaluation of the capacity parameter c seems to be quite important and is set at 30, however, the authors only describe trying values of 25 and 30, and claim that this does not impact FOI inference, however it is not clear that this is the case. What happens if the carrying capacity is increased substantially? Alternatively, this would be more convincing if the authors provided a mathematical explanation of why the carrying capacity increase will not influence the FOI inference, but absent that, this should be mentioned and discussed as a limitation.

Thank you for this question. This parameter represents the carrying capacity of the queuing system, or the maximum number of blood-stage strains with which an individual human host can be co-infected. Empirical evidence, estimated using the varcoding method, suggests this value is 20 (2), providing a lower bound for parameter c. However, the varcoding method does not account for the limited overlap between co-infecting strains, which reduces the number of var genes detected in an individual, thereby affecting the basis of MOI estimation. Additional factors, such as the synchronicity of clones in their 48-hour life cycle on alternate days (6) and within-host competition of strains leading to low-parasitemia levels (7, 8), contribute to under-sampling of strains and are not accounted for in MOI estimation (9). To address these potential under-sampling issues, we previously tested values of 25 and 30.

This time, we systematically investigated a wider range of values, including substantially higher ones: 25, 30, 40, and 60. We found that the FOI inference results are similar across these values. Figure 3 in the main text and supplementary figures (Appendix 1-Figure 16-18) illustrates these findings.

The parameter c influences the steady-state queue length distribution based on the two-moment approximation with specific mean and variance combinations, primarily affecting the distribution’s tail when customer or infection flows are high. Smaller values of c lower the maximum possible queue length, making the system more prone to “overflow”. In such cases, customers or infections may find no space available upon their arrival, hence not incrementing the queue length.

Empirical MOI distributions for high-transmission endemic regions center around 4 or 5, mostly remaining below 10, with only a small fraction between 15-20 (2). These distributions do not support parameter combinations resulting in frequent overflow for a system with c equal to 25 or 30. As one increases the value of c further, these parameter combinations would cause the MOI distributions to shift to larger values inconsistent with the empirical MOI distributions. We therefore do not expect substantially higher values for parameter c to noticeably change either the relative shape of the likelihood or the MLE.

We have included a subsection on parameter c in the Materials and Methods section of the revised manuscript (Line 596-612).

Reviewer #2 (Public Review):

Summary:

The authors combine a clever use of historical clinical data on infection duration in immunologically naive individuals and queuing theory to infer the force of infection (FOI) from measured multiplicity of infection (MOI) in a sparsely sampled setting. They conduct extensive simulations using agent-based modeling to recapitulate realistic population dynamics and successfully apply their method to recover FOI from measured MOI. They then go on to apply their method to real-world data from Ghana before and after an indoor residual spraying campaign.

Strengths:

(1) The use of historical clinical data is very clever in this context.

(2) The simulations are very sophisticated with respect to trying to capture realistic population dynamics.

(3) The mathematical approach is simple and elegant, and thus easy to understand.

Weaknesses:

(1) The assumptions of the approach are quite strong and should be made more clear. While the historical clinical data is a unique resource, it would be useful to see how misspecification of the duration of infection distribution would impact the estimates.

We thank the reviewer for bringing up the limitation of our proposed methods due to their reliance on a known and fixed duration of infection distribution from historical clinical data. Please see our response to Reviewer 1, Comment 2a, for a detailed discussion on this matter.

(2) Seeing as how the assumption of the duration of infection distribution is drawn from historical data and not informed by the data on hand, it does not substantially expand beyond MOI. The authors could address this by suggesting avenues for more refined estimates of infection duration.

We thank the reviewer for pointing out a potential improvement to our work. We acknowledge that FOI is inferred from MOI and thus depends on the information contained in MOI. However, MOI by definition is a number and not a rate parameter. FOI for naïve hosts is a fundamental parameter for epidemiological models of complex infectious diseases like falciparum malaria, in both agent-based and equation-based formulations. FOI of non-naïve hosts is typically a function of their immune status, body size, and the FOI of naïve hosts. Thus, knowing the FOI of naïve hosts helps parameterize and validate these models by reducing degrees of freedom. In this sense, we believe the transformation from MOI to FOI is valuable.

Measuring infection duration is challenging, making the simultaneous estimation of infection duration and FOI an attractive alternative, as the referee noted. This, however, would require closely monitored cohort studies or densely sampled cross-sectional surveys to reduce issues like identifiability. For instance, a higher arrival rate of infections paired with a shorter infection duration could generate a similar MOI distribution to a lower arrival rate with a longer infection duration. In some cases, incorrect combinations of rate and duration might even produce an MOI distribution that appears closer to the targeted distribution. Such cohort studies and densely sampled cross-sectional surveys have not been and will not be widely available across different geographical locations and times. This work utilizes more readily available data from sparsely sampled single-time-point cross-sectional surveys, which precludes more sophisticated derivation of time-varying average arrival rates of infections and lacks the resolution to simultaneously estimate arrival rates and infection duration. In the revised manuscript, we have elaborated on this matter and added a paragraph in the Discussion section (Line 306-309).

(3) It is unclear in the example how their bootstrap imputation approach is accounting for measurement error due to antimalarial treatment. They supply two approaches. First, there is no effect on measurement, so the measured MOI is unaffected, which is likely false and I think the authors are in agreement. The second approach instead discards the measurement for malaria-treated individuals and imputes their MOI by drawing from the remaining distribution. This is an extremely strong assumption that the distribution of MOI of the treated is the same as the untreated, which seems unlikely simply out of treatment-seeking behavior. By imputing in this way, the authors will also deflate the variability of their estimates.

We thank the reviewer for pointing out aspects of the work that can be further clarified. Disentangling the effect of drug treatment on measurements like infection duration is challenging. Since our methods rely on the known and fixed distribution of infection duration from historical data of naïve patients with neurosyphilis infected with malaria as a therapy, drug treatment can potentially violate this assumption. In the previous manuscript, we did not attempt to directly address the impact of drug treatment. Instead, we considered two extreme scenarios that bound reality, well summarized by the reviewer. Reality lies somewhere in between these two extremes, with antimalarial treatment significantly affecting measurements in some individuals but not in others. Nonetheless, the results of FOI inference do not differ significantly across both extremes.

The impact of the drugs likely depends on their nature, efficiency, and duration. We note that treatment information was collected via a routine questionnaire, with participant self-reporting that they had received an antimalarial treatment in the previous two-weeks before the surveys (i.e., participants that reported they were sick, sought treatment, and were provided with an antimalarial treatment). No confirmation through hospital or clinic records was conducted, as it was beyond the scope of the study. Additionally, many of these sick individuals seek treatment at local chemists, which may limit the relevance of hospital or clinic records, if they are even available. Consequently, information on the nature, efficiency, and duration of administrated drugs was incomplete or lacking. As this is not the focus of this work, we do not elaborate on the impact of drug treatment in the revised manuscript.

The reviewer correctly noted that this imputation might not add additional information and could reduce MOI variability. Therefore, in the revised manuscript, we reported FOI estimates with drug-treated 1-5-year-olds excluded. Additionally, we discarded the infection status and MOI values of treated individuals and sampled their MOI from non-treated microscopy-positive individuals, imputing a positive MOI for treated and uninfected individuals. We also reported FOI estimates based on these MOI values. This scenario provides an upper bound for FOI estimates. Note that we do not assume that the MOI distribution for treated individuals is the same as that for untreated individuals. Rather, we aim to estimate what their MOI would have been, and consequently, determine what the FOI per individual per year in the combined population would be, had these individuals not received antimalarial treatment. The results of FOI inference do not differ significantly between these two approaches. They can serve as general solutions to antimalarial treatment issues for others applying our FOI inference methods. These details can be found in the revised manuscript (Line 185-210, 462-484).

- For similar reasons, their imputation of microscopy-negative individuals is also questionable, as it also assumes the same distributions of MOI for microscopy-positive and negative individuals.

We thank the reviewer for this comment. The reviewer correctly noted that we imputed the MOI values for microscopy-negative but PCR-positive 1-5-year-olds by sampling from the microscopy-positive 1-5-year-olds, under the assumption that both groups have similar MOI distributions. This approach was motivated by the analysis of our Ghana surveys, which shows no clear relationship between MOI (or the number of var genes detected within an individual host, on the basis of which our MOI values were estimated) and the parasitemia levels of those hosts. Parasitemia levels underlie the difference in detection sensitivity between PCR and microscopy.

In the revised manuscript, we elaborated on this issue and included formal regression tests showing the lack of a relationship between MOI/the number of var genes detected within an individual host and the parasitemia levels of those hosts (Line 445-451, Appendix 1-Figure 7). We also described potential reasons or hypotheses behind this observation (Line 452-461).

Reviewer #3 (Public Review):

Summary:

It has been proposed that the FOI is a method of using parasite genetics to determine changes in transmission in areas with high asymptomatic infection. The manuscript attempts to use queuing theory to convert multiplicity of infection estimates (MOI) into estimates of the force of infection (FOI), which they define as the number of genetically distinct blood-stage strains. They look to validate the method by applying it to simulated results from a previously published agent-based model. They then apply these queuing theory methods to previously published and analysed genetic data from Ghana. They then compare their results to previous estimates of FOI.

Strengths:

It would be great to be able to infer FOI from cross-sectional surveys which are easier and cheaper than current FOI estimates which require longitudinal studies. This work proposes a method to convert MOI to FOI for cross-sectional studies. They attempt to validate this process using a previously published agent-based model which helps us understand the complexity of parasite population genetics.

Weaknesses:

(1) I fear that the work could be easily over-interpreted as no true validation was done, as no field estimates of FOI (I think considered true validation) were measured. The authors have developed a method of estimating FOI from MOI which makes a number of biological and structural assumptions. I would not call being able to recreate model results that were generated using a model that makes its own (probably similar) defined set of biological and structural assumptions a validation of what is going on in the field. The authors claim this at times (for example, Line 153) and I feel it would be appropriate to differentiate this in the discussion.

We thank the reviewer for this comment, although we think there is a mis-understanding on what can and cannot be practically validated in the sense of a “true” measure of FOI that would be free from assumptions for a complex disease such as malaria. We would not want the results to be over-interpreted, and we have extended the discussion of what we have done to test the methods in the revised manuscript (Line 314-328). Performance evaluation via simulation output is common and often necessary for statistical methods. These simulations can come from dynamical or descriptive models, each making their own assumptions to simplify reality. Our stochastic agent-based model (ABM) of malaria transmission, used in this study, has successfully replicated several key patterns from high-transmission endemic regions in the field, including aspects of strain diversity not represented and captured by simpler models (10).

In what sense this ABM makes a set of biological and structural assumptions that are “probably similar” to those of the queuing methods we present is not clear to us. We agree that using models with different structural assumptions from the method being tested is ideal. Our FOI inference methods based on queuing theory require the duration of infection distribution and the MOI distribution among sampled individuals. However, these FOI inference methods are agnostic to the specific biological mechanisms governing these distributions.

Another important point raised by this comment is what would be the “true” FOI value against which to validate our methods. Empirical MOI-FOI pairs from cohort studies tracking FOI directly are still lacking. Direct FOI measurements are prone to errors because differentiating new infections from the temporary absence of an old infection in the peripheral blood and its subsequent re-emergence remains challenging. Reasons for this challenge include the low resolution of the polymorphic markers used in cohort studies, which cannot fully differentiate hyper-diverse antigenic strains, and the complexity of within-host dynamics and competitive interaction of co-infecting strains (6, 8, 9). Alternative approaches also do not provide a “true” FOI estimation free from assumptions. These approaches involve fitting simplified epidemiological models to densely sampled/repeated cross-sectional surveys for FOI inference. In this case, no FOI is measured directly, and thus, there are no FOI values available for benchmarking against fitted FOI values. The evaluation or validation of these model-fitting approaches is typically based on their ability to capture other epidemiological quantities that are easier to sample or measure, such as prevalence or incidence, with criteria such as the Akaike information criterion (AIC). This type of evaluation is similar to the one done in this work. We selected FOI values that maximize the likelihood of observing the given MOI distribution. Furthermore, we paired our estimated FOI values for Ghana surveys with the independently measured EIR (Entomological Inoculation Rate), a common field measure of transmission intensity. We ensured that our resulting FOI-EIR points align with existing FOI-EIR pairs and the relationship between these quantities from previous studies. We acknowledge that, like model-fitting approaches, our validation for the field data is also indirect and further complicated by high variance in the relationship between EIR and FOI from previous studies.

Prompted by the reviewer’s comment, we elaborated on these points in the revised manuscript, emphasizing the indirect nature and existing constraints of our validation with field data in the Discussion section (Line 314-328). Additionally, we clarified certain basic assumptions of our agent-based model in Appendix 1-Simulation data.

(2) Another aspect of the paper is adding greater realism to the previous agent-based model, by including assumptions on missing data and under-sampling. This takes prominence in the figures and results section, but I would imagine is generally not as interesting to the less specialised reader. The apparent lack of impact of drug treatment on MOI is interesting and counterintuitive, though it is not really mentioned in the results or discussion sufficiently to allay my confusion. I would have been interested in understanding the relationship between MOI and FOI as generated by your queuing theory method and the model. It isn't clear to me why these more standard results are not presented, as I would imagine they are outputs of the model (though happy to stand corrected - it isn't entirely clear to me what the model is doing in this manuscript alone).

We thank the reviewer for this comment. Please refer to our response to Reviewer 2, comment (3), as we made changes in the revised manuscript regarding antimalarial drug treated individuals. We reported two sets of FOI estimates. In the first, we excluded these treated individuals from the analysis as suggested by Reviewer 2. In the second, we discarded their infection status and MOI estimates and sampling from non-treated individuals.

The reviewer correctly noted the surprising lack of impact of antimalarial treatment on MOI estimates. This pattern is indeed interesting and counterintuitive. The impact of the drugs likely depends on their nature, efficiency, and duration. We note that treatment information was collected via a routine questionnaire, with participant self-reporting that they had received an antimalarial treatment in the previous two-weeks before the surveys (i.e., participants that reported they were sick, sought treatment, and were provided with an antimalarial treatment). No confirmation through hospital or clinic or pharmacy records was conducted, as it was beyond the scope of the study. Additionally, many of these sick individuals seek treatment at local chemists, which may limit the relevance of hospital or clinic records, if they are even available. Consequently, information on the nature, efficiency, and duration of administrated drugs was incomplete or lacking. As this is not the focus of this work, we do not elaborate on the impact of drug treatment in the revised manuscript.

Regarding the last point of the reviewer, on understanding the relationship between MOI and FOI, we are not fully clear about what was meant. We are also confused about the statement on what the “model is doing in this manuscript alone”. We interpret the overall comment as the reviewer suggesting a better understanding of the relationship between MOI and FOI generated by the two-moment approximation method and the agent-based model. This could involve exploring the relationship between the moments of their distributions, possibly by fitting models such as simple linear regression models. Although this approach is in principle possible, it falls outside the focus of our work. Moreover, it would be challenging to evaluate the performance of this alternative approach given the lack of MOI-FOI pairs from empirical settings with directly measured FOI values (from large cohort studies). Nonetheless, we note that the qualitative relationship between the two quantities is intuitive. Higher FOI values should correspond to higher MOI values. Less variable FOI values should result in more narrow or concentrated MOI distributions, whereas more variable FOI values should lead to more spread-out MOI distributions. We described this qualitative relationship between MOI and FOI in the revised manuscript (Line 499-502).

As mentioned in the response to the reviewer’s previous point (1), we hope that our clarification of the basic assumptions underlying our agent-based model in Appendix 1-Simulation data helps the reviewer gain a better sense of the model. We appreciate agent-based models involve more assumptions and parameters than typical equation-based models in epidemiology, and their description can be difficult to follow. We have extended this description to rely less on previous publications. As for other ABMs, the population dynamics of the disease is followed over time by tracking individual hosts and strains. This allows us to implement specific immune memory to the large number of strains arising from the var multigene family. There is no equation-based formulation of the transmission dynamics that can incorporate immune memory in the presence of such large variation as well as recombination of the strains. We rely on this model because large strain diversity at high transmission underlies superinfection of individual hosts, and therefore, MOI values larger than one. We relied on the estimation of MOI with a method based on var gene sampling, and therefore, simulated such sampling for individual hosts (which requires an ABM and one that represents such genes and resulting strains explicitly).

(3) I would suggest that outside of malaria geneticists, the force of infection is considered to be the entomological inoculation rate, not the number of genetically distinct blood-stage strains. I appreciate that FOI has been used to explain the latter before by others, though the authors could avoid confusion by stating this clearly throughout the manuscript. For example, the abstract says FOI is "the number of new infections acquired by an individual host over a given time interval" which suggests the former, please consider clarifying.

We thank the reviewer for this helpful comment, as it is crucial to avoid any confusion regarding basic definitions. EIR, the entomological inoculation rate, is closely related to the FOI, force of infection, but they are not equivalent. EIR focuses on the rate of arrival of infectious bites and is measured as such by focusing on the mosquito vectors that are infectious and arrive to bite a given host. Not all these bites result in actual infection of the human host. Epidemiological models of malaria transmission clearly make this distinction, as FOI is defined as the rate at which a host acquires infection. This definition comes from more general models of the population dynamics of infectious diseases. For simpler diseases without super-infection, the typical SIR models define FOI as the rate at which a susceptible individual becomes infected. In the context of malaria, FOI refers to the number of new infections acquired by an individual host over a given time interval. This distinction between EIR and FOI is the reason why studies have investigated their relationship, with the nonlinearity of this relationship reflecting the complexity of the underlying biology and how host immunity influences the outcome of an infectious bite.

We added “blood-stage strains” to the definition of FOI in the previous manuscript, as pointed out by the reviewer, for the following reason. After an individual host acquires an infection/strain from an infectious mosquito bite, the strain undergoes a multi-stage life cycle within the host, including the liver stage and asexual blood stage. Liver-stage infections can fail to advance to the blood stage due to immunity or exceeding the blood-stage carrying capacity. Only active blood-stage infections are detectable in all direct measures of FOI. Quantities used in indirect model-fitting approaches for estimating FOI are also based on or reflect these blood-stage strains/infections. Only these blood-stage strains/infections are transmissible to other individuals, impacting disease dynamics. Ultimately, the FOI we seek to estimate is the one defined as specified above, as well as in both the previous and revised manuscripts, consistent with the epidemiological literature. We expanded on this point in the revised manuscript (Line 641-656).

(4) Line 319 says "Nevertheless, overall, our paired EIR (directly measured by the entomological team in Ghana (Tiedje et al., 2022)) and FOI values are reasonably consistent with the data points from previous studies, suggesting the robustness of our proposed methods". I would agree that the results are consistent, given that there is huge variation in Figure 4 despite the transformed scales, but I would not say this suggests a robustness of the method.

We thank the reviewer for this comment and have modified the relevant sentences to use “consistent” instead of “robust” (Line 229-231).

(5) The text is a little difficult to follow at times and sometimes requires multiple reads to understand. Greater precision is needed with the language in a few situations and some of the assumptions made in the modelling process are not referenced, making it unclear whether it is a true representation of the biology.

We thank the reviewer for this comment. As mentioned in the response to Reviewer 1 and in response to your previous points, we have shortened, reorganized and rewritten parts of the text in the revised manuscript to improve clarity and readability.

Reviewer #1 (Recommendations For The Authors):

Minor comments:

Bar graphs in Figures 6 and 7 are not an appropriate way to rigorously compare whether your estimated MOI (under different approaches) is comparable to your true MOIs. Particularly in Figure 6 it is very difficult to clearly compare what is going on. If anything in Figure 7 it looks like as MOI gets higher, Bayesian methods and barcoding are overestimating relative to the truth. The large Excel file that shows KS statistics could be better summarized (and include p-values not in a separate table) and further discussion of how these methods perform on metrics other than the mean value would be important given that MOI distributions can be heavily right skewed and these high MOI values contain a large proportion of genetic diversity which can be highly informative for the purposes of this estimation.

We appreciate the reviewer’s comment. It appears there may have been some misinterpretation of the pattern in Figure 7 in the previous manuscript. We believe the reviewer meant “as MOI gets higher, Bayesian methods and varcoding are UNDERESTIMATING relative to the truth” rather than “OVERESTIMATING”.

We agree with the reviewer that the comparison of MOI distributions can be improved. To better quantify the difference between the MOI distribution from the original varcoding method and its Bayesian formulation relative to true MOIs, we replaced the KS test conducted in the previous manuscript with two alternative, more powerful tests: the Cramer-von Mises Test and the Anderson-Darling Test. The Cramer-von Mises Test quantifies the sum of the squared differences between the two cumulative distribution functions, while the Anderson-Darling Test, a modification of the Cramer-von Mises Test, gives more weight to the tails of the distribution, as noted by the reviewer. We have summarized the results, including test statistics and their associated p-values, in a supplementary table (Line 135-149, Line 862-883, supplementary file 1-MOImethodsPerformance.xlsx and supplementary file 7-BayesianImprovement.xlsx).

Throughout the text the authors use "consistent" to describe their estimation of FOI, I know this is meant in the colloquial use of the word but consider changing this word to replicable or something similar. When talking about estimators, usually, consistency implies asymptotic convergence in probability which we do not know whether the proposed estimator does.

We thank the reviewer for this suggestion. We changed “consistent” to “replicable” in the revised manuscript.

I think there is an issue with the numbering of the figures, they are just numbered continuously between the main text and appendix between 1 and 15, but in the text, there is a different numbering system between the main text and appendix figures.

We thank the reviewer for this comment. We have double-checked to ensure that the numbering of the figures is consistent with the text in the revised manuscript. Figures are numbered continuously between the main text and the appendix. When referring to these figures in the text, we provide a prefix (i.e., Appendix 1) indicating whether the figure is in the main text or Appendix 1, followed by the figure number.

The description of the bootstrap for 95% CI is a bit sparse, did bootstrap distributions look symmetric? If not did authors use a skewness adjustment to ensure good coverage? Also, is the bootstrap unit of resampling at the individual level, the simulation scenario level, population level?

We checked the bootstrap distributions and calculated their skewness. The majority fall within the range of -0.5 to 0.5, with a few exceptions falling within the range of 0.5-0.75 (supplementary file 6-FOIBootstrapSkewness.xlsx). We considered them as fairly symmetric and thus did not use a skewness adjustment.

In Figures 8 and 9 the x-axes seem to imply there are both the true and estimated MOI distributions on the plot but only 1 color of grey is clearly visible. If there are 2 distributions the color or size needs to be changed or if not consider re-labeling the x-axis.

We thank the reviewer for this comment. There was a mistake in the x-axis labels in Figure 8 and 9. Only the estimated MOI distributions were shown because the true ones are not available for the Ghana field surveys. The labels should simply be “Estimated MOIvar”.

Reviewer #2 (Recommendations For The Authors):

(1) Throughout the results section there are lots of vague statements such as "differ only slightly", "exhibit a somewhat larger, but still small, difference", etc. Please include the exact values and ranges within the text where appropriate because it can be difficult to discern from the figure.

We thank the reviewer for this useful comment. In the revised manuscript, we have provided exact values and ranges where appropriate (supplementary file 1- MOImethodsPerformance.xlsx, supplementary file 3- FOImethodsPerformance.xlsx, and supplementary file 7-BayesianImprovement.xlsx).

(2) Truncate decimals to 2 places.

We thank the reviewer for this comment. In the revised manuscript, we have truncated decimals to two places where applicable.

(3) The queueing theory notation in the methods section is unfamiliar, specifically things like "M/M/c/k", please define the variables used.

We thank the reviewer for this useful comment. In the revised manuscript, we have defined all the variables used. Please refer to our responses to Reviewer 1 Point (1) a.

Reviewer #3 (Recommendations For The Authors):

(1) The work takes many of the models and data from a previous paper published in eLife in 2023 (the 4 most senior authors of this previous manuscript are the 4 authors of the current manuscript). This previous paper introduced some new terminology "census population" which was highlighted as being potentially confusing by 2 of the 3 reviewers of the original article. This was somewhat rebuffed by the authors, though their response was ambiguous about whether the terminology would be changed in any potential future revision. The census population terminology does not appear in this manuscript, though the same data is being used. Publication of similar papers with the same data and different terminology could generate confusion, so I would encourage authors to be consistent and make sure the two papers are in line. To this end, it feels like this paper would be better suited to be classified as a "Research Advances" on this original manuscript and linked, which is a nice functionality that eLife offers.

We thank the reviewer for this comment, but we do not think our work would fall under the criteria of “Research Advances” based on our previous paper pointed out by the reviewer. The reviewer correctly noted that the current work and the previous paper used the same datasets. However, they have different goals and are not related in terms of content.

The previous paper examined how epidemiological quantities and diversity measurements of the local parasite population change following the initiation of effective control interventions and subsequently as this control wanes. These quantities included MOI and census population size (MOI was estimated using the Bayesian formulation of the varcoding method, and the census population size was derived from summing MOIvar across individuals in the human population). In contrast, our current work focused on a different goal: inferring FOI based on MOI. We proposed two methods from queuing theory and illustrated them with MOI estimates obtained with the Bayesian formulation of the "varcoding" method. Although the method applied to estimate MOI is indeed the same as that of the paper mentioned by the reviewer, the proposed methods should be applicable to MOI estimates obtained in any other way, as stated in the Abstract in the previous manuscript. That is, the methods we present in the current paper are independent from the way the MOI estimation has been carried out. Our results are not about the MOI values themselves but rather on an illustration of the methods for converting those MOI values to FOI. In fact, there are different ways to obtain MOI estimates for Plasmodium falciparum (9). The most common approach for determining MOI involves size-polymorphic antigenic markers, such as msp1, msp2, msp3, glurp, ama1, and csp. Similarly, microsatellites, also termed simple sequence repeat (SSR), are another type of size-polymorphic marker that can be amplified to estimate MOI by determining the number of alleles detected. Combinations of genome-wide single nucleotide polymorphisms (SNPs) have also been used to estimate MOI.

The result section of the current manuscript begins by evaluating how different kinds of errors/sampling limitations affect the estimation of MOI using the Bayesian formulation of the varcoding method. Only that brief section, which is not the core or primary objective of the manuscript, could be considered an extension and an advancement related to the other paper. We considered the effect of these errors on the resulting estimates of FOI.

We further note that, as the reviewer pointed out, the census population size is not utilized at all in our current work. We are unclear on why this quantity is mentioned here. Our previous paper has been revised and can be found in eLife as such. We have not changed this terminology and have provided a clear explanation for why we chose it. The reviewer seems to have read the previous response to version 1 posted on December 28, 2023 (Note that version 2 and the associated response was posted on November 20, 2024). Regardless, this is not the place for a discussion on another paper on a quantity that is irrelevant to the current work being reviewed.

We understand that the reviewer’s impression may have been influenced by the previous emphasis on the Bayesian formulation of the varcoding method in our manuscript. With the reorganization and rewriting of parts of the manuscript, we hope the revised version will clearly convey the central goal of our work.

(2) Similar statements that could be toned down. 344 ".... two-moment approximation approach and Little's law are shown to provide consistent and good FOI estimates,.....", 374 "Thus, the flexibility and generality of these two proposed methods allow robust estimation of an important metric for malaria transmission"

We thank the reviewer for this comment. We have modified the descriptive terms for the performance of our methods. Please also refer to our responses to Reviewer 1, Point (1) c and your previous Point (1).

(3) Various assumptions seem to have been made which are not justified. For example, heterogeneous mixing is defined as 2/3rd of the population receives 90% of the bites. A reference for this would be good.

In this work, we considered heterogenous transmission arising from 2/3 of the population receiving approximately 94% of all bites, because we believe this distribution introduces a reasonable and sufficient amount of heterogeneity in exposure risk across individuals. We are not aware of field studies justifying this degree of heterogeneity.

(4) The work assumes children under 5 have no immunity (Line 648 says "It is thus safe to consider negligible the impact of immune memory accumulated from previous infections on the duration of a current infection." ). Is there supporting evidence for this and what would happen if this wasn't the case?

We thank the reviewer for this helpful comment. Please refer to our responses to Reviewer 1 Point (2) a.

(5) Similarly, there are a few instances of a need for more copy-editing. The text says "We continue with the result of the heterogeneous exposure risk scenarios in which a high-risk group ( 2/3 of the total population) receives around 94% of all bites whereas a low-risk group ( 1/3 of the total population) receives the remaining bites (Appendix 1-Figure 5C)." whereas the referenced caption says "For example, heterogeneous mixing is defined as 2/3rd of population receives 90% of the bites."

We believe there was a misinterpretation of the legend caption. In the referenced caption, we stated “2/3rd of population receives MORE THAN 90% of the bites”, which aligns with “around 94% of all bites”. Nonetheless, to maintain consistency in the revised manuscript, we have updated the description to uniformly state “approximately 94% of all bites” throughout.

(6) The term "measurement error" is used to describe the missing potential under-sampling of var genes. Given this would only go one way isn't the term "bias" more appropriate?

We understand that, in general English, “bias” might seem more precise for describing a deviation in one direction. However, in malaria epidemiology and in models for malaria and other infectious diseases, “measurement error” is a general term that describes deviations introduced in the process of measurement and sampling, which can confound or add noise to the true values being collected. This term is commonly used, and we have adhered to it in the revised manuscript.

(7) Line 739 "Though FOI and EIR both reflect transmission intensity, the former refers directly to detectable blood-stage infections whereas the latter concerns human-vector contact rates." In my mind this is not true, the EIR is the number of potentially invading parasites (a contact rate between parasites in mosquitoes and humans if you will). The human-vector contact rate is the human biting rate.

We thank the reviewer for this comment. We have clarified the definition regarding FOI and EIR in our response to your previous comment (3) and in the revised manuscript. We agree that the term “human-vector contact rates” was not precise enough for EIR. We intended “human-infectious vector contact rates”, and we have updated the text to reflect this change (Line 644-645).

References and Notes

(1) Maire, N. et al. A model for natural immunity to asexual blood stages of Plasmodium falciparum malaria in endemic areas. Am J Trop Med Hyg., 75(2 Suppl):19-31 (2006).

(2) Tiedje, K. E. et al. Measuring changes in Plasmodium falciparum census population size in response to sequential malaria control interventions. eLife, 12 (2023).

(3) Andrade C. M. et al. Infection length and host environment influence on Plasmodium falciparum dry season reservoir. EMBO Mol Med.,16(10):2349-2375 (2024).

(4) Zhang X. and Deitsch K. W. The mystery of persistent, asymptomatic Plasmodium falciparum infections, Current Opinion in Microbiology, 70:102231 (2022).

(5) Tran, T. M. et al. An Intensive Longitudinal Cohort Study of Malian Children and Adults Reveals No Evidence of Acquired Immunity to Plasmodium falciparum Infection, Clinical Infectious Diseases, 57(1):40–47 (2013).

(6) Farnert, A., Snounou, G., Rooth, I., Bjorkman, A. Daily dynamics of Plasmodium falciparum subpopulations in asymptomatic children in a holoendemic area. Am J Trop Med Hyg., 56(5):538-47 (1997).

(7) Read, A. F. and Taylor, L. H. The Ecology of Genetically Diverse Infections, Science, 292:1099-1102 (2001).

(8) Sondo, P. et al. Genetically diverse Plasmodium falciparum infections, within-host competition and symptomatic malaria in humans. Sci Rep 9(127) (2019).

(9) Labbe, F. et al. Neutral vs. non-neutral genetic footprints of Plasmodium falciparum multiclonal infections. PLoS Comput Biol, 19(1) (2023).

(10) He, Q. et al. Networks of genetic similarity reveal non-neutral processes shape strain structure in Plasmodium falciparum. Nat Commun 9(1817) (2018).

Version 2

Abstract

Abstract

High multiplicity of infection or MOI, the number of genetically distinct parasite strains co-infecting a single human host, characterizes infectious diseases including falciparum malaria at high transmission. This high MOI accompanies high asymptomatic Plasmodium falciparum prevalence despite high exposure, creating a large transmission reservoir challenging intervention. High MOI and asymptomatic prevalence are enabled by immune evasion of the parasite achieved via vast antigenic diversity. Force of infection or FOI, the number of new infections acquired by an individual host over a given time interval, is the dynamic sister quantity of MOI, and a key epidemiological parameter for monitoring antimalarial interventions and assessing vaccine or drug efficacy in clinical trials. FOI remains difficult, expensive, and labor-intensive to accurately measure, especially in high-transmission regions, whether directly via cohort studies or indirectly via the fitting of epidemiological models to repeated cross-sectional surveys. We propose here the application of queuing theory to obtain FOI from MOI, in the form of either a two-moment approximation method or Little’s Law. We illustrate these two methods with MOI estimates obtained under sparse sampling schemes with the “varcoding” approach. The two methods use infection duration data from naive malaria therapy patients with neurosyphilis. Consequently, they are suitable for FOI inference in subpopulations with a similar immune profile and the highest vulnerability, for example, infants or toddlers. Both methods are evaluated with simulation output from a stochastic agent-based model, and are applied to an interrupted time-series study from Bongo District in northern Ghana before and immediately after a three-round transient indoor residual spraying (IRS) intervention. The sampling of the simulation output incorporates limitations representative of those encountered in the collection of field data, including under-sampling of var genes, missing data, and antimalarial drug treatment. We address these limitations in MOI estimates with a Bayesian framework and an imputation bootstrap approach. Both methods yield good and replicable FOI estimates across various simulated scenarios. Applying these methods to the subpopulation of children aged 1-5 years in Ghana field surveys shows over a 70% reduction in annual FOI immediately post-intervention. The proposed methods should be applicable to geographical locations lacking cohort or cross-sectional studies with regular and frequent sampling but having single-time-point surveys under sparse sampling schemes, and for MOI estimates obtained in different ways. They should also be relevant to other pathogens whose immune evasion strategies are based on large antigenic variation resulting in high MOI.

Introduction

Despite substantial intervention efforts, falciparum malaria in high-transmission regions remains a major public health concern, causing mortality among young children and considerable economic burden, particularly in sub-Saharan Africa (World Health Organization, 2023). It is thus still important to robustly evaluate the effects of intervention efforts in these regions, including on transmission intensity. Force of infection (FOI), defined as the number of new Plasmodium falciparum infections acquired by an individual host over a given time interval, is a key metric reflecting the risk of infection and clinical episodes (Mueller et al., 2012). While other metrics may describe the relationship between transmission intensity and the burden of malaria illness on global or continental scales (Carneiro et al., 2010; Beier et al., 1994), FOI can relate local variation in malaria burden to transmission (Mueller et al., 2012). Despite being a key epidemiological parameter for malaria surveillance, FOI is difficult, expensive, and labor-intensive to measure accurately, whether directly through cohort studies or indirectly through the fitting of epidemiological models. As molecular tools for parasite genomics become more readily available, they are contributing to new approaches. In particular, molecular advances are providing a basis for estimating a sister “static” quantity, the multiplicity of infection (MOI), defined as the number of genetically distinct parasite strains co-infecting a single human host (Zhong et al., 2018; Chang et al., 2017; Ruybal-Pesántez et al., 2022; Tiedje et al., 2022; Labbé et al., 2023). We can therefore ask whether we can go further and on the basis of MOI obtain the dynamical, rate, quantity of force of infection?

Early efforts to directly measure FOI included clearing infections and observing time to reinfection (Macdonald, 1950; Pull and Grab, 1974; Msuya and Curtis, 1991; Alonso et al., 2004). Molecular approaches now enable the genotyping of individual parasite infections (Mueller et al., 2012; Hofmann et al., 2017), but differentiating new infections from the temporary absence of an old infection in the peripheral blood and its subsequent re-emergence (Daubersies et al., 1996) remains challenging due to the low resolution of polymorphic markers and the complex within-host dynamics of malaria infection. Determining molecular FOI remains challenging, labor-intensive, and costly, requiring close long-term monitoring and genotyping of a large cohort.

Alternatives to direct measurements involve cross-sectional surveys with FOI estimated by fitting simple epidemiological models (Bekessy et al., 1976; Muench, 1959; Smith and Vounatsou, 2003; Felger et al., 2012; Mugenyi et al., 2017). These model-fitting procedures require empirical data sampled regularly and frequently, such as age-stratified large cohorts sampled six times a year, to account for FOI and infection duration heterogeneity (Laishram et al., 2012; Simpson et al., 2002; Langhorne et al., 2008; Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014) influenced by the interplay between host immunity and antigen composition of infections (Piper et al., 1999; Molineaux et al., 2002; Doolan et al., 2009; Barry et al., 2007). Additionally, they may face identifiability issues with model parameters (Mugenyi et al., 2017). Hence, these indirect measurements share limitations with direct ones.

Due to the described challenges, FOI has not become a readily available epidemiological quantity across geographical locations and times. In contrast, various approaches have been proposed to estimate MOI from clinical samples using size-polymorphic antigenic markers, microsatellites, and panels of biallelic single nucleotide polymorphisms (SNPs) (Felger et al., 1994; Konaté et al., 1999; Anderson et al., 2000; Daniels et al., 2008; Chang et al., 2017). Because Plasmodium parasites reproduce asexually during haploid stages within human hosts (Guttery et al., 2012), polymorphic genotypes indicate multiclonal infection. As methods for disentangling the molecular complexity of natural parasite infections across various transmission settings emerge and mature, MOI becomes more commonly and easily surveyed across space and time. Although MOI remains one of the most frequently used genetic metrics of parasite transmission (Arnot, 1998; Sondo et al., 2020), it is by definition a number and not a rate.

A natural correlation exists however between MOI and FOI, mediated by infection duration, offering an opportunity to convert MOI into FOI. This conversion has been challenging due to MOI estimates often being obtained from sparse sampling schemes, like single-time-point surveys at the end of wet (high-transmission) and dry (low-transmission) seasons (Tiedje et al., 2022; Abukari et al., 2019). Although these MOI estimates have been useful (Earland et al., 2019a; Lee et al., 2006), the sparse sampling scheme has limited their translation into a transmission rate.

In this work, we propose using these MOI estimates for FOI inference with two mathematical modeling frameworks based on queuing theory (Choi et al., 2005; Little and Graves, 2008). The two methods require infection duration values, for which we relied on data from naive malaria therapy patients with neurosyphilis (Collins and Jeffery, 1999; Maire et al., 2006). Consequently, our approach is suited for FOI inference in subpopulations with a similar immune profile and the highest vulnerability, for example, infants or toddlers. We evaluate them through numerical simulation of an extended stochastic agent-based model (ABM) (He et al., 2018; Zhan et al., 2024b) for both closed and open systems with constant or seasonal transmission. We consider both homogeneous and heterogeneous transmission, with the latter including a high-risk group of hosts that receives the majority of the infectious bites. We also examine different statistical distributions for the times between local transmission events. We incorporate limitations representative of those encountered in the collection of field data into the sampling of simulation output, including undersampling of var genes, missing data, and antimalarial drug treatment. We address these limitations in MOI estimates with a Bayesian framework and an imputation bootstrap approach. Both methods provide good and replicable FOI estimates across simulated scenarios. After validating with simulations, we apply the two methods to empirical data from an interrupted time-series study in Bongo District, northern Ghana, that involved a three-round transient indoor residual spraying (IRS) intervention (Tiedje et al., 2022, 2023). We focus on children aged 1-5 years whose immune profiles are closer to naive patients than the rest of the population, an aspect we discuss later. We then explore the relationship between FOI and another commonly used surrogate for transmission intensity, the entomological inoculation rate or EIR, defined as the number of infectious bites received by an individual over a given time period (Shaukat et al., 2010). This relationship underscores the challenges of relating transmission to malaria burden at local scales and achieving substantial reductions in transmission in high-transmission regions.

Results

The Bayesian formulation of the varcoding method, combined with the bootstrap imputation approach, effectively addresses sampling limitations often encountered in collecting field data for MOI estimates

Because our FOI inference relies on MOI estimates, we first investigate the impact of various sampling limitations on these estimates. We use the Bayesian formulation of the varcoding method and the bootstrap imputation approach (Materials and Methods) to address sampling limitations often encountered in collecting field data for MOI estimates: under-sampling of var genes, missing data, antimalarial drug treatment, and their combination. For this investigation, we utilize simulation output from an agent-based model (ABM) of malaria transmission with known true MOI values (He et al., 2018; Zhan et al., 2024b). Key assumptions and processes for the ABM, and the experimental design for simulation output are summarized in Appendix 1-Simulation data, with illustrations in Appendix 1-Figure 5-6.

Our results indicate that MOI estimates obtained using the Bayesian formulation of the varcoding method and the bootstrap imputation approach closely match true MOI values in most cases. To assess the difference between estimated and true MOI distributions, we use the Cramer-von Mises and Anderson-Darling tests. The Cramer-von Mises test quantifies the sum of the squared differences between cumulative distribution functions, while the Anderson-Darling test, a modification of the former, gives more weight to the tails of distributions. Most p-values are non-significant (> 0.05), indicating insufficient evidence to conclude that the estimated and true distributions differ, with few exceptions in pre-IRS or low-coverage IRS scenarios. The Bayesian formulation of varcoding tends to underestimate MOI because it does not account for the limited overlap between co-infecting strains, which reduces the number of var genes detected per individual. This underestimation bias can be more pronounced in certain high-transmission situations where many hosts have a high true MOI, such as the aforementioned exceptions in pre-IRS or low-coverage IRS scenarios. Consequently, this underestimation in MOI leads to an underestimation of FOI estimates, as described in the next section. Detailed results of both tests are provided in the supplementary file 1-MOImethodsPerformance.xlsx.

The distributions of MOI estimates across different surveys in Ghana and from the simulated outputs are not Poisson-distributed (Appendix 1-Test of deviation from Poisson homogeneity in MOI estimates, supplementary file 2-deviationFromPoissonTest.xlsx) (Potthoff and Whittinghill, 1966; Lloyd-Smith et al., 2005). This deviation suggests that either the arrival of infections deviates from a homogeneous Poisson process, or some latent heterogeneity causes over-dispersion in infection duration. This deviation makes converting MOI to FOI more challenging, but the proposed methods are flexible and applicable (Materials and Methods-Inferring FOI from MOI estimates).

The Two-Moment Approximation and Little’s Law methods give good and replicable estimates for FOI across various simulated scenarios

We begin with a homogeneous exposure risk scenario for seasonal transmission in a closed system (Appendix 1-Figure 6A-C, and Appendix 1-Simulation data). The times between local transmission events follow a Gamma distribution (Appendix 1-Figure 6C). We infer FOI using true MOI values and MOI estimates obtained via the Bayesian formulation or the bootstrap imputation approach, accounting for one or all sampling limitations. Details on deriving the confidence intervals are provided in Appendix 1-Confidence intervals for FOI inference.

Across pre-IRS and three IRS coverage levels, the 95% confidence intervals and the bootstrap distributions of FOI estimates are narrow. FOI estimates based on true MOI values, and those corrected for missing data and antimalarial treatment, closely match true FOI values (Figure 1). FOI estimates based on MOI estimates corrected for under-sampling of var genes and those corrected for all sampling limitations show slight underestimations (Figure 1), due to the underestimation of MOI, as described in the previous section.

Confidence intervals for estimated mean FOI values in simulated scenarios of homogeneous exposure risk, before and during IRS interventions at three different coverage levels.

The times between local transmission events follow a Gamma distribution, with seasonal transmission in a closed system. FOI estimates are derived from true MOI values and MOI estimates obtained through the Bayesian formulation or the bootstrap imputation approach correcting for all or individual sampling limitations. The true mean FOI per host per year is computed by dividing the total number of infections acquired by the population by the total number of hosts in the population. Each boxplot shows minimum, 5% quantile, median, 95% quantile, and maximum values.

To quantify the difference between inferred and true FOI values, we check if the true FOI lies within the bootstrap distribution and calculate the relative deviation, which is defined as the true FOI value minus the median of the bootstrap distribution for the estimate, normalized by the true FOI value. These details are in supplementary file 3-FOImethodsPerformance.xlsx.

We continue with a heterogeneous exposure risk scenario in which a high-risk group ( of the population) receives approximately 94% of bites, while a low-risk group ( of the population) receives the remainder (Appendix 1-Figure 6C). Transmission is seasonal and the system is semi-open (Appendix 1-Figure 6A-B, and Appendix 1-Simulation data). The times between local transmission events are Gamma-distributed (Appendix 1-Figure 6C). As before, FOI estimates across pre-IRS and three IRS coverage levels show narrow 95% confidence intervals and bootstrap distributions close to true FOI values (Figure 2), with slight underestimations for FOI estimates based on MOI estimates corrected for under-sampling of var genes and those corrected for all sampling limitations.

Confidence intervals for estimated mean FOI values in simulated scenarios of heterogeneous exposure risk, before and during IRS interventions at three different coverage levels.

The times between local transmission events follow a Gamma distribution, with seasonal transmission in a semi-open system. FOI estimates are derived from true MOI values and MOI estimates obtained through the Bayesian formulation of the varcoding method or the bootstrap imputation approach correcting for all or individual sampling limitations. The true mean FOI per host per year is computed by dividing the total number of infections acquired by the population by the total number of hosts in the population. Each boxplot shows minimum, 5% quantile, median, 95% quantile, and maximum values.

The performance of the two methods across additional simulated scenarios is shown in Appendix 1-Figure 11-15.

The Two-Moment Approximation and Little’s Law methods give replicable FOI estimates for empirical surveys conducted in the Bongo District of northern Ghana

After validating with simulations, we apply the two methods to empirical surveys in the Bongo District of northern Ghana. We first derive their MOI estimates. Due to the high but imperfect detection power of PCR, we assume three levels of sensitivity: 0% (high detectability), 5% (mid detectability), and 10% (low detectability) of PCR-negative individuals carrying infection (Materials and Methods-The under-sampling of infections in empirical surveys).

Antimalarial treatment, sought in response to symptoms or perceived transmission risk, can impact infection duration. Consequently, this may potentially violate the assumption that the two methods use infection duration data from naive malaria therapy patients with neurosyphilis. We address this issue by either excluding treated individuals from the analysis or by discarding their infection status and MOI estimates, instead sampling from non-treated individuals. Since the latter samples non-zero MOIs for these treated and uninfected individuals, it results in an upper bound for FOI estimates. Note that in this latter case, we do not assume that the MOI distribution for treated individuals is the same as that for untreated individuals. Rather, we aim to estimate what their MOI would have been, and consequently, determine what the FOI per individual per year in the combined population would be, had these individuals not received antimalarial treatment. Further details can be found in Materials and Methods-Antimalarial drug treatment of infections in empirical surveys.

Next, we apply the two-moment approximation and Little’s Law methods to derive FOI estimates from empirical MOI estimates. Both methods yield replicable FOI estimates. The FOI estimates are similar across the three PCR sensitivity levels and the two approaches for handling treated individuals. The 95% confidence intervals and the full sampling distributions from boot-strap analysis are concentrated (Figure 3, Appendix 1-Figure 16-18). Notably, there is a significant reduction in FOI, exceeding 70%, indicating that the three-round IRS intervention, although transient, was highly effective.

Confidence intervals for the estimated mean FOI values in Ghana surveys before and immediately after a transient three-round IRS intervention

(A) The estimated FOI values when excluding these treated individuals from the analysis. (B) The estimated FOI values when discarding the infection status and MOI estimates of treated individuals and sampling from non-treated ones. Since this case samples non-zero MOIs for these treated and uninfected individuals, it results in an upper bound for FOI estimates. Each boxplot shows minimum, 5% quantile, median, 95% quantile, and maximum. The value of c is set to 30. FOI estimates with other values of c can be found in Appendix 1-Figure 16-18.

The inferred FOI and directly measured EIR from the Ghana surveys align with the relationship between these two quantities in previous studies

We plot the measured annual EIR against the estimated annual FOI, and the transmission efficiency (the ratio between FOI and EIR) against the measured annual EIR from previous field studies summarized by (Smith et al., 2010) (Figure 4). In these studies, FOI estimation was based on fitting a simple epidemiological model to age-stratified prevalence data from cross-sectional parasitological studies (Smith et al., 2010; Pull and Grab, 1974), while EIR was estimated using various methods such as exit bait collection, human bait collection, pyrethrum spray collection, night bite collection, and outdoor resting collection (Smith et al., 2010). The red line indicates the functional curve fitted to these data points (Materials and Methods-Conversion between FOI and the Entomological Inoculation Rate, or EIR), initially proposed by (Smith et al., 2010).

The saturation in FOI with increasing EIR and their non-linear relationship from previous field studies.

(A) and (B) present our empirical estimates (with c = 30) when excluding treated individuals from the analysis. (C) and (D) show our estimates when discarding the infection status and MOI estimates of treated individuals and instead sampling from non-treated ones. Since this case samples non-zero MOIs for these treated and uninfected individuals, it results in an upper bound for FOI estimates. The black points represent paired EIR-FOI values from the literature, as summarized by (Smith et al., 2010), with crosses indicating instances where multiple estimates or ranges were reported or estimated for the same location. The red curve represents the best-fit to these paired EIR-FOI values (Smith et al., 2010). The turquoise hollow diamond and plus represent the Ghana data, showing our FOI estimates using the two methods and the EIR measured in the field by the entomological team (Tiedje et al., 2022).

The data show that mean annual FOI values are consistently below an empirical limit of 20. Due to the highly non-linear relationship between FOI and EIR, there is no single constant factor to convert FOI to EIR (or vice versa) across different settings. In high-transmission regions, transmission efficiency is extremely low: annual EIR can reach a few hundred to a thousand, while annual FOI ranges from about 5 to 10. For future applications, when using our proposed methods to estimate FOI from MOI under sparse sampling schemes, we can rely on this functional curve to convert FOI estimates to corresponding EIR values. There is however an inherently high variance in this conversion. Overall, our paired EIR (measured directly by the entomological team in Ghana (Tiedje et al., 2022)) and FOI estimates align well with previous studies, indicating the consistency of our methods.

The variance inferred by the two-moment approximation method reflects transmission intensity and heterogeneity across individuals

We have focused on presenting the mean FOI in both simulation output and empirical data. The two-moment approximation method also provides estimates for the variance of inter-arrival times of infections. When examining the inferred variance across simulated scenarios, We find that the individual-level mean FOI, when aggregated across the local population, reflects the total number of infections accumulated by that population, regardless of the variance magnitude for FOI.

The inferred variance is most significantly correlated with the inferred mean FOI. Specifically, a smaller mean FOI is associated with a larger variance (Appendix 1-Figure 19). Overall, seasonal runs exhibit slightly greater variance than non-seasonal runs. Runs with heterogeneous transmission (Appendix 1-Figure 6C) have slightly higher variance compared to homogeneous transmission runs. These findings align with expectations, as both seasonality and transmission heterogeneity increase the dispersion of FOI and inter-arrival times of infections.

Discussion

Building on estimates of the multiplicity of infection under sparse sampling schemes, we demonstrate the feasibility of converting these values into estimates of the force of infection. In various simulation scenarios using an extended stochastic agent-based model, the two-moment approximation and Little’s Law methods provide good and replicable FOI estimates based on MOI estimates obtained with the Bayesian formulation of the varcoding method and the bootstrap imputation approach, despite common sampling limitations.

Both methods tend to slightly underestimate FOI due to a bias in underestimating MOI values. The current MOI estimation procedure, specifically the Bayesian formulation of the varcoding method, does not correct for the limited overlap of var genes between co-infecting strains. This limited overlap reduces the number of var genes identified per individual, thereby causing the underestimation bias.

Studies have shown parasite co-transmission from single mosquito bites in high-transmission regions (Nkhoma et al., 2020; Wong et al., 2017), predominantly based on clinical or symptomatic infections (Andolina et al., 2021; Lindblade et al., 2013), although asymptomatic cases constitute the majority of the malaria transmission reservoir in these regions. Co-transmitted recombinant parasites, more closely related to each other than parasites from different bites, further reduce the number of identified var genes (Wong et al., 2022; Nkhoma et al., 2020; Wong et al., 2017) per individual. Low or variable parasite densities due to factors like small blood volumes, genomic DNA quality, clinical status, and within host dynamics also affect MOI estimation (Peyerl-Hoffmann et al., 2001; Bruce et al., 2000; Okell et al., 2012; Farnert et al., 1997; Färnert et al., 2008; Barry et al., 2021; Hergott et al., 2024). These issues are common to all measures of MOI and direct measures of FOI. We did not correct for these factors when estimating MOI and FOI for the Ghana surveys. In the future, the Bayesian formulation could be extended to account for these confounders in identifying the number of co-infecting strains.

Our proposed methods leverage infection duration data from malaria therapy patients with neurosyphilis who had no prior malaria exposure, making them well-suited for FOI inference in highly vulnerable subpopulations with similarly naive immune profiles. For our Ghana surveys, we focus on children aged 1-5 years, as their immune profiles are closer to those of naive patients than those of older individuals. In the pre-IRS phase of the Ghana surveys, an estimated mean FOI of about 5 per host per year suggests that a 4-year-old child would have experienced around 20 infections, which makes them far from naive. However, the extreme documented diversity of var genes (Tiedje et al., 2023) means that even with 20 infections, a 4-year-old may have developed immunity to only a small fraction of the total antigenic diversity encoded by these genes. Consequently, they are not as immunologically experienced as it might initially seem. Moreover, studies have shown that long-lived infections in older children and adults can persist for months or even years, including through the dry season. This persistence is driven by high antigenic variation of var genes and associated incomplete immunity. Additionally, parasites can skew PfEMP1 expression to produce less adhesive erythrocytes, enhancing splenic clearance, reducing virulence, and maintaining extended periods of sub-clinical parasitemia (Andrade et al., 2024; Tran et al., 2013; Zhang and Deitsch, 2022). The impact of immunity on infection duration with age for falciparum malaria remains a challenging open question.

We acknowledge, however, the limitations this aspect of the two methods introduces. To reduce misspecification in infection duration and fully utilize our proposed methods, future data collection and sampling could prioritize subpopulations with minimal prior infections and an immune profile similar to naïve adults, such as infants and toddlers. As these individuals are also the most vulnerable, prioritizing them aligns with the short-term priority of all intervention efforts: to monitor and protect the most vulnerable individuals from severe symptoms and death.

The application of both methods generally requires an evolving time series observed over a long period or many realizations of the same series at a single time. However, empirical surveys often rely on sparse sampling schemes with single-time-point observations per host available. We therefore use the population-level queue length distribution by aggregating MOI estimates across all sampled individuals to approximate the steady-state queue length distribution at the individual level, without explicitly considering individual heterogeneity due to transmission. The estimated FOI represents the FOI experienced by an individual host whose queue length distribution is approximated from the collection of all sampled individuals. Our estimator, combined with demographic information on population size, estimates the total number of Plasmodium falciparum infections acquired by all hosts in the population per year. We evaluate the impact of individual heterogeneity due to transmission on FOI inference using simulations. Even for significant heterogeneity across individuals, our methods show performance comparable to that of homogeneous scenarios. Additionally, our methods perform similarly for both non-seasonal and seasonal transmission scenarios. Alternative frameworks such as fitting a non-homogeneous Poisson process or Gamma process to derive time-varying average arrival rates, or simultaneous estimation of the arrival rates and infection duration, cannot be considered due to their requirements for densely sampled time-series data.

After validating our methods with simulations, we applied them to surveys in Bongo District of northern Ghana, a high-transmission endemic region where estimating multiplicity of infection and related FOI has been challenging. The three-round transient IRS intervention proved strong and effective, resulting in a significant reduction in FOI by more than 70%.

Our Ghana surveys lack direct FOI measurements, which prevents us from directly evaluating our methods as we did with simulation outputs. Empirical MOI-FOI pairs from cohort studies are still lacking, and direct FOI measurements are prone to errors due to challenges in differentiating new infections from the temporary absence and re-emergence of old infections. These challenges arise from the low resolution of polymorphic markers used in cohort studies and the complexity of within-host dynamics. Alternative approaches that fit epidemiological models to densely sampled cross-sectional surveys do not provide direct FOI measurements, leaving no available “true” values for benchmarking against fitted FOI estimates. In these approaches, model parameterization relies on capturing certain epidemiological quantities, such as prevalence or incidence, similar to the one done in this work. We selected FOI values that maximize the likelihood of observing given MOI distributions. Additionally, we pair our estimated FOI value for Ghana surveys with independently measured EIR (Tiedje et al., 2022). We demonstrate alignment between the paired EIR and our FOI estimates with those from previous studies. We acknowledge however that, like these alternative approaches, our validation for field data is indirect and further complicated by high variance in the relationship between EIR and FOI from previous studies.

The FOI estimates obtained through these methods go beyond describing basic malaria epidemiology and evaluating intervention outcomes. FOI for naïve hosts is a fundamental parameter for epidemiological models. FOI of non-naïve hosts is typically a function of their immune status, body size, and the FOI of naïve hosts. Additionally, these FOI estimates can inform process-based models for the population dynamics of complex infectious diseases, serving as priors for parameterizing and validating more complex agent-based or equation-based models, in a way that reduces computational cost, improves efficiency, and minimizes identifiability issues.

A key characteristic of malaria transmission is the saturation in FOI at high transmission, and the highly non-linear relationship between FOI and EIR. While EIR can reach several hundred to a thousand per year, annual FOI typically saturates below 20 (Smith et al., 2010). Different choices of field measures of EIR and FOI cannot account for this drastic difference in magnitude. Transmission is highly inefficient in high-transmission regions with high annual EIRs. The difference between these two quantities is mediated primarily by immunity, or within-host dynamics, measurement bias, and heterogeneous transmission (Donovan et al., 2007; Macdonald, 1950; John et al., 2005; Doolan and Martinez-Alier, 2006). Mathematical models commonly use the probability of transmission from an infectious mosquito bite to bridge FOI and EIR, as a general parameter encapsulating a variety of processes.

FOI saturation poses significant challenges to intervention efforts in high-transmission endemic regions. In these areas, intervention efforts must dramatically reduce EIR by several orders of magnitude to bring FOI below saturation levels. In other words, high-coverage interventions are needed to achieve and start observing any noticeable impacts on individual exposure risk. Theoretical models suggest a sharp non-linear transition towards sustainable low transmission or elimination, influenced by the high antigenic diversity of P. falciparum (de Roos et al., 2023; Zhan et al., 2024a). The same molecular information used here to estimate MOI and FOI in Bongo District underlies this diversity. Recent ABM modeling on resurgence dynamics indicates that highly resilient systems can compensate for reduced FOI following intervention through heterogeneity in infection duration and selection for longer-lasting infections (Zhan et al., 2024a). Consequently, high prevalence and MOI levels can be even maintained during interventions in the regime below FOI saturation.

The proposed methods are applicable to evaluate transmission intensity in many pathogens exhibiting multi-genomic infection due to large antigenic diversity, including those encoding such variation with multigene families (Deitsch et al., 2009). Easier estimation of changes in transmission intensity should enhance the efficiency and evaluation of control programs across a broader range of infectious diseases.

Materials and methods

High genetic diversity of var and the associated strain structure of limiting similarity

We briefly describe the biology of the malaria parasite P. falciparum that underpins our MOI estimation procedure, the Bayesian formulation of the varcoding method.

In high-transmission endemic regions, human hosts remain susceptible to malaria re-infection throughout their lifetime (Doolan et al., 2009). High asymptomatic prevalence and high MOI result from high transmission rates and incomplete host immunity due to the parasite’s high antigenic variation (Deitsch et al., 2009). Parasites achieve this variation and evade the immune system by encoding key variant surface antigens (VSAs) using multigene families (Deitsch et al., 2009). One important multigene family in the malaria parasite P. falciparum is known as var, which encodes PfEMP1 (Plasmodium falciparum erythrocyte membrane protein 1), the major VSA during the blood stage of infection (Zhang and Deitsch, 2022; Baruch et al., 1995; Smith et al., 1995; Su et al., 1995). Each parasite carries 50-60 var genes across its chromosomes, encoding different variants of this protein, which are expressed sequentially.

Empirical sequencing of var genes focuses on the DBLα tag, a conserved ~450bp region encoding the immunogenic Duffy-binding-like alpha domain of PfEMP1 (Tiedje et al., 2023; Ruybal-Pesántez et al., 2017, 2022; Day et al., 2017). Bioinformatic analyses of a large database of exon 1 sequences of var genes revealed a predominantly 1-to-1 DBLα-var relationship, meaning each DBLα tag typically represents a unique var gene (Tan et al., 2023). Hereafter, we use DBLα types and var genes interchangeably.

In high-transmission endemic regions, local parasite populations exhibit a vast pool of var gene variants, ranging from thousands to tens of thousands (Day et al., 2017; Tiedje et al., 2022). These variants are generated primarily through mitotic recombination, but also through meiotic recom-bination, mutation, and host/mosquito vector migration (Claessens et al., 2014; Frank et al., 2008; Freitas-Junior et al., 2000; Bopp et al., 2013). This large pool, combined with negative frequency-dependent selection (NFDS) mediated by hosts’ specific immunity, results in the limited overlap of var genes among individual repertoires (individual parasite genomes) and isolates (sets of in-dividual parasite genomes co-infecting individual hosts) (Day et al., 2017; He et al., 2018). Major groups of var genes are classified based on their 5’-flanking region, called ups, which controls gene expression: upsA and upsB/C (non-upsA) Rask et al. (2010a). Non-upsA DBLα sequences are ~ 20 times more diverse and less conserved among repertoires than the upsA DBLα sequences. Hence our MOI estimation leverages non-upsA DBLα types, as detailed in the following section.

Bayesian formulation of the “varcoding” method for MOI estimation

The limited overlap of var repertoires allows MOI estimation based on the number of non-upsA DBLα types identified from an isolate. The original varcoding method assumes a constant repertoire length, i.e., number of non-upsA DBLα types in a parasite genome, to convert the number of types identified in an isolate to the estimated MOI. This method does not account for the measurement error (Appendix 1-Figure 6D) in this length introduced by the under-sampling of var genes in an infection. We recently extended this method to a Bayesian formulation that considers this error and provides a posterior distribution of MOI values for each sampled individual (Tiedje et al., 2023). We documented the steps of this Bayesian formulation, compared two ways of obtaining population-level MOI distribution (either pooling the maximum a posteriori MOI estimates or calculating a mixture distribution), and examined the impact of different priors (Tiedje et al., 2023). In our analyses here, we provide the estimated population-level MOI distribution obtained from a mixture distribution using a uniform prior for individuals.

Empirical surveys from Ghana

We use empirical data from an interrupted time-series study conducted in Bongo District, northern Ghana. This study involves four age-stratified cross-sectional surveys of ~2,000 participants each, conducted between 2012 and 2016. The study assessed the impacts of a transient threeround IRS intervention, combined with long-lasting insecticidal nets (LLINs), on the asymptomatic P. falciparum reservoir (Tiedje et al., 2017, Tiedje et al., 2022, Tiedje et al., 2023). Surveys were conducted at the end of the wet/high-transmission season (i.e., October) or the dry/low-transmission season (i.e., May/June). The study consists of two phases: (1) Pre-IRS: two surveys before the IRS intervention (Survey 1 in October 2012; Survey 2 in May/June 2013); and (2) Immediately post-IRS: two surveys immediately following the three-round IRS intervention (Survey 3 in October 2015; Survey 4 in May/June 2016) (Appendix 1-Figure 6E). Details on the study area/population, study design, malaria control interventions (IRS and LLINs), inclusion/exclusion criteria, data collection/generation procedures, and routine surveys of antimalarial treatment usage in a previous window of time have been previously described (Tiedje et al., 2017, Tiedje et al., 2022, Tiedje et al., 2023).

The under-sampling of infections in empirical surveys

The empirical MOI estimates in various epidemiological studies, including ours in Bongo District, northern Ghana, rely on microscopy-positive individuals (Tiedje et al., 2023, Tiedje et al., 2022). Due to microscopy’s limited sensitivity, a significant fraction of individuals who carry infections are undetected. A subset of Ghana surveys also include submicroscopic infections detected by PCR (Tiedje et al., 2023, Tiedje et al., 2022), which is significantly more sensitive and can detect a higher fraction, if not 100%, of individuals with P. falciparum infections. Using surveys with both microscopy and PCR detection, we estimate conversion factors of 0.76 for untreated children and 0.67 for antimalarial-treated children aged 1–5 years.

For surveys with both detection methods, we directly calculate the number of microscopynegative but PCR-positive individuals. For surveys with only microscopy data, we use the estimated conversion factors to estimate the number of microscopy-negative but PCR-positive individuals. Additionally, we account for the high but not exactly known sensitivity of PCR by assuming its detectability ranges from relatively low (10% of all PCR-negative individuals carrying undetected infections) to perfect (none of PCR-negative individuals carrying undetected infections). We calculate the number of individuals with undetected infections by PCR for each sensitivity level.

After estimating the number of individuals with undetected infections (both microscopy-negative but PCR-positive and PCR-negative but infected), we sample from existing MOI estimates of microscopy-positive individuals not under antimalarial treatment (see the following section) to represent the missing MOI data.

Similarly, for individuals who are microscopy-positive but lacking var information due to factors like low DNA quality, we sample values from existing MOI estimates of microscopy-positive individuals not under antimalarial treatment to represent the missing MOI data.

We assume that microscopy-negative but PCR-positive children aged 1-5 years and microscopypositive children aged 1-5 years have similar MOI distributions. This assumption is supported by our analysis of Ghana surveys, which shows no clear relationship between parasitemia levels and MOI (or the number of var genes detected within an individual host, on the basis of which our MOI values were estimated) (Appendix 1-Figure 7). We scale the parasitemia levels and the number of non-ups A var genes or MOI estimates before performing the regression. Parasitemia levels underlie the difference in detection sensitivity between PCR and microscopy.

This lack of clear relationship can be attributed to several factors. One factor is immune regulation of parasite density, where host immunity may limit parasite density without reducing the diversity of co-infecting strains (Eldh et al., 2019), leading to individuals with low parasitemia but high MOI. Another factor is asynchronous parasite dynamics, where different parasite clones replicate asynchronously (Färnert et al., 1997), resulting in varied parasite densities that do not directly correlate with the number of distinct strains present. This could explain why individuals with low parasitemia still exhibit multiple strains. Lastly, competition among parasite strains suppress the growth of individual clones, lowering parasite densities while maintaining high strain diversity, thus reducing the expected correlation between MOI and parasitemia (Sondo et al., 2019; Earland et al., 2019b).

Antimalarial drug treatment of infections in empirical surveys

Individuals may seek and receive antimalarial treatment in response to symptoms or perceived transmission risk. In our surveys from northern Ghana, over 50% of children aged 1-5 years responded that they had received an antimalarial treatment in the previous two-weeks (i.e., participants that reported they were sick, sought treatment, and were provided with an antimalarial treatment) prior to the wet/high-transmission survey before IRS (i.e., Survey 1, Appendix 1-Figure 6E) (Tiedje et al., 2022). This fraction is significantly lower for the dry/low-transmission survey before IRS and the surveys collected immediately after IRS (i.e., Survey 2-4, Appendix 1-Figure 6E) (Tiedje et al., 2022).

Disentangling the effect of drug treatment on measurements like infection duration is challenging. Since our methods use infection duration data from naïve malaria therapy patients with neurosyphilis, drug treatment can potentially violate this assumption. We propose two solutions: (1)exclude treated individuals from the analysis; (2) remove treated individuals’ samples and use a bootstrap imputation approach based on the remaining population. Specifically, we sample from the MOI estimates of untreated microscopy-positive individuals to represent MOI estimates for treated individuals, which corrects for individuals who have used antimalarial drugs and show either no infection (MOI= 0) or infection (MOI> 0). Hence, this solution provides an upper bound for FOI estimates. Numerical simulations show our bootstrap imputation approach is robust even with a significant fraction of treated individuals, as seen in our Ghana surveys.

The final distribution of MOI estimates at the population level includes values for microscopy-positive individuals, imputed values for individuals with missing MOI information or false negatives, imputed values for treated individuals (for the second solution), and true zeros for uninfected individuals. This distribution is used for FOI inference.

Inferring FOI from MOI estimates

Malaria transmission in relation to queuing theory

In a cohort of individuals acquiring and clearing infections independently, infections follow a homo-geneous Poisson process with a rate equal to the mean FOI, and each infection has an exponentially distributed duration. At equilibrium, MOI is a Poisson variable with a mean equal to the mean FOI divided by the mean clearance rate (Dietz et al., 1974). However, the arrival of infections often deviates from a homogeneous Poisson process due to factors like seasonality and heterogeneity in exposure risk. Infection durations can also exhibit higher variance than the mean. For instance, under heterogeneous biting, the MOI distribution is better represented by a negative binomial distribution, which account for higher variance. In this case, there is no straightforward relationship between the parameters of the negative binomial distribution and the mean FOI and clearance rate. We conduct a formal test for deviations from Poisson in simulated and empirical MOI distributions, detailed in Appendix 1-Test of deviation from Poisson homogeneity in MOI estimates and supplementary file 2-deviationFromPoissonTest.xlsx.

Despite these complexities, the qualitative relationship between MOI and FOI is intuitive when information on infection duration is known. Higher FOI values should correspond to higher MOI values. Less variable FOI values should result in narrower or more concentrated MOI distributions, whereas more variable FOI values should lead to more spread-out MOI distributions.

The process of acquiring infectious bites is structurally analogous to stochastic queuing theory, which relates queue length to the intensity of arrivals, priority schedules, and service and waiting times. Modeled using differential equations (with the Kolmogorov equations), queueing systems comprise three main components: queue length, the intensity of arrivals, and service times. Knowing any two allows for the inference of the third.

In malaria transmission, hosts resemble service facilities composed of a collection of servers, with each infection akin to a customer. Just as service facilities have a carrying capacity for the number of customers they can serve simultaneously, hosts have a carrying capacity for blood-stage infections, which limits the maximum number of infections they can harbor (Appendix 1-Figure 8). Empirical MOI estimates provide information on queue length. Hypothetically, knowing the service times, i.e., infection durations, can help infer the intensity of arrivals, i.e., the rate at which hosts acquire infections or FOI.

However, determining infection durations is challenging in endemic areas where multi-genomic infections are common. Popular polymorphic markers often fail to distinguish between co-infecting strains (Argyropoulos et al., 2023), complicating the tracking of the emergence and clearance of individual strains from the peripheral blood. Complex within-host dynamics further complicate tracking unless daily sampling is conducted, which is impractical in real settings. Frequent ectopic recombination of var genes complicates assigning genes to specific chromosomal locations. This difficulty in phasing compromises the integrity of individual strains, making it hard to isolate them and track their first appearance and subsequent clearance in blood over time. Additionally, infection duration varies widely across age groups, geographical locations, and sampling times (Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014; Bretscher et al., 2011).

We therefore propose focusing on FOI inference in subpopulations with naive or near-naïve immune profiles. Their infection duration can be approximated by that of naive hosts, as seen in a historical medical study of neurosyphilis patients intentionally infected with malaria as a treatment. Between 1940 and 1963, 318 syphilis patients were infected with a single strain of P. falciparum (Collins and Jeffery, 1999; Maire et al., 2006), and data on fever and parasite counts in the blood were recorded. Since these patients had no prior P. falciparum infections, the documented infection duration reflects that of naive infections.

In Ghana surveys, we focus on children aged 1-5 years, who have accumulated far fewer infections and less immune memory compared to older individuals, an aspect we discuss in the Discussion section. We treat these children as nearly naive and approximate their duration infection with that of naive hosts. Using their MOI estimates, we infer FOI for these children using the following two methods from queuing theory.

A two-moment approximation for a queue of finite capacity

Analysis of multi-server models is challenging, with exact results available only for specific cases, such as the previously mentioned M/M/c/k models. In these models, M represents exponential inter-arrival and service times, c is the number of servers, and k is the maximum queue capacity, including both customers being served and those waiting. Additional models like M/G/c/c and GI/M/c/c +r queues also require exponential distributions for either inter-arrival or service times, where G and GI represent generic random variables and independent and identically distributed (i.i.d.) generic random variables, respectively. These models can often be too restrictive for realworld scenarios as well.

We examine a two-moment approximation method introduced by Kim and Cha (Choi et al., 2005). This method considers the GI/G/c/c + r queue, where inter-arrival times (GI) and service times (G) of customers are independent sequences of i.i.d. general random variables A and S, respectively. There are c (≥ 1) identical servers in parallel and r (≥ 0) waiting places.

Let N denote the number of customers in the system at an arbitrary time, and NA(ND) denotes the number of customers that an arriving customer finds (that a departing customer leaves behind) in steady state. Customers who arrive to find c + r customers in the system depart immediately, leaving those c + r customers behind. Let and denote the probabilities that N = n, NA = n, and ND = n, respectively, for 0 ≤ n ≤ c + r. These probabilities are expressed in terms of the following quantities:

Where , is the residual inter-arrival time at the departure instant of a customer who leaves behind n customers in the system, and , is the residual service time of a randomly chosen busy server at the arrival instant (the departure instant) of a customer who finds (leaves behind) n customers in the system. From these definitions, we have and We set . We assume that all the above quantities are well defined and finite.

Using Theorems 4.3.19 and 4.3.43 from Franken et al (Heyman, 1985), the steady-state queue-length distribution can be derived:

Where

And, by normalization, :

Since quantities , and are difficult to compute in general, Kim and Cha propose an approximation to the exact expression which replaces these unknown arrival- and departure-average quantities by their corresponding (well-known) time-average counterparts, which are exact for exponential inter-arrival and service times. That is: where is the square coefficient of variation (SCV) of a random variable X with distribution function F.

Therefore, a two-moment approximation for the steady-state queue-length distribution is:

Where

And, by normalization, :

Likelihood formulation and parameter estimation

We vary the mean and variance parameters for inter-arrival times across wide ranges (supplementary file 4-meanAndVarianceParams.xlsx). For each mean and variance combination, we calculate the steady-state queue length distribution, i.e., the probability density distribution of MOI, using the two-moment approximation method. The goal is to identify the parameter combination that minimizes the negative log-likelihood (or maximizes the likelihood) of observed MOI distributions from simulated outputs or Ghana surveys: with the likelihood defined as follows: where N is the number of individual hosts, mi is the MOI estimate for individual i, m is the vector of MOI estimates for all hosts, μ and σ represent the mean and variance parameters, respectively, and P is the steady state queue length distribution from the two-moment approximation method with specific mean and variance parameter values, as defined in the previous section.

The shape of the negative log likelihood for both simulated outputs and Ghana surveys is concave upwards around the trough, signifying a clear minimum point (Appendix 1-Figure 9). We tested the impact of different grid value choices on the FOI inference results by refining the grid to include more points, ensuring the FOI inference results are consistent. Specifically, we reduce the grid width for the mean parameter to half and a quarter of the original width, and for the variance parameter to half, a quarter, an eighth, and a sixteenth of the original width. The FOI inference results remain either unchanged or within a 1% deviation from those based on the original grid width (Appendix 1-Figure 10).

Details for deriving confidence intervals for the estimated parameters are provided in Appendix 1-Confidence intervals for FOI inference.

Choice of c and r for the GI/G/c/c + r queue in the two-moment approximation method

When applying the two-moment approximation method, values for the number of parallel servers (c) and waiting places (r) need to be specified. Since MOI is defined exclusively for blood-stage infections, r is set to 0 by default. The parameter c corresponds to the carrying capacity of blood-stage infections. The maximum MOI observed in empirical data from Bongo, based on the varcoding method, is 20. Certain factors which reduce the number of var genes identified in an individual, and thus affect MOI estimation, are not explicitly accounted for in the current MOI estimation (see Discussion), so the actual carrying capacity could be higher. For simplicity, we assume the value of c to be 30 in the simulation. Provided that c is kept consistent across simulations and the two-moment approximation method, this choice should not affect FOI inference. For empirical surveys from Bongo, we set c to 25, 30, 40, and 60 to systematically investigate its impact on FOI inference results. The FOI inference results are similar across these values (Figure 3 and Appendix 1-Figure 16-18).

In general, the choice of c depends on the maximum MOI observed in specific empirical datasets. To account for factors that cause underestimation of MOI, c should be set higher than the observed maximum MOI. Since Bongo District of northern Ghana is a high-transmission endemic region, we expect the range of its c to be applicable to other empirical datasets.

The mean arrival rate of infection from Little’s Law

The second method is known as Little’s Law (Little and Graves, 2008), which describes a relationship between the three main components of queuing systems. This law states that the average number of items in a queuing system L equals the average arrival rate λ multiplied by the average waiting time of an item in the system, W. Reformulating Little’s Law for malaria transmission, the average arrival rate of infection λ equals the average number of blood-stage infections present in an individual L divided by the average duration of infection W.

The relationship is simple and general, holding true regardless of the number of servers (carrying capacity of blood-stage infections in hosts), the service time distribution (infection duration distribution), the distribution of inter-arrival times, the order of service, or the queue structure.

Population-level MOI distribution for approximating time-series observation of MOI per host or many realizations at the same sampling time per host

Both the two-moment approximation and Little’s Law methods require either the steady-state queue length distribution (MOI distribution) or the moments of this distribution. Obtaining such a steady-state queue length distribution necessitates either densely tracked time-series observations per host or many realizations at the same sampling time per host. However, under sparse sampling schemes, we only have two one-time-point observations per host: one at the end of wet/high-transmission and another at the end of dry/low-transmission. This is typically the case for empirical data, although numerical simulations could circumvent this limitation and generate such output.

Nonetheless, we have a population-level queue length distribution from both simulation outputs and empirical data by aggregating MOI estimates across all sampled individuals. We use this distribution to represent and approximate the steady-state queue length distribution at the individual level, not explicitly considering individual heterogeneity due to transmission. We assess the impact of individual heterogeneity due to transmission on FOI inference using simulation outputs (Appendix 1-Figure 6C). The performance of our methods across various simulated scenarios is reported and discussed in the Results and Discussion sections.

Conversion between FOI and the Entomological Inoculation Rate, or EIR

As an indirect proxy for transmission intensity, malariologists typically measure EIR by counting the number of infectious bites a human host receives within a fixed time interval (Shaukat et al., 2010). EIR is considered a standard metric of malaria transmission. Although both FOI and EIR reflect transmission intensity, FOI directly concerns detectable blood-stage infections, while EIR pertains to human-infectious vector contact rates. FOI is defined as the rate at which a host acquires infections, with the focus specifically on blood-stage strains for the following reason. Only blood-stage infections are detectable in all direct measures of FOI. Quantities used in indirect model-fitting approaches for estimating FOI are also based on or reflect these blood-stage strains/infections. Only these blood-stage strains/infections are transmissible to other individuals, impacting disease dynamics.

Studies comparing annual P. falciparum EIR and FOI estimates from age-stratified prevalence in cross-sectional parasitological studies have found significantly different magnitudes for these two quantities (Smith et al., 2010). The number of blood-stage infections per infectious bite (FOI/EIR) is referred to as transmission efficiency. Multiple studies indicate that malaria transmission is inefficient in high-intensity settings, and the reasons for this have been debated. Potential causes include heterogeneous biting, immunity, and measurement bias.

We utilize a functional curve with empirically derived parameters under the assumption of heterogeneous transmission, describing the highly non-linear relationship between reported EIR-FOI pairs (Smith et al., 2010). The functional curve, with FOI h, EIR E, and the corresponding parameters b = 0.55, α = 4.6 and t = 43 days, is as follows:

The EIR-FOI pairs (Smith et al., 2010) and the functional curve provide a basis for converting between these two quantities. EIR data for a subset of surveys in Bongo District, northern Ghana, were obtained (Tiedje et al., 2022). Combined with FOI estimates from our two proposed methods, we generate an EIR-FOI pair for empirical surveys in Bongo District. This enables us to evaluate whether our EIR-FOI pair aligns with historical data and the functional curve with the best-fit parameter values.

eLife Assessment

The ability to estimate the force of infection for Plasmodium falciparum from other more directly measurable epidemiological quantities would contribute to malaria epidemiology. The authors propose a method to accomplish this using genetic data from the var genes of the Pf genome and novel applications of existing methods from queueing theory. After revising the manuscript, this is a useful contribution to the field, and the authors provide solid evidence to support it.

Reviewer #1 (Public review):

Summary

In their paper Zhan et al. have used Pf genetic data from simulated data and Ghanaian field samples to elucidate a relationship between multiplicity of infection (MOI) (the number of distinct parasite clones in a single host infection) and force of infection (FOI). Specifically, they use sequencing data from the var genes of Pf along with Bayesian modeling to estimate MOI individual infections and use these values along with methods from queueing theory that rely on various assumptions to estimate FOI. They compare these estimates to known FOIs in a simulated scenario and describe the relationship between these estimated FOI values and another commonly used metric of transmission EIR (entomological inoculation rate).

This approach does fill an important gap in malaria epidemiology, namely estimating force of infection, which is currently complicated by several factors including superinfection, unknown duration of infection, and highly genetically diverse parasite populations. The authors use a new approach borrowing from other fields of statistics and modeling and make extensive efforts to evaluate their approach under a range of realistic sampling scenarios. However, the write-up would greatly benefit from added clarity both in the description of methods, and in the presentation of the results. Without these clarifications, rigorously evaluating whether the author's proposed method of estimating FOI is sound remains difficult. Additionally, there are several limitations that call into question the stated generalizability of this method that should at minimum be further discussed by authors and in some cases require a more thorough evaluation.

Major comments:

(1) Description and evaluation of FOI estimation procedure.

a. The methods section describing the two-moment approximation and accompanying appendix is lacking several important details. Equations on line 891 and 892 are only a small part of the equations in Choi et al. and do not adequately describe the procedure notably several quantities in those equations are never defined some of them are important to understand the method (e.g. A, S as the main random variables for inter-arrival times and service times, aR and bR which are the known time average quantities, and these also rely on the squared coefficient of variation of the random variable which is also never introduced in the paper). Without going back to the Choi paper to understand these quantities, and to understand the assumptions of this method it was not possible to follow how this works in the paper. At minimum, all variables used in the equations should be clearly defined.

b. Additionally, the description in the main text of how queueing procedure can be used to describe malaria infections would benefit from a diagram currently as written it's very difficult to follow.

c. Just observing the box plots of mean and 95% CI on a plot with the FOI estimate (Figures 1, 2 and 10-14) is not sufficient to adequately assess the performance of this estimator. First, it is not clear whether authors are displaying the bootstrapped 95%Cis or whether they are just showing the distribution of the mean FOI taken over multiple simulations, and then it seems that they are also estimating mean FOI per host on an annual basis. Showing a distribution of those per host estimates would also be helpful. Second, a more quantitative assessment of the ability of the estimator to recover the truth across simulations (e.g. proportion of simulations where the truth is captured in the 95% CI or something like this) is important in many cases it seems that the estimator is always underestimating the true FOI and may not even contain the true value in the FOI distribution (e.g. figure 10, figure 1 under the mid IRS panel). But it's not possible to conclude on way or the other based on this visualization. This is a major issue since it calls into question whether there is in fact data to support that these methods give good and consistent FOI estimates.

d. Furthermore authors state in the methods that the choice of mean and variance (and thus second moment) parameters for inter arrival times are varied widely, however, it's not clear what those ranges are there needs to be a clear table or figure caption showing what combinations of values were tested and which results are produced from them, this is an essential component of the method and it's impossible to fully evaluate its performance without this information. This relates to the issue of selecting the mean and variance values that maximize the likelihood of observing a given distribution of MOI estimates, this is very unclear since no likelihoods have been written down in the methods section of the main text, which likelihood are the authors referring to, is this the probability distribution of the steady state queue length distribution? At other places the authors refer to these quantities as Maximum Likelihood estimators, how do they know they have found the MLE? There are no derivations in the manuscript to support this. The authors should specify and likelihood and include in an appendix why their estimation procedure is in fact maximizing this likelihood preferably with evidence of the shape of the likelihood, and how fine the grid of values they tested are for their mean and variance since this could influence the overall quality of the estimation procedure.

(2) Limitation of FOI estimation procedure.

a. The authors discuss the importance of duration of infection to this problem. While I agree that empirically estimating this is not possible, there are other options besides assuming that all 1-5 year olds have the same duration of infection distribution as naïve adults co-infected with syphilis. E.g. it would be useful to test a wide range of assumed infection duration and assess their impact on the estimation procedure. Furthermore, if the authors are going to stick to the described method for duration of infection, the potentially limited generalizability of this method needs to be further highlighted in both the introduction, and the discussion. In particular, for an estimated mean FOI of about 5 per host per year in the pre-IRS season as estimated in Ghana (Figure 3) it seems that this would not translate to 4 year old being immune naïve, and certainly this would not necessarily generalize well to a school-aged child population or an adult population.

b. The evaluation of the capacity parameter c seems to be quite important, and is set at 30, however, the authors only describe trying values of 25 and 30, and claim that this does not impact FOI inference, however it is not clear that this is the case. What happens if carrying capacity is increased substantially? Alternatively, this would be more convincing if the authors provided a mathematical explanation of why the carrying capacity increasing will not influence the FOI inference, but absent that, this should be mentioned and discussed as a limitation.

Comments on revisions:

The authors have adequately responded to all comments.

Reviewer #2 (Public review):

Summary:

The authors combine a clever use of historical clinical data on infection duration in immunologically naive individuals and queuing theory to infer the force of infection (FOI) from measured multiplicity of infection (MOI) in a sparsely sampled setting. They conduct extensive simulations using agent based modeling to recapitulate realistic population dynamics and successfully apply their method to recover FOI from measured MOI. They then go on to apply their method to real world data from Ghana before and after an indoor residual spraying campaign.

Strengths:

- The use of historical clinical data is very clever in this context

- The simulations are very sophisticated with respect to trying to capture realistic population dynamics

- The mathematical approach is simple and elegant, and thus easy to understand

Weaknesses:

- The assumptions of the approach are quite strong, and the authors have made clear that applicability is constrained to individuals with immune profiles that are similar to malaria naive patients with neurosyphilis. While the historical clinical data is a unique resource and likely directionally correct, it remains somewhat dubious to use the exact estimated values as inputs to other models without extensive sensitivity analysis.

Reviewer #3 (Public review):

Summary:

It has been proposed that the FOI is a method of using parasite genetics to determine changes in transmission in areas with high asymptomatic infection. The manuscript attempts to use queuing theory to convert multiplicity of infection estimates (MOI) into estimates of the force of infection (FOI), which they define as the number of genetically distinct blood-stage strains. They look to validate the method by applying them to simulated results from a previously published agent based model. They then apply these queuing theory methods to previously published and analysed genetic data from Ghana. They then compare their results to previous estimates of FOI.

Strengths:

It would be great to be able to infer FOI from cross sectional surveys which are easier and cheaper than current FOI estimates which require longitudinal studies. This work proposes a method to convert MOI to FOI for cross sectional studies. They attempt to validate this process using a previously published agent based model which helps us understand the complexity of parasite population genetics.

Weaknesses:

(1) I fear that the work could be easily over-interpreted as no true validation was done as no field estimates of FOI (I think considered true validation) were measured. You have developed a method of estimating FOI from MOI which makes a number of biological and structural assumptions. I would not call being able to recreate model results that were generated using a model that makes its own (probably similar) defined set of biological and structural assumptions acts as a validation of what is going on in the field. The authors claim this at times (for example, Line 153 ) and I feel it would be appropriate to differentiate this in the discussion.

(2) Another aspect of the paper is adding greater realism to the previous agent based model, by including assumptions on missing data and under sampling. This takes prominence in the figures and results section, but I would imagine is generally not as interesting to the less specialised reader. The apparent lack of impact of drug treatment on MOI is interesting and counterintuitive, though it is not really mentioned in the results or discussion sufficiently to allay my confusion. I would have been interested in understanding the relationship between MOI and FOI as generated by your queuing theory method and the model. It isn't clear to me why these more standard results are not presented, as I would imagine they are outputs of the model (though happy to stand corrected - it isn't entirely clear to me what the model is doing in this manuscript alone).

(3) I would suggest that outside of malaria geneticists, the force of infection is considered to be the entomological inoculation rate, not the number of genetically distinct blood-stage strains. I appreciate that FOI has been used to explain the later before by others, though the authors could avoid confusion by stating this clearly throughout the manuscript. For example, the abstract says FOI is "the number of new infections acquired by an individual host over a given time interval" which suggests the former, please consider clarifying.

(4) Line 319 says "Nevertheless, overall, our paired EIR (directly measured by the entomological team in Ghana (Tiedje et al., 2022)) and FOI values are reasonably consistent with the data points from previous studies, suggesting the robustness of our proposed methods". I would agree that the results are consistent, given that there is huge variation in Figure 4 despite the transformed scales, but I would not say this suggests a robustness of the method.

(5) The text is a little difficult to follow at times, and sometimes requires multiple reads to understand. Greater precision is needed with the language in a few situations and some of the assumptions made in the modelling process are not referenced, making it unclear whether it is a true representation of the biology.

Comments on revisions:

I think the authors gave a robust but thorough response to our reviews and made some important changes to the manuscript which certainly clarify things for me.

Author response:

The following is the authors’ response to the previous reviews

Public Reviews:

Reviewer #1 (Public review):

The authors have adequately responded to all comments.

We thank Reviewer 1 for their positive assessment of our previous round of revisions.

Reviewer #2 (Public review):

Summary:

The authors combine a clever use of historical clinical data on infection duration in immunologically naive individuals and queuing theory to infer the force of infection (FOI) from measured multiplicity of infection (MOI) in a sparsely sampled setting. They conduct extensive simulations using agent based modeling to recapitulate realistic population dynamics and successfully apply their method to recover FOI from measured MOI. They then go on to apply their method to real world data from Ghana before and after an indoor residual spraying campaign.

Strengths:

- The use of historical clinical data is very clever in this context

- The simulations are very sophisticated with respect to trying to capture realistic population dynamics

- The mathematical approach is simple and elegant, and thus easy to understand

Weakness:

The assumptions of the approach are quite strong, and the authors have made clear that applicability is constrained to individuals with immune profiles that are similar to malaria naive patients with neurosyphilis. While the historical clinical data is a unique resource and likely directionally correct, it remains somewhat dubious to use the exact estimated values as inputs to other models without extensive sensitivity analysis.

We thank reviewer 2 for their comments on our previous round of revisions. The statement here that “it remains somewhat dubious to use the exact estimated values as inputs to other models” suggests that we may not have been sufficiently clear on how infection duration is represented in our agent-based model (ABM) of malaria population dynamics. Because our analysis uses simulated outputs from the ABM to validate the performance of the two queuing-theory methods, we believe this point warrants clarification, which we provide below.

When simulating with the ABM, we do not use empirical estimates of infection duration in immunologically naïve individuals from the historical clinical data as direct inputs. Instead, infection duration emerges from the within-host dynamics modeled in the ABM (lines 800-816, second paragraph of the subsection Within-host dynamics in Appendix 1-Simulation data of the previous revision). Briefly, each Plasmodium falciparum parasite carries approximately 50-60 var genes, each encoding a distinct variant surface antigen expressed during the blood stage of infection. Empirical evidence[1,2] indicates that these var genes are expressed largely sequentially. If a host has previously encountered the antigenic product of a given var gene and retains immunity to it, subject to waning at empirically estimated rates[3,4], the corresponding parasite subpopulation is rapidly cleared. Conversely, if the host is naïve to that gene, it takes approximately seven days for the immune system to mount an effective antibody response, resulting in a rapid decline or elimination of the expressed variant[5]. This seven-day timescale aligns with the duration of each successive parasitemia peak observed in Plasmodium falciparum infections[6,7], each arising primarily from the expression of a single var gene and occasionally from a small number of var genes.

In our previous analyses, we therefore modeled an average expression duration of seven days per gene in naïve hosts. Specifically, the switching time to the next gene was drawn from an exponential distribution with a mean of seven days. Each var gene is represented as a linear combination of two epitopes (alleles), based on the empirical characterization of two hypervariable regions in the var tag region[8], and immunity is acquired against these alleles. Immunity to one allele of a given gene reduces its average expression duration by approximately half, whereas immunity to both alleles results in an immediate switch to another var gene within the infection. Consequently, the total duration of infection is proportional to the number of unseen alleles by the host across all var genes expressed during that infection (lines 800-816, second paragraph of the subsection Within-host dynamics in Appendix 1-Simulation data of the previous revision).

Prompted by the reviewer’s comments, in this revision we additionally tested mean expression durations of 7.5 and 8 days per var gene, together with an extension of the within-host rules. These values were applied in combination with the extended within-host rules (see the next paragraph for motivation and details). Although differences among the three mean expression durations are modest at the per-gene level, when aggregated across all var genes expressed within an individual parasite, the resulting total infection duration can differ by on the order of several months. The resulting distributions of infection duration across immunologically naïve individuals and those aged 1-5 years, together with those generated under our previous simulation settings, span a range of means and variances that lies above and below, but encompasses, scenarios comparable to the historical clinical data from naïve neurosyphilis patients treated with P. falciparum malaria. We have provided example supplementary figures illustrating that the distributions of infection duration from the simulated outputs overlap with, and closely resemble, the empirical distribution from the historical clinical data (Appendix 1-Figure 27-32).

We considered the following modification of the within-host rules. In our previous ABM simulations, we had assumed that an infection would clear only once the parasite had exhausted its entire var gene repertoire, that is, after every var gene had been expressed and recognized. However, biological evidence indicates that clearance can occur earlier for several reasons, including stochastic extinction before full repertoire exhaustion. Even if some var genes remain unexpressed, an infection can terminate due to demographic stochasticity once parasite densities fall to very low levels. This decline in parasite densities may result from non-variant-specific immune mechanisms or from cross-immunity among var genes that share sequence similarity or alleles[9,10,11], both of which can substantially reduce parasite numbers. To model the possibility of termination or clearance before full repertoire exhaustion, we implemented a simple scenario in which there is a small probability of clearing the current infection while a given var gene-whether non-final or final-is being expressed. This probability is a function of the host’s pre-existing immunity to the two epitopes (alleles) of that gene, thereby capturing in a parsimonious manner the effects of cross-immunity among sequence- or allele-sharing var genes in reducing parasitemia. Specifically, it is modeled as a Bernoulli draw whose success probability equals the immunity level against the gene (0 for no immunity to either epitope, 0.5 for immunity to one epitope, and 1 for immunity to both epitopes) multiplied by a constant factor of 0.025. Thus, the probability scales with pre-existing variant-specific immunity to the gene but remains small overall, while introducing additional variance into the emergent distribution of total infection duration across hosts.

We acknowledge that the ABM used to simulate malaria population dynamics cannot capture all mechanisms and complexities underlying within-host processes, many of which remain poorly understood. However, we emphasize that the resulting distributions of infection duration generated by the ABM span a broad range of means, variances, and shapes, including distributions that closely match those observed in the clinical historical data. Because the queueing-theory methods rely on only the mean and variance of infection duration to estimate the force of infection (FOI), these scenarios, which collectively span and encompass values comparable to the empirical ones, provide an appropriate basis for evaluating the performance of the methods using simulated outputs. We have added supplementary figures (see Appendix 1-Figure 16-22) illustrating the corresponding FOI inference results when we allow for clearance before the complete expression of the var repertoire, and the accuracy of FOI estimation remains comparable across all the scenarios examined.

Finally, we emphasize that the application of the queuing-theory methods to the simulated outputs and to the Ghana field survey data involve two self-contained steps. For the simulations, FOI is inferred directly from the emergent distributions of infection duration generated by the ABM. For the Ghana surveys, FOI is inferred using the historical clinical data, which remains one of the few credible and widely used empirical sources for infection duration in immunologically naïve individuals[6]. By exploring different mean expression durations and within-host rules in the ABM, which generates distributions of infection duration that span and encompass those comparable to the empirical distribution, we demonstrate that the queueing-theory methods perform comparably across diverse scenarios and are well suited for application to the Ghana field surveys.

We expanded the section on within-host dynamics in Appendix 1 to elaborate on this point (Lines 817-854).

Reviewer #3 (Public review):

I think the authors gave a robust but thorough response to our reviews and made some important changes to the manuscript which certainly clarify things for me.

We thank Reviewer 3 for their positive feedback on our previous round of revisions.

References

(1) Zhang, X. & Deitsch, K. W. The mystery of persistent, asymptomatic Plasmodium falciparum infections. Curr. Opin. Microbiol 70, 102231 (2022).

(2) Deitsch, K. W. & Dzikowski, R. Variant gene expression and antigenic variation by malaria parasites. Annu. Rev. Microbiol. 71, 625–641 (2017).

(3) Collins, W. E., Skinner, J. C. & Jeffery, G. M. Studies on the persistence of malarial antibody response. American journal of epidemiology, 87(3), 592–598 (1968).

(4) Collins, W. E., Jeffery, G. M. & Skinner, J. C. Fluorescent Antibody Studies in Human Malaria. II. Development and Persistence of Antibodies to Plasmodium falciparum. The American journal of tropical medicine and hygiene, 13, 256–260 (1964).

(5) Gatton, M. L., & Cheng, Q. Investigating antigenic variation and other parasite-host interactions in Plasmodium falciparum infections in naïve hosts. Parasitology, 128(Pt 4), 367–376 (2004).

(6) Maire, N., Smith, T., Ross, A., Owusu-Agyei, S., Dietz, K., & Molineaux, L. A model for natural immunity to asexual blood stages of Plasmodium falciparum malaria in endemic areas. The American journal of tropical medicine and hygiene, 75(2 Suppl), 19–31 (2006).

(7) Chen D. S., Barry A. E., Leliwa-Sytek A., Smith T-A., Peterson I., Brown S. M., et al. A Molecular Epidemiological Study of var Gene Diversity to Characterize the Reservoir of Plasmodium falciparum in Humans in Africa. PLoS ONE 6(2): e16629 (2011).

(8) Larremore D. B., Clauset A., & Buckee C. O. A Network Approach to Analyzing Highly Recombinant Malaria Parasite Genes. PLoS Comput Biol 9(10): e1003268 (2013).

(9) Holding T. & Recker M. Maintenance of phenotypic diversity within a set of virulence encoding genes of the malaria parasite Plasmodium falciparum. J. R. Soc. Interface.1220150848 (2015).

(10) Crompton, P. D., Moebius, J., Portugal, S., Waisberg, M., Hart, G., Garver, L. S., Miller, L. H., Barillas-Mury, C., & Pierce, S. K. Malaria immunity in man and mosquito: insights into unsolved mysteries of a deadly infectious disease. Annual review of immunology, 32, 157–187 (2014).

(11) Langhorne, J., Ndungu, F., Sponaas, AM. et al. Immunity to malaria: more questions than answers. Nat Immunol 9, 725–732 (2008).

Version 3

Abstract

High multiplicity of infection or MOI, the number of genetically distinct parasite strains co-infecting a single human host, characterizes infectious diseases including falciparum malaria at high transmission. This high MOI accompanies high asymptomatic Plasmodium falciparum prevalence despite high exposure, creating a large transmission reservoir challenging intervention. High MOI and asymptomatic prevalence are enabled by immune evasion of the parasite achieved via vast antigenic diversity. Force of infection or FOI, the number of new infections acquired by an individual host over a given time interval, is the dynamic sister quantity of MOI, and a key epidemiological parameter for monitoring antimalarial interventions. FOI remains difficult, expensive, and labor-intensive to accurately measure, especially in high-transmission regions, whether directly via cohort studies or indirectly via the fitting of epidemiological models to repeated cross-sectional surveys. We propose here the application of queuing theory to obtain FOI from MOI, in the form of either a two-moment approximation method or Little’s Law. We illustrate these two methods with MOI estimates obtained under sparse sampling schemes with the “varcoding” approach. The two methods use infection duration data from naive malaria therapy patients with neurosyphilis. Consequently, they are suitable for FOI inference in subpopulations with a similar immune profile and the highest vulnerability, for example, infants or toddlers. Both methods are evaluated with simulation output from a stochastic agent-based model, and are applied to an interrupted time-series study from Bongo District in northern Ghana before and immediately after a three-round transient indoor residual spraying (IRS) intervention. The sampling of the simulation output incorporates limitations representative of those encountered in the collection of field data, including under-sampling of var genes, missing data, and antimalarial drug treatment. We address these limitations in MOI estimates with a Bayesian framework and an imputation bootstrap approach. Both methods yield good and replicable FOI estimates across various simulated scenarios. Applying these methods to the subpopulation of children aged 1-5 years in Ghana field surveys shows over a 70% reduction in annual FOI immediately post-intervention. The proposed methods should be applicable to geographical locations lacking cohort or cross-sectional studies with regular and frequent sampling but having single-time-point surveys under sparse sampling schemes, and for MOI estimates obtained in different ways. They should also be relevant to other pathogens whose immune evasion strategies are based on large antigenic variation resulting in high MOI.

Introduction

Despite substantial intervention efforts, falciparum malaria in high-transmission regions remains a major public health concern, causing mortality among young children and considerable economic burden, particularly in sub-Saharan Africa (World Health Organization, 2023). It is thus still impor-tant to robustly evaluate the effects of intervention efforts in these regions, including on trans-mission intensity. Force of infection (FOI), defined as the number of new Plasmodium falciparum infections acquired by an individual host over a given time interval, is a key metric reflecting the risk of infection and clinical episodes (Mueller et al., 2012). While other metrics may describe the rela-tionship between transmission intensity and the burden of malaria illness on global or continental scales (Carneiro et al., 2010; Beier et al., 1994), FOI can relate local variation in malaria burden to transmission (Mueller et al., 2012). Despite being a key epidemiological parameter for malaria surveillance, FOI is difficult, expensive, and labor-intensive to measure accurately, whether directly through cohort studies or indirectly through the fitting of epidemiological models. As molecu-lar tools for parasite genomics become more readily available, they are contributing to new ap-proaches. In particular, molecular advances are providing a basis for estimating a sister “static” quantity, the multiplicity of infection (MOI), defined as the number of genetically distinct parasite strains co-infecting a single human host (Zhong et al., 2018; Chang et al., 2017; Ruybal-Pesántez et al., 2022; Tiedje et al., 2022; Labbé et al., 2023). We can therefore ask whether we can go further and on the basis of MOI obtain the dynamical, rate, quantity of force of infection?

Early efforts to directly measure FOI included clearing infections and observing time to re-infection (Macdonald, 1950; Pull and Grab, 1974; Msuya and Curtis, 1991; Alonso et al., 2004). Molecular approaches now enable the genotyping of individual parasite infections (Mueller et al., 2012; Hofmann et al., 2017), but differentiating new infections from the temporary absence of an old infection in the peripheral blood and its subsequent re-emergence (Daubersies et al., 1996) re-mains challenging due to the low resolution of polymorphic markers and the complex within-host dynamics of malaria infection. Determining molecular FOI remains challenging, labor-intensive, and costly, requiring close long-term monitoring and genotyping of a large cohort.

Alternatives to direct measurements involve cross-sectional surveys with FOI estimated by fit-ting simple epidemiological models (Bekessy et al., 1976; Muench, 1959; Smith and Vounatsou, 2003; Felger et al., 2012; Mugenyi et al., 2017). These model-fitting procedures require empiri-cal data sampled regularly and frequently, such as age-stratified large cohorts sampled six times a year, to account for FOI and infection duration heterogeneity (Laishram et al., 2012; Simpson et al., 2002; Langhorne et al., 2008a; Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014) influenced by the interplay between host immunity and antigen composition of infections (Piper et al., 1999; Molineaux et al., 2002; Doolan et al., 2009; Barry et al., 2007). Additionally, they may face identifiability issues with model parameters (Mugenyi et al., 2017). Hence, these indirect mea-surements share limitations with direct ones.

Due to the described challenges, FOI has not become a readily available epidemiological quan-tity across geographical locations and times. In contrast, various approaches have been proposed to estimate MOI from clinical samples using size-polymorphic antigenic markers, microsatellites, and panels of biallelic single nucleotide polymorphisms (SNPs) (Felger et al., 1994; Konaté et al., 1999; Anderson et al., 2000; Daniels et al., 2008; Chang et al., 2017). Because Plasmodium parasites reproduce asexually during haploid stages within human hosts (Guttery et al., 2012), polymorphic genotypes indicate multiclonal infection. An alternative approach, termed varcoding, leverages the extreme diversity of the var multigene family, which encodes the major variant surface antigen dur-ing blood-stage infection, and the resulting zero or extremely low similarity between repertoires with respect to their var gene composition shaped by immune selection. As methods for disentan-gling the molecular complexity of natural parasite infections across various transmission settings emerge and mature, MOI becomes more commonly and easily surveyed across space and time. Although MOI remains one of the most frequently used genetic metrics of parasite transmission (Arnot, 1998; Sondo et al., 2020), it is by definition a number and not a rate.

A natural correlation exists however between MOI and FOI, mediated by infection duration, offering an opportunity to convert MOI into FOI. This conversion has been challenging due to MOI estimates often being obtained from sparse sampling schemes, like single-time-point surveys at the end of wet (high-transmission) and dry (low-transmission) seasons (Tiedje et al., 2022; Abukari et al., 2019). Although these MOI estimates have been useful (Earland et al., 2019a; Lee et al., 2006), the sparse sampling scheme has limited their translation into a transmission rate.

In this work, we propose using these MOI estimates for FOI inference with two mathematical modeling frameworks based on queuing theory (Choi et al., 2005; Little and Graves, 2008). The two methods require infection duration values, for which we relied on data from naive malaria therapy patients with neurosyphilis (Collins and Jeffery, 1999; Maire et al., 2006). Consequently, our approach is suited for FOI inference in subpopulations with a similar immune profile and the highest vulnerability, for example, infants or toddlers. We evaluate them through numerical simu-lation of an extended stochastic agent-based model (ABM) (He et al., 2018; Zhan et al., 2024b) for both closed and open systems with constant or seasonal transmission. We consider both homo-geneous and heterogeneous transmission, with the latter including a high-risk group of hosts that receives the majority of the infectious bites. We also examine different statistical distributions for the times between local transmission events. We incorporate limitations representative of those encountered in the collection of field data into the sampling of simulation output, including under-sampling of var genes, missing data, and antimalarial drug treatment. We address these limitations in MOI estimates with a Bayesian framework and an imputation bootstrap approach. Both meth-ods provide good and replicable FOI estimates across simulated scenarios. After validating with simulations, we apply the two methods to empirical data from an interrupted time-series study in Bongo District, northern Ghana, that involved a three-round transient indoor residual spraying (IRS) intervention (Tiedje et al., 2022, 2023). We focus on children aged 1-5 years whose immune profiles are closer to naive patients than the rest of the population, an aspect we discuss later. We then explore the relationship between FOI and another commonly used surrogate for trans-mission intensity, the entomological inoculation rate or EIR, defined as the number of infectious bites received by an individual over a given time period (Shaukat et al., 2010). This relationship un-derscores the challenges of relating transmission to malaria burden at local scales and achieving substantial reductions in transmission in high-transmission regions.

Results

The Bayesian formulation of the varcoding method, combined with the bootstrap imputation approach, effectively addresses sampling limitations often encountered in collecting field data for MOI estimates

Because our FOI inference relies on MOI estimates, we first investigate the impact of various sam-pling limitations on these estimates. We use the Bayesian formulation of the varcoding method and the bootstrap imputation approach (Materials and Methods) to address sampling limitations often encountered in collecting field data for MOI estimates: under-sampling or imperfect detection of var genes, missing data, antimalarial drug treatment, and their combination. For this investigation, we utilize simulation output from an agent-based model (ABM) of malaria transmission with known true MOI values (He et al., 2018; Zhan et al., 2024b). Key assumptions and processes for the ABM, and the experimental design for simulation output are summarized in Appendix 1-Simulation data, with illustrations in Appendix 1-Figure 5-6.

Our results indicate that MOI estimates obtained using the Bayesian formulation of the varcoding method and the bootstrap imputation approach closely match true MOI values in most cases. To assess the difference between estimated and true MOI distributions, we use the Cramer-von Mises and Anderson-Darling tests. The Cramer-von Mises test quantifies the sum of the squared differ-ences between cumulative distribution functions, while the Anderson-Darling test, a modification of the former, gives more weight to the tails of distributions. Most p-values are non-significant (> 0.05), indicating insufficient evidence to conclude that the estimated and true distributions dif-fer, with few exceptions in pre-IRS or low-coverage IRS scenarios. The Bayesian formulation of varcoding tends to underestimate MOI because it assumes that each co-infecting strain contributes a distinct set of var genes. In practice, limited overlap among co-infecting strains reduces the number of var genes detected per individual relative to this expectation, thereby leading to sys-tematic underestimation of MOI. This underestimation bias can be more pronounced in certain high-transmission situations where many hosts have a high true MOI, such as the aforementioned exceptions in pre-IRS or low-coverage IRS scenarios. Consequently, this underestimation in MOI leads to an underestimation of FOI estimates, as described in the next section. Detailed results of both tests are provided in the supplementary file 1-MOImethodsPerformance.xlsx.

The distributions of MOI estimates across different surveys in Ghana and from the simulated outputs are not Poisson-distributed (Appendix 1-Test of deviation from Poisson homogeneity in MOI estimates, supplementary file 2-deviationFromPoissonTest.xlsx) (Potthoff and Whittinghill, 1966; Lloyd-Smith et al., 2005). This deviation suggests that infection arrivals depart from a ho-mogeneous Poisson process. In addition, infection durations often deviate from an exponential distribution, violating the assumptions under which Poisson MOI arises. Together, these depar-tures complicate conversion from MOI to FOI in the presence of a finite host carrying capacity, but the proposed methods are flexible and applicable (Materials and Methods-Inferring FOI from MOI estimates).

The Two-Moment Approximation and Little’s Law methods give good and replicable estimates for FOI across various simulated scenarios

We begin with a homogeneous exposure risk scenario for seasonal transmission in a closed system (Appendix 1-Figure 6A-C, and Appendix 1-Simulation data). The times between local transmission events follow a Gamma distribution (Appendix 1-Figure 6C). We infer FOI using the true MOI val-ues and the MOI estimates obtained via the Bayesian formulation or the bootstrap imputation approach, with each method accounting for the specific sampling limitations applied in that sce-nario, whether a single limitation or all limitations combined. Details on deriving the confidence intervals are provided in Appendix 1-Confidence intervals for FOI inference.

Across pre-IRS and three IRS coverage levels, the 95% confidence intervals and the bootstrap distributions of FOI estimates are narrow. FOI estimates based on the true MOI values, as well as those based on MOI estimates obtained under (i) the missing data limitation and (ii) the antimalarial treatment limitation, each accounted for by the bootstrap imputation approach, closely match the true FOI values (Figure 1). FOI estimates based on MOI estimates obtained under (i) the under-sampling or imperfect detection of var genes and (ii) all sampling limitations combined, accounted for by the Bayesian formulation alone in the former case and by the Bayesian formulation together with the bootstrap imputation approach in the latter, show slight underestimations (Figure 1), due to the underestimation of MOI described in the previous section.

Confidence intervals for estimated mean FOI values in simulated scenarios of homogeneous exposure risk, before and during IRS interventions at three different coverage levels.

The times between local transmission events follow a Gamma distribution, with seasonal transmission in a closed system. FOI estimates are derived from true MOI values and MOI estimates obtained through the Bayesian formulation or the bootstrap imputation approach correcting for all or individual sampling limitations. The true mean FOI per host per year is computed by dividing the total number of infections acquired by the population by the total number of hosts in the population. Each boxplot shows minimum, 5% quantile, median, 95% quantile, and maximum values.

To quantify the difference between inferred and true FOI values, we check if the true FOI lies within the bootstrap distribution and calculate the relative deviation, which is defined as the true FOI value minus the median of the bootstrap distribution for the estimate, normalized by the true FOI value. These details are in supplementary file 3-FOImethodsPerformance.xlsx.

We continue with a heterogeneous exposure risk scenario in which a high-risk group ( of the population) receives approximately 94% of bites, while a low-risk group ( of the population) re-ceives the remainder (Appendix 1-Figure 6C). Transmission is seasonal and the system is semi-open (Appendix 1-Figure 6A-B, and Appendix 1-Simulation data). The times between local transmission events are Gamma-distributed (Appendix 1-Figure 6C). As before, FOI estimates across pre-IRS and three IRS coverage levels show narrow 95% confidence intervals and bootstrap distributions close to true FOI values (Figure 2), with slight underestimations for FOI estimates based on MOI esti-mates corrected for under-sampling or imperfect detection of var genes and those corrected for all sampling limitations.

Confidence intervals for estimated mean FOI values in simulated scenarios of heterogeneous exposure risk, before and during IRS interventions at three different coverage levels.

The times between local transmission events follow a Gamma distribution, with seasonal transmission in a semi-open system. FOI estimates are derived from true MOI values and MOI estimates obtained through the Bayesian formulation of the varcoding method or the bootstrap imputation approach correcting for all or individual sampling limitations. The true mean FOI per host per year is computed by dividing the total number of infections acquired by the population by the total number of hosts in the population. Each boxplot shows minimum, 5% quantile, median, 95% quantile, and maximum values.

The performance of the two methods across additional simulated scenarios is shown in Ap-pendix 1-Figure 11-22.

The Two-Moment Approximation and Little’s Law methods give replicable FOI esti-mates for empirical surveys conducted in the Bongo District of northern Ghana

After validating with simulations, we apply the two methods to empirical surveys in the Bongo District of northern Ghana. We first derive their MOI estimates. Due to the high but imperfect detection power of PCR, we assume three levels of sensitivity: 0% (high detectability), 5% (mid detectability), and 10% (low detectability) of PCR-negative individuals carrying infection (Materials and Methods-The under-sampling of infections in empirical surveys).

Antimalarial treatment, sought in response to symptoms or perceived transmission risk, can impact the duration of an ongoing infection and may therefore violate the assumption underlying the two methods, which rely on infection duration data from naive malaria therapy patients with neurosyphilis. We address this issue by either excluding treated individuals from the analysis or by discarding their infection status and MOI estimates, instead sampling from non-treated individuals with MOI > 0. Since the latter samples non-zero MOIs for these treated and uninfected individuals, it results in an upper bound for FOI estimates. Note that in this latter case, we do not assume that the MOI distribution for treated individuals is the same as that for untreated individuals. Rather, we aim to estimate what their MOI would have been, and consequently, determine what the FOI per individual per year in the combined population would be, had these individuals not received antimalarial treatment. Further details can be found in Materials and Methods-Antimalarial drug treatment of infections in empirical surveys.

Next, we apply the two-moment approximation and Little’s Law methods to derive FOI esti-mates from empirical MOI estimates. Both methods yield replicable FOI estimates. The FOI es-timates are similar across the three PCR sensitivity levels and the two approaches for handling treated individuals. The 95% confidence intervals and the full sampling distributions from boot-strap analysis are concentrated (Figure 3, Appendix 1-Figure 23-25). Notably, there is a significant reduction in FOI, exceeding 70%, indicating that the three-round IRS intervention, although tran-sient, was highly effective.

Confidence intervals for the estimated mean FOI values in Ghana surveys before and immediately after a transient three-round IRS intervention

(A) The estimated FOI values when excluding these treated individuals from the analysis. (B) The estimated FOI values when discarding the infection status and MOI estimates of treated individuals and sampling from non-treated ones with MOI > 0. Since this case samples non-zero MOIs for these treated and uninfected individuals, it results in an upper bound for FOI estimates. Each boxplot shows minimum, 5% quantile, median, 95% quantile, and maximum. The value of c is set to 30. FOI estimates with other values of c can be found in Appendix 1-Figure 23-25.

The inferred FOI and directly measured EIR from the Ghana surveys align with the relationship between these two quantities in previous studies

We plot the measured annual EIR against the estimated annual FOI, and the transmission efficiency (the ratio between FOI and EIR) against the measured annual EIR from previous field studies sum-marized by (Smith et al., 2010) (Figure 4). In these studies, FOI estimation was based on fitting a simple epidemiological model to age-stratified prevalence data from cross-sectional parasitologi-cal studies (Smith et al., 2010; Pull and Grab, 1974), while EIR was estimated using various methods such as exit bait collection, human bait collection, pyrethrum spray collection, night bite collection, and outdoor resting collection (Smith et al., 2010). The red line indicates the functional curve fit-ted to these data points (Materials and Methods-Conversion between FOI and the Entomological Inoculation Rate, or EIR), initially proposed by (Smith et al., 2010).

The saturation in FOI with increasing EIR and their non-linear relationship from previous field studies.

(A) and (B) present our empirical estimates (with c = 30) when excluding treated individuals from the analysis. (C) and (D) show our estimates when discarding the infection status and MOI estimates of treated individuals and instead sampling from non-treated ones with MOI > 0. Since this case samples non-zero MOIs for these treated and uninfected individuals, it results in an upper bound for FOI estimates. The black points represent paired EIR-FOI values from the literature, as summarized by (Smith et al., 2010), with crosses indicating instances where multiple estimates or ranges were reported or estimated for the same location. The red curve represents the best-fit to these paired EIR-FOI values (Smith et al., 2010). The turquoise hollow diamond and plus represent the Ghana data, showing our FOI estimates using the two methods and the EIR measured in the field by the entomological team (Tiedje et al., 2022).

The data show that mean annual FOI values are consistently below an empirical limit of 20. Due to the highly non-linear relationship between FOI and EIR, there is no single constant factor to con-vert FOI to EIR (or vice versa) across different settings. In high-transmission regions, transmission efficiency is extremely low: annual EIR can reach a few hundred to a thousand, while annual FOI ranges from about 5 to 10. For future applications, when using our proposed methods to estimate FOI from MOI under sparse sampling schemes, we can rely on this functional curve to convert FOI estimates to corresponding EIR values. There is however an inherently high variance in this con-version. Overall, our paired EIR (measured directly by the entomological team in Ghana (Tiedje et al., 2022)) and FOI estimates align well with previous studies, indicating the consistency of our methods.

The variance inferred by the two-moment approximation method reflects trans-mission intensity and heterogeneity across individuals

We focused primarily on the mean FOI in both simulation outputs and empirical data. However, the two-moment approximation method also yields estimates of the variance in infection interarrival times. When examining the inferred variance across simulated scenarios, we find that the esti-mated individual-level mean FOI, when aggregated across the local population, robustly reflects the total number of infections accumulated within that population. This relation holds across a wide range of assumptions, including seasonality, systems openness, heterogeneity in exposure risks, and the distribution of infection interarrival times. In other words, it remains robust regard-less of the magnitude of variance in FOI.

The inferred variance is most significantly correlated with the inferred mean FOI. Specifically, a smaller mean FOI is associated with a larger variance (Appendix 1-Figure 26). Overall, seasonal runs exhibit greater variance than non-seasonal runs. Runs with heterogeneous transmission (Ap-pendix 1-Figure 6C) have higher variance compared to homogeneous transmission runs. These findings align with expectations, as both seasonality and transmission heterogeneity increase the dispersion of FOI and inter-arrival times of infections.

Discussion

Building on estimates of the multiplicity of infection under sparse sampling schemes, we demon-strate the feasibility of converting these values into estimates of the force of infection. In various simulation scenarios using an extended stochastic agent-based model, the two-moment approx-imation and Little’s Law methods provide good and replicable FOI estimates based on MOI esti-mates obtained with the Bayesian formulation of the varcoding method and the bootstrap impu-tation approach, despite common sampling limitations.

Both methods tend to slightly underestimate FOI due to a bias in underestimating MOI val-ues. The current MOI estimation procedure, specifically the Bayesian formulation of the varcoding method, does not correct for the limited overlap of var genes between co-infecting strains. This limited overlap reduces the number of var genes identified per individual, while the Bayesian for-mulation implicitly assumes that each co-infecting strain contributes a unique set of var genes, thereby introducing a downward bias in the MOI estimates.

Studies have shown parasite co-transmission from single mosquito bites in high-transmission regions (Nkhoma et al., 2020; Wong et al., 2017), predominantly based on clinical or symptomatic infections (Andolina et al., 2021; Lindblade et al., 2013), although asymptomatic cases constitute the majority of the malaria transmission reservoir in these regions. Co-transmitted recombinant parasites, more closely related to each other than parasites from different bites, further reduce the number of identified var genes (Wong et al., 2022; Nkhoma et al., 2020; Wong et al., 2017) per individual. Low or variable parasite densities due to factors like small blood volumes, genomic DNA quality, clinical status, and within host dynamics also affect MOI estimation (Peyerl-Hoffmann et al., 2001; Bruce et al., 2000; Okell et al., 2012; Farnert et al., 1997; Färnert et al., 2008; Barry et al., 2021; Hergott et al., 2024). These issues are common to all measures of MOI and direct measures of FOI. We did not correct for these factors when estimating MOI and FOI for the Ghana surveys. In the future, the Bayesian formulation could be extended to account for these confounders in estimating the number of co-infecting strains.

Our proposed methods leverage infection duration data from malaria therapy patients with neurosyphilis who had no prior malaria exposure, making them well-suited for FOI inference in highly vulnerable subpopulations with similarly naive immune profiles. For our Ghana surveys, we focus on children aged 1-5 years, as their immune profiles are closer to those of naive patients than those of older individuals. In the pre-IRS phase of the Ghana surveys, an estimated mean FOI of about 5 per host per year suggests that a 4-year-old child would have experienced around 20 infections, which makes them appear far from naive. However, the extreme documented diversity of var genes (Tiedje et al., 2023) means that even with 20 infections, a 4-year-old may have de-veloped immunity to only a small fraction of the total antigenic diversity encoded by these genes. Consequently, they are not as immunologically experienced as it might initially seem. Moreover, studies have shown that long-lived infections in older children and adults can persist for months or even years, including through the dry season. This persistence is driven by high antigenic variation of var genes and associated incomplete immunity. Additionally, parasites can skew PfEMP1 expres-sion to produce less adhesive erythrocytes, enhancing splenic clearance, reducing virulence, and maintaining extended periods of sub-clinical parasitemia (Andrade et al., 2024; Tran et al., 2013; Zhang and Deitsch, 2022). The impact of immunity on infection duration with age for falciparum malaria remains a challenging open question.

We acknowledge, however, the limitations this aspect of the two methods introduces. To re-duce misspecification in infection duration and fully utilize our proposed methods, future data collection and sampling could prioritize subpopulations with minimal prior infections and an im-mune profile similar to naïve adults, such as infants and toddlers. As these individuals are also the most vulnerable, prioritizing them aligns with the short-term priority of all intervention efforts: to monitor and protect the most vulnerable individuals from severe symptoms and death.

The application of both methods generally requires a time series observed over a long period or many realizations of the same series at a single time. However, empirical surveys often rely on sparse sampling schemes with single-time-point observations per host available. We therefore use the population-level queue length distribution by aggregating MOI estimates across all sampled in-dividuals to approximate the steady-state queue length distribution at the individual level, without explicitly considering individual heterogeneity due to transmission. The estimated FOI represents the FOI experienced by an individual host whose queue length distribution is approximated from the collection of all sampled individuals. Our estimator, combined with demographic information on population size, estimates the total number of Plasmodium falciparum infections acquired by all hosts in the population per year. We evaluate the impact of individual heterogeneity due to trans-mission on FOI inference using simulations. Even for significant heterogeneity across individuals, our methods show performance comparable to that of homogeneous scenarios. Additionally, our methods perform similarly for both non-seasonal and seasonal transmission scenarios. Alterna-tive frameworks such as fitting a non-homogeneous Poisson process or Gamma process to derive time-varying average arrival rates, or simultaneous estimation of the arrival rates and infection duration, cannot be considered due to their requirements for densely sampled time-series data.

After validating our methods with simulations, we applied them to surveys in Bongo District of northern Ghana, a high-transmission endemic region where estimating multiplicity of infection and related FOI has been challenging. The three-round transient IRS intervention proved strong and effective, resulting in a significant reduction in FOI by more than 70%.

Our Ghana surveys lack direct FOI measurements, which prevents us from directly evaluating our methods as we did with simulation outputs. Empirical MOI-FOI pairs from cohort studies are still lacking, and direct FOI measurements are prone to errors due to challenges in differentiating new infections from the temporary absence and re-emergence of old infections. These challenges arise from the low resolution of polymorphic markers used in cohort studies and the complexity of within-host dynamics. Alternative approaches that fit epidemiological models to densely sampled cross-sectional surveys do not provide direct FOI measurements, leaving no available “true” values for benchmarking against fitted FOI estimates. In these approaches, model parameterization relies on capturing certain epidemiological quantities, such as prevalence or incidence, similar to the one done in this work. We selected FOI values that maximize the likelihood of observing given MOI distributions. Additionally, we pair our estimated FOI value for Ghana surveys with independently measured EIR (Tiedje et al., 2022). We demonstrate alignment between the paired EIR and our FOI estimates with those from previous studies. We acknowledge however that, like these alternative approaches, our validation for field data is indirect and further complicated by high variance in the relationship between EIR and FOI from previous studies.

The FOI estimates obtained through these methods go beyond describing basic malaria epi-demiology and evaluating intervention outcomes. FOI for naïve hosts is a fundamental parameter for epidemiological models. FOI of non-naïve hosts is typically a function of their immune status, body size, and the FOI of naïve hosts. Additionally, these FOI estimates can inform process-based models for the population dynamics of complex infectious diseases, serving as priors for parame-terizing and validating more complex agent-based or equation-based models, in a way that reduces computational cost, improves efficiency, and minimizes identifiability issues.

A key characteristic of malaria transmission is the saturation in FOI at high transmission, and the highly non-linear relationship between FOI and EIR. While EIR can reach several hundred to a thousand per year, annual FOI typically saturates below 20 (Smith et al., 2010). Different choices of field measures of EIR and FOI cannot account for this drastic difference in magnitude. Transmis-sion is highly inefficient in high-transmission regions with high annual EIRs. The difference between these two quantities is mediated primarily by immunity, or within-host dynamics, measurement bias, and heterogeneous transmission (Donovan et al., 2007; Macdonald, 1950; John et al., 2005; Doolan and Martinez-Alier, 2006). Mathematical models commonly use the probability of transmis-sion from an infectious mosquito bite to bridge FOI and EIR, as a general parameter encapsulating a variety of processes.

FOI saturation poses significant challenges to intervention efforts in high-transmission endemic regions. In these areas, intervention efforts must dramatically reduce EIR by several orders of mag-nitude to bring FOI below saturation levels. In other words, high-coverage interventions are needed to achieve and start observing any noticeable impacts on individual exposure risk. Theoretical models suggest a sharp non-linear transition towards sustainable low transmission or elimination, influenced by the high antigenic diversity of P. falciparum (de Roos et al., 2023; Zhan et al., 2024a). The same molecular information used here to estimate MOI and FOI in Bongo District underlies this diversity. Recent ABM modeling on resurgence dynamics indicates that highly resilient systems can compensate for reduced FOI following intervention through heterogeneity in infection dura-tion and selection for longer-lasting infections (Zhan et al., 2024a). Consequently, high prevalence and MOI levels can be even maintained during interventions in the regime below FOI saturation.

The proposed methods are applicable to evaluate transmission intensity in many pathogens exhibiting multi-genomic infection due to large antigenic diversity, including those encoding such variation with multigene families (Deitsch et al., 2009). Easier estimation of changes in transmission intensity should enhance the efficiency and evaluation of control programs across a broader range of infectious diseases.

Materials and Methods

High genetic diversity of var and the associated strain structure of limiting similar-ity

We briefly describe the biology of the malaria parasite P. falciparum that underpins our MOI esti-mation procedure, the Bayesian formulation of the varcoding method.

In high-transmission endemic regions, human hosts remain susceptible to malaria re-infection throughout their lifetime (Doolan et al., 2009). High asymptomatic prevalence and high MOI result from high transmission rates and incomplete host immunity due to the parasite’s high antigenic variation (Deitsch et al., 2009). Parasites achieve this variation and evade the immune system by encoding key variant surface antigens (VSAs) using multigene families (Deitsch et al., 2009). One important multigene family in the malaria parasite P. falciparum is known as var, which encodes PfEMP1 (Plasmodium falciparum erythrocyte membrane protein 1), the major VSA during the blood stage of infection (Zhang and Deitsch, 2022; Baruch et al., 1995; Smith et al., 1995; Su et al., 1995). Each parasite carries 50-60 var genes across its chromosomes, encoding different variants of this protein, which are expressed largely sequentially (Appendix 1–Simulation data, subsection “An ex-tended var model,” sub-subsection “Within-host dynamics”).

Empirical sequencing of var genes focuses on the DBLα tag, a conserved ∼450bp region en-coding the immunogenic Duffy-binding-like alpha domain of PfEMP1 (Tiedje et al., 2023; Ruybal-Pesántez et al., 2017, 2022; Day et al., 2017). Bioinformatic analyses of a large database of exon 1 sequences of var genes revealed a predominantly 1-to-1 DBLα-var relationship, meaning each DBLα tag typically represents a unique var gene (Tan et al., 2023). Hereafter, we use DBLα types and var genes interchangeably.

In high-transmission endemic regions, local parasite populations exhibit a vast pool of var gene variants, ranging from thousands to tens of thousands (Day et al., 2017; Tiedje et al., 2022). These variants are generated primarily through mitotic recombination, but also through meiotic recom-bination, mutation, and host/mosquito vector migration (Claessens et al., 2014; Frank et al., 2008; Freitas-Junior et al., 2000; Bopp et al., 2013). This large pool, combined with negative frequency-dependent selection (NFDS) mediated by hosts’ specific immunity, results in the limited overlap of var genes among individual repertoires (individual parasite genomes) and isolates (sets of in-dividual parasite genomes co-infecting individual hosts) (Day et al., 2017; He et al., 2018). Major groups of var genes are classified based on their 5’-flanking region, called ups, which controls gene expression: upsA and upsB/C (non-upsA) (Rask et al., 2010a). Non-upsA DBLα sequences are ∼ 20 times more diverse and less conserved among repertoires than the upsA DBLα sequences. Hence our MOI estimation leverages non-upsA DBLα types, as detailed in the following section.

Bayesian formulation of the “varcoding” method for MOI estimation

The limited overlap of var repertoires allows MOI estimation based on the number of non-upsA DBLα types identified from an isolate. The original varcoding method assumes a constant reper-toire length, i.e., number of non-upsA DBLα types in a parasite genome, to convert the number of types identified in an isolate to the estimated MOI. This method does not account for the measure-ment error (Appendix 1-Figure 6D) in this length introduced by the under-sampling or imperfect detection of var genes in an infection. We recently extended this method to a Bayesian formula-tion that considers this error and provides a posterior distribution of MOI values for each sampled individual (Tiedje et al., 2023). We documented the steps of this Bayesian formulation, compared two ways of obtaining population-level MOI distribution (either pooling the maximum a posteriori MOI estimates or calculating a mixture distribution), and examined the impact of different priors (Tiedje et al., 2023). In our analyses here, we provide the estimated population-level MOI distribu-tion obtained from a mixture distribution using a uniform prior for individuals.

Empirical surveys from Ghana

We use empirical data from an interrupted time-series study conducted in Bongo District, north-ern Ghana. This study involves four age-stratified cross-sectional surveys of ∼2,000 participants each, conducted between 2012 and 2016. The study assessed the impacts of a transient three-round IRS intervention, combined with long-lasting insecticidal nets (LLINs), on the asymptomatic P. falciparum reservoir (Tiedje et al., 2017, 2022, 2023). Surveys were conducted at the end of the wet/high-transmission season (i.e., October) or the dry/low-transmission season (i.e., May/June). The study consists of two phases: (1) Pre-IRS: two surveys before the IRS intervention (Survey 1 in October 2012; Survey 2 in May/June 2013); and (2) Immediately post-IRS: two surveys immediately following the three-round IRS intervention (Survey 3 in October 2015; Survey 4 in May/June 2016) (Appendix 1-Figure 6E). Details on the study area/population, study design, malaria control inter-ventions (IRS and LLINs), inclusion/exclusion criteria, data collection/generation procedures, and routine surveys of antimalarial treatment usage in a previous window of time have been previously described (Tiedje et al., 2017, 2022, 2023).

The under-sampling of infections in empirical surveys

The empirical MOI estimates in various epidemiological studies, including ours in Bongo District, northern Ghana, rely on microscopy-positive individuals (Tiedje et al., 2023, 2022). Due to mi-croscopy’s limited sensitivity, a significant fraction of individuals who carry infections are unde-tected. A subset of Ghana surveys also include submicroscopic infections detected by PCR (Tiedje et al., 2023, 2022), which is significantly more sensitive and can detect a higher fraction, if not 100%, of individuals with P. falciparum infections. Using surveys with both microscopy and PCR detection, we estimate conversion factors of 0.76 for untreated children and 0.67 for antimalarial-treated children aged 1–5 years.

For surveys with both detection methods, we directly calculate the number of microscopy-negative but PCR-positive individuals. For surveys with only microscopy data, we use the estimated conversion factors to estimate the number of microscopy-negative but PCR-positive individuals. Additionally, we account for the high but not exactly known sensitivity of PCR by assuming its de-tectability ranges from relatively low (10% of all PCR-negative individuals carrying undetected in-fections) to perfect (none of PCR-negative individuals carrying undetected infections). We calculate the number of individuals with undetected infections by PCR for each sensitivity level.

After estimating the number of individuals with undetected infections (both microscopy-negative but PCR-positive and PCR-negative but infected), we sample from existing MOI estimates of microscopy-positive individuals not under antimalarial treatment (see the following section) to represent the missing MOI data.

Similarly, for individuals who are microscopy-positive but lacking var information due to fac-tors like low DNA quality, we sample values from existing MOI estimates of microscopy-positive individuals not under antimalarial treatment to represent the missing MOI data.

We assume that microscopy-negative but PCR-positive children aged 1-5 years and microscopy-positive children aged 1-5 years have similar MOI distributions. This assumption is suggested by our analysis of Ghana surveys, which shows no clear relationship between parasitemia levels and MOI (or the number of var genes detected within an individual host, on the basis of which our MOI values were estimated) (Appendix 1-Figure 7). We scale the parasitemia levels and the number of non-ups A var genes or MOI estimates before performing the regression. Parasitemia levels underlie the difference in detection sensitivity between PCR and microscopy.

This lack of clear relationship can be attributed to several factors. One factor is immune reg-ulation of parasite density, where host immunity may limit parasite density without reducing the diversity of co-infecting strains (Eldh et al., 2019), leading to individuals with low parasitemia but high MOI. Another factor is asynchronous parasite dynamics, where different parasite clones repli-cate asynchronously (Färnert et al., 1997), resulting in varied parasite densities that do not directly correlate with the number of distinct strains present. This could explain why individuals with low parasitemia still exhibit multiple strains. Lastly, competition among parasite strains suppress the growth of individual clones, lowering parasite densities while maintaining high strain diversity, thus reducing the expected correlation between MOI and parasitemia (Sondo et al., 2019; Earland et al., 2019b).

Antimalarial drug treatment of infections in empirical surveys

Individuals may seek and receive antimalarial treatment in response to symptoms or perceived transmission risk. In our surveys from northern Ghana, over 50% of children aged 1-5 years re-sponded that they had received an antimalarial treatment in the previous two-weeks (i.e., partic-ipants that reported they were sick, sought treatment, and were provided with an antimalarial treatment) prior to the wet/high-transmission survey before IRS (i.e., Survey 1, Appendix 1-Figure 6E) (Tiedje et al., 2022). This fraction is significantly lower for the dry/low-transmission survey be-fore IRS and the surveys collected immediately after IRS (i.e., Survey 2-4, Appendix 1-Figure 6E) (Tiedje et al., 2022).

Disentangling the effect of drug treatment on measurements like infection duration is chal-lenging. Since our methods use infection duration data from naïve malaria therapy patients with neurosyphilis, drug treatment can potentially violate this assumption. We propose two solutions: (1) exclude treated individuals from the analysis; (2) remove treated individuals’ samples and use a bootstrap imputation approach based on the remaining population. Specifically, we sample from the MOI estimates of untreated microscopy-positive individuals to represent MOI estimates for treated individuals, which corrects for individuals who have used antimalarial drugs and show ei-ther no infection (MOI = 0) or infection (MOI > 0). Hence, this solution provides an upper bound for FOI estimates. Numerical simulations show our bootstrap imputation approach is robust even with a significant fraction of treated individuals, as seen in our Ghana surveys.

The final distribution of MOI estimates at the population level includes values for microscopy-positive individuals, imputed values for individuals with missing MOI information or false negatives, imputed values for treated individuals (for the second solution), and true zeros for uninfected in-dividuals. This distribution is used for FOI inference.

Inferring FOI from MOI estimates

Malaria transmission in relation to queuing theory

In a cohort of individuals acquiring and clearing infections independently, infections occurs as a homogeneous Poisson process with a rate equal to the mean FOI, and each infection has an ex-ponentially distributed duration. At equilibrium, MOI follows a Poisson distribution with a mean equal to the mean FOI divided by the mean clearance rate (Dietz et al., 1974). In practice, indi-viduals are often capacity-limited, such that they can only carry up to a certain number of concur-rent infections due to within-host competition or immune regulation. New infections that arrive when a host is already at capacity can be simply blocked rather than queued or delayed, resulting in an MOI that follows a Poisson distribution truncated at the carrying capacity and normalized. However, infection arrivals frequently deviate from a homogeneous Poisson process because of factors such as seasonality and heterogeneity in exposure risk, and infection duration may also deviates from an exponential distribution. Consequently, the MOI distribution is often overdis-persed. In this case, no simple analytical relationship exists between the MOI distribution and the mean FOI or clearance rate. We formally test for deviations from Poisson homogeneity against a negative binomial alternative in both simulated and empirical MOI distributions, as detailed in Appendix 1-Test of deviation from Poisson homogeneity in MOI estimates and supplementary file 2-deviationFromPoissonTest.xlsx. Notably, infection durations observed in malaria-naive patients from historical neurosyphilis treatment studies, where patients were intentionally infected with malaria, also exhibit a standard deviation substantially different from their mean (Collins and Jef-fery, 1999; Maire et al., 2006) (see details below).

Despite these complexities, the qualitative relationship between MOI and FOI is intuitive when information on infection duration is known. Higher FOI values should correspond to higher MOI values. Less variable FOI values should result in narrower or more concentrated MOI distributions, whereas more variable FOI values should lead to more spread-out MOI distributions.

The process of acquiring infectious bites is structurally analogous to stochastic queuing theory, which relates queue length to the intensity of arrivals, priority schedules, and service and waiting times. Modeled using differential equations (with the Kolmogorov equations), queueing systems comprise three main components: queue length, the intensity of arrivals, and service times. Know-ing any two allows for the inference of the third.

In malaria transmission, hosts resemble service facilities composed of a collection of servers, with each infection akin to a customer. Just as service facilities have a carrying capacity for the number of customers they can serve simultaneously, hosts have a carrying capacity for blood-stage infections, which limits the maximum number of infections they can harbor (Appendix 1-Figure 8). Empirical MOI estimates provide information on queue length. Hypothetically, knowing the service times, i.e., infection durations, can help infer the intensity of arrivals, i.e., the rate at which hosts acquire infections or FOI.

However, determining infection durations is challenging in endemic areas where multi-genomic infections are common. Popular polymorphic markers often fail to distinguish between co-infecting strains (Argyropoulos et al., 2023), complicating the tracking of the emergence and clearance of individual strains from the peripheral blood. Complex within-host dynamics further complicate tracking unless daily sampling is conducted, which is impractical in real settings. Frequent ectopic recombination of var genes complicates assigning genes to specific chromosomal locations. This difficulty in phasing compromises the integrity of individual strains, making it hard to isolate them and track their first appearance and subsequent clearance in blood over time. Additionally, infec-tion duration varies widely across age groups, geographical locations, and sampling times (Childs and Buckee, 2014; Chang et al., 2016; Ashley and White, 2014; Bretscher et al., 2011).

We therefore propose focusing on FOI inference in subpopulations with naive or near-naive immune profiles. Their infection duration can be approximated by that of naive hosts, as seen in a historical medical study of neurosyphilis patients intentionally infected with malaria as a treat-ment. Between 1940 and 1963, 318 syphilis patients were infected with a single strain of P. falci-parum (Collins and Jeffery, 1999; Maire et al., 2006), and data on fever and parasite counts in the blood were recorded. Since these patients had no prior P. falciparum infections, the documented infection duration reflects that of naive infections.

In Ghana surveys, we focus on children aged 1-5 years, who have accumulated far fewer in-fections and less immune memory compared to older individuals, an aspect we discuss in the Discussion section. We treat these children as nearly naive and approximate their duration infec-tion with that of naive hosts. Using their MOI estimates, we infer FOI for these children using the following two methods from queuing theory.

A two-moment approximation for a queue of finite capacity

Analysis of multi-server models is challenging, with exact results available only for specific cases, such as the previously mentioned M∕M∕c∕k models. In these models, M represents exponential inter-arrival and service times, c is the number of servers, and k is the maximum queue capacity, including both customers being served and those waiting. Additional models like M∕G∕c∕c and GI∕M∕c∕c +r queues also require exponential distributions for either inter-arrival or service times, where G and GI represent generic random variables and independent and identically distributed (i.i.d.) generic random variables, respectively. These models can often be too restrictive for real-world scenarios as well.

We examine a two-moment approximation method introduced by Kim and Cha (Choi et al., 2005). This method considers the GI∕G∕c∕c + r queue, where inter-arrival times (GI) and service times (G) of customers are independent sequences of i.i.d. general random variables A and S, re-spectively. There are c (≥ 1) identical servers in parallel and r (≥ 0) waiting places. In this framework, overdispersion in MOI arises from temporally structured (non-Poisson) acquisition processes that generate bursts of infection, compounded by non-exponential infection durations, among other-wise homogeneous hosts.

Let N denote the number of customers in the system at an arbitrary time, and NA(ND) denotes the number of customers that an arriving customer finds (that a departing customer leaves behind) in steady state. Customers who arrive to find c + r customers in the system depart immediately, leaving those c + r customers behind. Let Pn, and denote the probabilities that N = n, NA = n, and ND = n, respectively, for 0 ≤ n ≤ c + r. These probabilities are expressed in terms of the following quantities: where , 0 ≤ n ≤ c +r, is the residual inter-arrival time at the departure instant of a customer who leaves behind n customers in the system, and , 1 ≤ n ≤ c + r, is the residual service time of a randomly chosen busy server at the arrival instant (the departure instant) of a customer who finds (leaves behind) n customers in the system. From these definitions, we have . We set . We assume that all the above quantities are well defined and finite.

Using Theorems 4.3.19 and 4.3.43 from Franken et al (Heyman, 1985), the steady-state queue-length distribution can be derived: where And, by normalization, : Since quantities are difficult to compute in general, Kim and Cha propose an ap-proximation to the exact expression which replaces these unknown arrival- and departure-average quantities by their corresponding (well-known) time-average counterparts, which are exact for ex-ponential inter-arrival and service times. That is: where is the square coefficient of variation (SCV) of a random variable X with distribu-tion function F.

Therefore, a two-moment approximation for the steady-state queue-length distribution is: where And, by normalization, :

Likelihood formulation and parameter estimation

We vary the mean and variance parameters for inter-arrival times across wide ranges (supplementary file 4-meanAndVarianceParams.xlsx). For each mean and variance combination, we calculate the steady-state queue length distribution, i.e., the probability density distribution of MOI, using the two-moment approximation method. The goal is to identify the parameter combination that minimizes the negative log-likelihood (or maximizes the likelihood) of observed MOI distributions from simulated outputs or Ghana surveys: with the likelihood defined as follows: where N is the number of individual hosts, mi is the MOI estimate for individual i, m is the vector of MOI estimates for all hosts, μ and σ represent the mean and variance parameters, respectively, and P is the steady state queue length distribution from the two-moment approximation method with specific mean and variance parameter values, as defined in the previous section.

The shape of the negative log likelihood for both simulated outputs and Ghana surveys is con-cave upwards around the trough, signifying a clear minimum point (Appendix 1-Figure 9). We tested the impact of different grid value choices on the FOI inference results by refining the grid to include more points, ensuring the FOI inference results are consistent. Specifically, we reduce the grid width for the mean parameter to half and a quarter of the original width, and for the variance parameter to half, a quarter, an eighth, and a sixteenth of the original width. The FOI inference results remain either unchanged or within a 1% deviation from those based on the original grid width (Appendix 1-Figure 10).

Details for deriving confidence intervals for the estimated parameters are provided in Appendix 1-Confidence intervals for FOI inference.

Choice of c and r for the GI∕G∕c∕c + r queue in the two-moment approximation method When applying the two-moment approximation method, values for the number of parallel servers (c) and waiting places (r) need to be specified. Since MOI is defined exclusively for blood-stage infec-tions, r is set to 0 by default. The parameter c corresponds to the carrying capacity of blood-stage infections. The maximum MOI observed in empirical data from Bongo, based on the varcoding method, is 20. Certain factors which reduce the number of var genes identified in an individual, and thus affect MOI estimation, are not explicitly accounted for in the current MOI estimation (see Discussion), so the actual carrying capacity could be higher. For simplicity, we assume the value of c to be 30 in the simulation. Provided that c is kept consistent across simulations and the two-moment approximation method, this choice should not affect FOI inference. For empirical surveys from Bongo, we set c to 25, 30, 40, and 60 to systematically investigate its impact on FOI inference results. The FOI inference results are similar across these values (Figure 3 and Appendix 1-Figure 23-25).

In general, the choice of c depends on the maximum MOI observed in a given empirical dataset under high transmission. To account for factors that may lead to underestimation of MOI, c should be set higher than the observed maximum MOI. Since Bongo District of northern Ghana is a high-transmission endemic region, we expect the range of its c to be applicable to other empirical datasets.

The mean arrival rate of infection from Little’s Law

The second method is known as Little’s Law (Little and Graves, 2008), which describes a relation-ship between the three main components of queuing systems. This law states that the average number of items in a queuing system L equals the average arrival rate λ multiplied by the average waiting time of an item in the system, W. Reformulating Little’s Law for malaria transmission, the average arrival rate of infection λ equals the average number of blood-stage infections present in an individual L divided by the average duration of infection W. The relationship is simple and general, holding true regardless of the number of servers (car-rying capacity of blood-stage infections in hosts), the service time distribution (infection duration distribution), the distribution of inter-arrival times, the order of service, or the queue structure. Population-level MOI distribution for approximating time-series observation of MOI per host or many realizations at the same sampling time per host Both the two-moment approximation and Little’s Law methods require either the steady-state queue length distribution (MOI distribution) or the moments of this distribution. Obtaining such a steady-state queue length distribution necessitates either densely tracked time-series observa-tions per host or many realizations at the same sampling time per host. However, under sparse sampling schemes, we only have two one-time-point observations per host: one at the end of wet/high-transmission and another at the end of dry/low-transmission. This is typically the case for empirical data, although numerical simulations could circumvent this limitation and generate such output.

Nonetheless, we have a population-level queue length distribution from both simulation out-puts and empirical data by aggregating MOI estimates across all sampled individuals. We use this distribution to represent and approximate the steady-state queue length distribution at the indi-vidual level, not explicitly considering individual heterogeneity due to transmission. We assess the impact of individual heterogeneity due to transmission on FOI inference using simulation outputs (Appendix 1-Figure 6C). The performance of our methods across various simulated scenarios is reported and discussed in the Results and Discussion sections.

Conversion between FOI and the Entomological Inoculation Rate, or EIR

As an indirect proxy for transmission intensity, malariologists typically measure EIR by counting the number of infectious bites a human host receives within a fixed time interval (Shaukat et al., 2010). EIR is considered a standard metric of malaria transmission. Although both FOI and EIR reflect transmission intensity, FOI directly concerns detectable blood-stage infections, while EIR pertains to human-infectious vector contact rates. FOI is defined as the rate at which a host acquires infec-tions, with the focus specifically on blood-stage strains for the following reason. Only blood-stage infections are detectable in all direct measures of FOI. Quantities used in indirect model-fitting approaches for estimating FOI are also based on or reflect these blood-stage strains/infections. Only these blood-stage strains/infections are transmissible to other individuals, impacting disease dynamics.

Studies comparing annual P. falciparum EIR and FOI estimates from age-stratified prevalence in cross-sectional parasitological studies have found significantly different magnitudes for these two quantities (Smith et al., 2010). The number of blood-stage infections per infectious bite (FOI/EIR) is referred to as transmission efficiency. Multiple studies indicate that malaria transmission is inefficient in high-intensity settings, and the reasons for this have been debated. Potential causes include heterogeneous biting, immunity or within-host dynamics, and measurement bias.

We utilize a functional curve with empirically derived parameters under the assumption of het-erogeneous transmission, describing the highly non-linear relationship between reported EIR-FOI pairs (Smith et al., 2010). The functional curve, with FOI 𝑕, EIR E, and the corresponding parameters b = 0.55, α = 4.6 and t = 43 days, is as follows: The EIR-FOI pairs (Smith et al., 2010) and the functional curve provide a basis for converting between these two quantities. EIR data for a subset of surveys in Bongo District, northern Ghana, were obtained (Tiedje et al., 2022). Combined with FOI estimates from our two proposed methods, we generate an EIR-FOI pair for empirical surveys in Bongo District. This enables us to evaluate whether our EIR-FOI pair aligns with historical data and the functional curve with the best-fit pa-rameter values.